# LM2GLMM

This is the repository for the R package LM2GLMM.
This R package is only aimed at the students and helpers of the course I am giving at the 
Freie Universität, Berlin. It is not aimed at general users.

## Tips for the students

### Solution 1 to install the package

1. install the R package ```devtools```

```{r, eval = FALSE}
install.packages("devtools")
```

2. install the R package LM2GLMM:

```{r, eval = FALSE}
devtools::install_github("courtiol/LM2GLMM")
```

### Solution 2 to install the package

1. Re(install) the dependencies:

```{r, eval = FALSE}
install.packages(c('car', 'caret', 'spaMM', 'lme4', 'lmtest', 'multcomp', 'mvtnorm', 'nortest', 'fastR', 'rgl', 'sem'))
```

2. Download the release version on the package, available here:
https://github.com/courtiol/LM2GLMM/releases

### Other tips:

To list the vignettes:
```{r, eval = FALSE}
browseVignettes(package = "LM2GLMM")
```

## Tips for the helpers

To rebuild the package using devtools:
```{r, eval = FALSE}
devtools::build_vignettes()
devtools::build(binary = TRUE)
devtools::build(binary = FALSE)
devtools::install()
```
