---
title: "Answers to Exercises: Let's practice more GLM"
author: "Alexandre Courtiol"
date: "`r Sys.Date()`"
output:
  html_vignette:
    toc: true
vignette: >
  %\VignetteIndexEntry{3.X Answers to Exercises: Let's practice more GLM}
  %\VignetteEncoding{UTF-8}
  %\VignetteEngine{knitr::rmarkdown}
---
```{r setup, include=FALSE}
library(LM2GLMM)
library(car)
knitr::opts_chunk$set(cache = FALSE, fig.align = "center", fig.width = 6, fig.height = 6)
```

## Disclamer

In this vignette I illustrate the key steps required to solve the exercises. By no means I am trying to provide a detailed report of the analysis of each dataset. Any such report would need additional analyses and results would have to be written in good English, not in telegraphic style or lines of codes. I have also written this vignette very quickly, so double check I did not mess anything up...


# Dataset: InsectSprays (difficulty = 1/5)

## Goal

* compare the effectiveness of various insecticides on the number of insects in agricultural experimental units
* compare the predictions from each insecticides from a GLM fit to those done while practicing the LM

Note: at this stage I will not check the assumptions and detail the dataset since we did it before and I will not look at the GLM assumptions since we have not yet seen this.


## LM model
```{r}
mod_insect <- lm(count ~ spray, data = InsectSprays)
```

## LM BoxCox
```{r}
InsectSprays$count_bis <- InsectSprays$count + 1
mod_insect_bis <- update(mod_insect, count_bis ~ .)
bc <- car::powerTransform(mod_insect_bis)
InsectSprays$count_bc <- car::bcPower(InsectSprays$count_bis, bc$lambda)
mod_insect_bc <- update(mod_insect_bis, count_bc ~ .)
```

## GLM
```{r}
mod_insect_glm <- glm(count ~ spray, data = InsectSprays, family = poisson())
```

## Tests comparing the fitted models to their respective null models
```{r}
anova(mod_insect, update(mod_insect, . ~ 1))
anova(mod_insect_bc, update(mod_insect_bc, . ~ 1))
anova(mod_insect_glm, update(mod_insect_glm, . ~ 1), test = "LR")
```

## Predictions
```{r}
data.for.pred <- data.frame(spray = levels(InsectSprays$spray))
pred_lm <- predict(mod_insect, newdata = data.for.pred, interval = "confidence")
pred_bc <- predict(mod_insect_bc, newdata = data.for.pred, interval = "confidence")
unboxcox <- function(x, lambda) (x*lambda + 1)^(1/lambda)
pred_bc_unboxcox <- unboxcox(x = pred_bc, lambda = bc$lambda) - 1  ## we remove one as we added one so to be able to do the BoxCox
pred_glm_eta <- predict(mod_insect_glm, newdata = data.for.pred, se.fit = TRUE)
pred_glm_table <- data.frame(fit = exp(pred_glm_eta$fit))
pred_glm_table$lwr <- exp(pred_glm_eta$fit + qnorm(0.025)*pred_glm_eta$se.fit)
pred_glm_table$upr <- exp(pred_glm_eta$fit + qnorm(0.975)*pred_glm_eta$se.fit)
```

## Plot
```{r}
plot(InsectSprays$count ~  as.numeric(InsectSprays$spray), xlim = c(0.5, 6.5), ylim = c(0, 30),
     ylab = "Predicted number of instects", xlab = "Sprays", axes = FALSE)
axis(2, las = 2)
axis(1, at = 1:6, labels = levels(InsectSprays$spray))
box()

points(pred_lm[, "fit"] ~  I(-0.2 + 1:6), col = "blue", pch = 20)

arrows(x0 = -0.2 + 1:6, x1 = -0.2 + 1:6, y0 = pred_lm[, "lwr"],
       y1 = pred_lm[, "upr"], code = 3, col = "blue", angle = 90, length = 0.1)

points(pred_bc_unboxcox[, "fit"] ~  I(-0.1 + 1:6), col = "green", pch = 20)

arrows(x0 = -0.1 + 1:6, x1 = -0.1 + 1:6, y0 = pred_bc_unboxcox[, "lwr"],
       y1 = pred_bc_unboxcox[, "upr"], code = 3, col = "green", angle = 90, length = 0.1)

points(pred_glm_table[, "fit"] ~  I(0.1 + 1:6), col = "red", pch = 20)

arrows(x0 = 0.1 + 1:6, x1 = 0.1 + 1:6, y0 = pred_glm_table[, "lwr"],
       y1 = pred_glm_table[, "upr"], code = 3, col = "red", angle = 90, length = 0.1)

legend("top", horiz = TRUE, fill = c("blue", "green", "red"), legend = c("LM", "LM + BoxCox", "GLM Poisson"), bty = "n")
```


# Classroom exercises


## Data
```{r}
set.seed(1L)
Aliens <- simulate_Aliens_GLM()
mod_binar <- glm(happy ~ humans_eaten, family = binomial(), data = Aliens)
mod_poiss <- glm(eggs  ~ humans_eaten, family = poisson(),  data = Aliens)
```

## Goal
 * plot the predictions and confidence intervals for ```mod_binar``` and ```mod_poiss``` for a number of humans eaten varying between 0 and 15.

## Solution for ```mod_binar```

### Confidence intervals 

```{r}
data.for.pred <- data.frame(humans_eaten = 0:15)
p <- predict(mod_binar, newdata = data.for.pred, se.fit = TRUE)
p$upr <- p$fit + qnorm(0.975) * p$se.fit
p$lwr <- p$fit + qnorm(0.025) * p$se.fit
p.mu <- data.frame(fit = plogis(p$fit),
                   upr = binomial()$linkinv(p$upr), ## same as plogis
                   lwr = 1/(1 + exp(-p$lwr))) ## same as plogis

plot(p.mu$fit ~ data.for.pred$humans_eaten, type = "l", lwd = 2, las = 1, ylim = c(0, 1),
     ylab = "Probability of being happy", xlab = "Number of humans eaten")
points(p.mu$upr ~  data.for.pred$humans_eaten, type = "l", lwd = 2, lty = 2)
points(p.mu$lwr ~  data.for.pred$humans_eaten, type = "l", lwd = 2, lty = 2)
```


### Confidence intervals using ```spaMM```

```{r, message = FALSE}
library(spaMM)
mod_binar_spaMM <- fitme(happy ~ humans_eaten, family = binomial(), data = Aliens)
p <- predict(mod_binar_spaMM, newdata = data.for.pred, intervals = "predVar")
plot(p ~ data.for.pred$humans_eaten, type = "l", lwd = 2, las = 1, ylim = c(0, 1),
     ylab = "Probability of being happy", xlab = "Number of humans eaten")
points(attr(p, "interval")[, 1] ~  data.for.pred$humans_eaten, type = "l", lwd = 2, lty = 2)
points(attr(p, "interval")[, 2] ~  data.for.pred$humans_eaten, type = "l", lwd = 2, lty = 2)
```


### Prediction intervals

```{r}
p <- predict(mod_binar, newdata = data.for.pred, se.fit = TRUE)
p$upr <- p$fit + qnorm(0.975) * sqrt(p$se.fit^2 + binomial()$variance(plogis(p$fit)))
p$lwr <- p$fit + qnorm(0.025) * sqrt(p$se.fit^2 + binomial()$variance(plogis(p$fit)))
p.mu <- data.frame(fit = plogis(p$fit),
                   upr = binomial()$linkinv(p$upr), ## same as plogis
                   lwr = 1/(1 + exp(-p$lwr))) ## same as plogis

plot(p.mu$fit ~ data.for.pred$humans_eaten, type = "l", lwd = 2, las = 1, ylim = c(0, 1),
     ylab = "Probability of being happy", xlab = "Number of humans eaten")
points(p.mu$upr ~  data.for.pred$humans_eaten, type = "l", lwd = 2, lty = 2)
points(p.mu$lwr ~  data.for.pred$humans_eaten, type = "l", lwd = 2, lty = 2)
```


### Prediction intervals with ```spaMM```

```{r}
p2 <- predict(mod_binar_spaMM, newdata = data.for.pred, intervals = "respVar")
plot(p2 ~ data.for.pred$humans_eaten, type = "l", lwd = 2, las = 1, ylim = c(0, 1),
     col = NULL, ylab = "Probability of being happy", xlab = "Number of humans eaten")
lwr <- attr(p2, "interval")[, 1]
upr <- attr(p2, "interval")[, 2]
polygon(y = c(lwr, rev(upr), lwr[1]),
        x = c(0:15, 15:0, 0), col = "green", border = NA)
points(p2 ~  data.for.pred$humans_eaten, type = "l", lwd = 2)
```


## Solution for ```mod_poiss```

### Condidence intervals

```{r}
p <- predict(mod_poiss, newdata = data.for.pred, se.fit = TRUE)
p$upr <- p$fit + qnorm(0.975) * p$se.fit
p$lwr <- p$fit + qnorm(0.025) * p$se.fit
p.mu <- data.frame(fit = exp(p$fit),
                   upr = poisson()$linkinv(p$upr), ## same as exp
                   lwr = mod_poiss$family$linkinv(p$lwr)) ## same as exp
plot(p.mu$fit ~ data.for.pred$humans_eaten, type = "l", lwd = 2, las = 1,
     ylim = range(Aliens$eggs),
     ylab = "Number of eggs", xlab = "Number of humans eaten")
points(p.mu$upr ~  data.for.pred$humans_eaten, type = "l", lwd = 2, lty = 2)
points(p.mu$lwr ~  data.for.pred$humans_eaten, type = "l", lwd = 2, lty = 2)
points(jitter(Aliens$eggs) ~ Aliens$humans_eaten, cex = 0.5)
rug(x = jitter(Aliens$humans_eaten), col = "red", side = 3)
```

Let's do the same with packages:

```{r, message = FALSE}
library(effects)
plot(allEffects(mod_poiss))

library(visreg)
visreg(mod_poiss, scale = "response")
```

### Prediction intervals

```{r}
p$upr <- p$fit + qnorm(0.975) * sqrt(p$se.fit^2 + poisson()$variance(exp(p$fit)))
p$lwr <- p$fit + qnorm(0.025) * sqrt(p$se.fit^2 + poisson()$variance(exp(p$fit)))
p.mu <- data.frame(fit = exp(p$fit),
                   upr = poisson()$linkinv(p$upr), ## same as exp
                   lwr = mod_poiss$family$linkinv(p$lwr)) ## same as exp
plot(p.mu$fit ~ data.for.pred$humans_eaten, type = "l", lwd = 2, las = 1,
     ylim = range(Aliens$eggs),
     ylab = "Number of eggs", xlab = "Number of humans eaten")
points(p.mu$upr ~  data.for.pred$humans_eaten, type = "l", lwd = 2, lty = 2)
points(p.mu$lwr ~  data.for.pred$humans_eaten, type = "l", lwd = 2, lty = 2)
points(jitter(Aliens$eggs) ~ Aliens$humans_eaten, cex = 0.5)
rug(x = jitter(Aliens$humans_eaten), col = "red", side = 3)
```




# Dataset: HSE98women, exercise menopause

## Goal

* study the influences of age, body mass index and smoking status upon the probability of being menopaused
* estimate the proportion of menopaused women in the population, depending on their age (40, 45, 50, 55, 60 yrs)

Note: here, for the sake of simplicity we simply define menopause as the lack of periods in old ages.


## Data preparation and exploration

We create the response variable:
```{r}
HSE98women$menopause <-  HSE98women$period == FALSE
```

We fit a first model to help us easily exploring the data used for fitting and not all data (there are many missing values in the dataset).

```{r}
mod_menop_logit <- glm(menopause ~ age + bmi + smoked, data = HSE98women, family = binomial())
nrow(HSE98women) ## original number of observations
nrow(mod_menop_logit$model)  ## number of observations considered in the model
```

We compare the variables before and after filtering to see if any obvious biais is introduced by the observations we discared:
```{r}
summary(HSE98women[, c("menopause", "age", "bmi", "smoked")])
summary(mod_menop_logit$model)
```

The only big discrepancy between the two datasets is that the minimal age is now 18 yrs old while the original dataset also contains children. That suggests that girls too young to have periods are not being considered, which will greatly help us to model menopause.

Let's check the correlation between our two continuous predictor:

```{r}
cor.test(HSE98women$age, HSE98women$bmi)
cor.test(mod_menop_logit$model$age, mod_menop_logit$model$bmi)
```

The correlation is not very high, especially in the dataset that is used for the fit, so multicollinearity should not be an issue.

We visualise the relationship between the predictor and the response:
```{r}
library(car)
scatterplot(menopause ~ age + smoked, data = mod_menop_logit$model)
scatterplot(menopause ~ bmi + smoked, data = mod_menop_logit$model)
```
There don't seem to be any obvious interaction between the age and smoked or between BMI and smoked, so we won't consider any. We will however consider that the effect of age may depend on the BMI because while old individuals should be menopaused no matter their BMI, we may expect that obesity play a role early on.


## Fitting the models

```{r}
mod_menop_logit   <- glm(menopause ~ age * bmi + smoked, data = HSE98women, family = binomial())
mod_menop_probit  <- glm(menopause ~ age * bmi + smoked, data = HSE98women, family = binomial(link = "probit"))
mod_menop_cauchit <- glm(menopause ~ age * bmi + smoked, data = HSE98women, family = binomial(link = "cauchit"))
```

Because it is very difficult to explore the shape of the relationship between the age and bmi upon the probability of being menopaused, we will also fit general additive models (GAM), which allows parameters to vary along the predictors.

```{r, message=FALSE}
library(mgcv)
mod_menop_logit_gam   <- gam(menopause ~ s(age, bmi) + smoked, data = HSE98women, family = binomial())
mod_menop_probit_gam  <- gam(menopause ~ s(age, bmi) + smoked, data = HSE98women, family = binomial(link = "probit"))
mod_menop_cauchit_gam <- gam(menopause ~ s(age, bmi) + smoked, data = HSE98women, family = binomial(link = "cauchit"))
```

```{r}
cbind(logit = AIC(mod_menop_logit),
      probit = AIC(mod_menop_probit),
      cauchit = AIC(mod_menop_cauchit),
      logit_GAM = AIC(mod_menop_logit_gam),
      probit_GAM = AIC(mod_menop_probit_gam),
      cauchit_GAM = AIC(mod_menop_cauchit_gam)
      )
```

The GAM logit model seems to be the best model, which will prevent us to express simply the effect of coefficients compared to a usual GLM. That's too bad but as the model is much better, we will go for it.

## Tests

We are going to test for the effect of the interaction by Likelihood ratio test:
```{r}
mod_menop_logit_gam_no_int <- gam(menopause ~ s(age) + s(bmi) + smoked, data = HSE98women, family = binomial())
LR <- -2 * (logLik(mod_menop_logit_gam_no_int)[[1]] - logLik(mod_menop_logit_gam)[[1]])
DF <- (mod_menop_logit_gam$df.null - mod_menop_logit_gam$df.residual) - (mod_menop_logit_gam_no_int$df.null - mod_menop_logit_gam_no_int$df.residual)
(PV <- 1 - pchisq(LR, DF))
```

The interaction between age and BMI is not significant (LRT, LR = ```r round(LR, 1)```, df = ```r round(DF, 2)```, p-value = ```r round(PV, 2)```).

Let's now test the effect of BMI by Likelihood ratio test:

```{r}
mod_menop_logit_gam_no_bmi <- gam(menopause ~ s(age) + smoked, data = HSE98women, family = binomial())
LR <- -2 * (logLik(mod_menop_logit_gam_no_bmi)[[1]] - logLik(mod_menop_logit_gam)[[1]])
DF <- (mod_menop_logit_gam$df.null - mod_menop_logit_gam$df.residual) - (mod_menop_logit_gam_no_bmi$df.null - mod_menop_logit_gam_no_bmi$df.residual)
(PV <- 1 - pchisq(LR, DF))
```

The effect of BMI is strongly significant (LRT, LR = ```r round(LR, 1)```, df = ```r round(DF, 2)```, p-value < 0.001).

Note that models with GAM have non integer degrees of freedom. This is normal.

Let's now test the effect of age by Likelihood ratio test:

```{r}
mod_menop_logit_gam_no_age <- gam(menopause ~ s(bmi) + smoked, data = HSE98women, family = binomial())
LR <- -2 * (logLik(mod_menop_logit_gam_no_age)[[1]] - logLik(mod_menop_logit_gam)[[1]])
DF <- (mod_menop_logit_gam$df.null - mod_menop_logit_gam$df.residual) - (mod_menop_logit_gam_no_age$df.null - mod_menop_logit_gam_no_age$df.residual)
(PV <- 1 - pchisq(LR, DF))
```

The effect of age is strongly significant (LRT, LR = ```r round(LR, 1)```, df = ```r round(DF, 2)```, p-value < 0.001).

Let's now test the effect of smoking by Likelihood ratio test:

```{r}
mod_menop_logit_gam_never_smoked <- gam(menopause ~ s(age, bmi), data = HSE98women, family = binomial())
LR <- -2 * (logLik(mod_menop_logit_gam_never_smoked)[[1]] - logLik(mod_menop_logit_gam)[[1]])
DF <- (mod_menop_logit_gam$df.null - mod_menop_logit_gam$df.residual) - (mod_menop_logit_gam_never_smoked$df.null - mod_menop_logit_gam_never_smoked$df.residual)
(PV <- 1 - pchisq(LR, DF))
```

The effect of have smooked is also significant (LRT, LR = ```r round(LR, 1)```, df = ```r round(DF, 2)```, p-value < 0.001).


## Predictions

Let's first explore the model output with the pacakge ```visreg```, which also works on GAM models:

```{r}
library(visreg)
visreg(mod_menop_logit_gam)
```

These plots show the patterns identified by the smoothing parameters in the GAM: the effect of age seems to change after 40 yrs and the effect of BMI seem to be changing between 23 and 32 yrs. It is difficult to account for such threshold effects by polynoms, so we will have to stick to GAM here...

Le't us do the same plots on the scale of the response variable:
```{r}
visreg(mod_menop_logit_gam, scale = "response")
```

We will now plot the predictions for the model with the logit link, the model with the cauchit link and the model with the GAM model with the logit link to explore their differences. We will predict the influence of age and consider the median BMI and smokers.

```{r}
age.for.pred <- seq(min(mod_menop_cauchit_gam$model$age), max(mod_menop_cauchit_gam$model$age), length = 100)
data.for.pred <- expand.grid(
  age = age.for.pred,
  bmi = median(mod_menop_cauchit_gam$model$bmi),
  smoked = TRUE
  )

  gam.p <- predict(mod_menop_logit_gam, newdata = data.for.pred, se.fit = TRUE)
gam.upr <- binomial(link = "logit")$linkinv(gam.p$fit + qnorm(0.975) * gam.p$se.fit)
gam.lwr <- binomial(link = "logit")$linkinv(gam.p$fit + qnorm(0.025) * gam.p$se.fit)
gam.fit <- binomial(link = "logit")$linkinv(gam.p$fit)

cauchit.p <- predict(mod_menop_cauchit, newdata = data.for.pred, se.fit = TRUE)
cauchit.upr <- binomial(link = "cauchit")$linkinv(cauchit.p$fit + qnorm(0.975) * cauchit.p$se.fit)
cauchit.lwr <- binomial(link = "cauchit")$linkinv(cauchit.p$fit + qnorm(0.025) * cauchit.p$se.fit)
cauchit.fit <- binomial(link = "cauchit")$linkinv(cauchit.p$fit)

logit.p <- predict(mod_menop_logit, newdata = data.for.pred, se.fit = TRUE)
logit.upr <- binomial(link = "logit")$linkinv(logit.p$fit + qnorm(0.975) * logit.p$se.fit)
logit.lwr <- binomial(link = "logit")$linkinv(logit.p$fit + qnorm(0.025) * logit.p$se.fit)
logit.fit <- binomial(link = "logit")$linkinv(logit.p$fit)

plot(gam.fit ~ age.for.pred, type = "l", las = 1, ylab = "Probability of being menopaused", xlab = "Age (yrs)", ylim = c(0, 1))
points(gam.upr ~ age.for.pred, type = "l", lty = 2)
points(gam.lwr ~ age.for.pred, type = "l", lty = 2)

points(cauchit.fit ~ age.for.pred, type = "l", lty = 1, col = "green")
points(cauchit.upr ~ age.for.pred, type = "l", lty = 2, col = "green")
points(cauchit.lwr ~ age.for.pred, type = "l", lty = 2, col = "green")

points(logit.fit ~ age.for.pred, type = "l", lty = 1, col = "red")
points(logit.upr ~ age.for.pred, type = "l", lty = 2, col = "red")
points(logit.lwr ~ age.for.pred, type = "l", lty = 2, col = "red")

legend("topleft", fill = c("black", "green", "red"), legend = c("logit GAM", "cauchit", "logit"), bty = "n")
```

Selecting the best model, let's now look at the effect of BMI, still using smokers:

```{r}
data.for.pred <- expand.grid(
  age = age.for.pred,
  bmi = c(18.5, 25, 30),
  smoked = TRUE
  )

gam.skiny.p <- predict(mod_menop_logit_gam, newdata = subset(data.for.pred, bmi == 18.5), se.fit = TRUE)
gam.skiny.upr <- binomial(link = "logit")$linkinv(gam.skiny.p$fit + qnorm(0.975) * gam.skiny.p$se.fit)
gam.skiny.lwr <- binomial(link = "logit")$linkinv(gam.skiny.p$fit + qnorm(0.025) * gam.skiny.p$se.fit)
gam.skiny.fit <- binomial(link = "logit")$linkinv(gam.skiny.p$fit)

gam.overweight.p <- predict(mod_menop_logit_gam, newdata = subset(data.for.pred, bmi == 25), se.fit = TRUE)
gam.overweight.upr <- binomial(link = "logit")$linkinv(gam.overweight.p$fit + qnorm(0.975) * gam.overweight.p$se.fit)
gam.overweight.lwr <- binomial(link = "logit")$linkinv(gam.overweight.p$fit + qnorm(0.025) * gam.overweight.p$se.fit)
gam.overweight.fit <- binomial(link = "logit")$linkinv(gam.overweight.p$fit)

gam.obese.p <- predict(mod_menop_logit_gam, newdata = subset(data.for.pred, bmi == 30), se.fit = TRUE)
gam.obese.upr <- binomial(link = "logit")$linkinv(gam.obese.p$fit + qnorm(0.975) * gam.obese.p$se.fit)
gam.obese.lwr <- binomial(link = "logit")$linkinv(gam.obese.p$fit + qnorm(0.025) * gam.obese.p$se.fit)
gam.obese.fit <- binomial(link = "logit")$linkinv(gam.obese.p$fit)

plot(gam.skiny.fit ~ age.for.pred, type = "l", las = 1, ylab = "Probability of being menopaused", xlab = "Age (yrs)", ylim = c(0, 1))
points(gam.skiny.upr ~ age.for.pred, type = "l", lty = 2)
points(gam.skiny.lwr ~ age.for.pred, type = "l", lty = 2)

points(gam.overweight.fit ~ age.for.pred, type = "l", lty = 1, col = "green")
points(gam.overweight.upr ~ age.for.pred, type = "l", lty = 2, col = "green")
points(gam.overweight.lwr ~ age.for.pred, type = "l", lty = 2, col = "green")

points(gam.obese.fit ~ age.for.pred, type = "l", lty = 1, col = "red")
points(gam.obese.upr ~ age.for.pred, type = "l", lty = 2, col = "red")
points(gam.obese.lwr ~ age.for.pred, type = "l", lty = 2, col = "red")

legend("topleft", fill = c("black", "green", "red"), legend = c("skiny", "overweight", "obese"), bty = "n")
```

We can see on this plot that obesity seems to accelerate the onset of menopause, but it does not seem to make any difference later on.

We will now predict the proportion of women in menopause at 40, 45, 50, 55 and 60 yrs. To illustrate the effect ofthe BMI will will perform prediction for a BMI of 21.75 (middle of normal BMI range according to WHO) and 30 (lower limit for obesity) Kg/(m^2). Again, we will perform the predictions considering smoker since we have data for this category. We will also compute the confidence intervals (prediction intervals make little sense in the context of the prediction of proportion...).

```{r}
data.for.pred <- expand.grid(
  age = c(40, 45, 50, 55, 60),
  bmi = c(21.75, 30),
  smoked = TRUE
  )
pred_normal <- predict(mod_menop_logit_gam, newdata = subset(data.for.pred, bmi == 21.75), se.fit = TRUE)
tab_normal <- cbind(
  predict = plogis(pred_normal$fit),
  lwr.CI = plogis(pred_normal$fit + qnorm(0.025)*pred_normal$se.fit),
  upr.CI = plogis(pred_normal$fit + qnorm(0.975)*pred_normal$se.fit)
)
rownames(tab_normal) <- c(40, 45, 50, 55, 60)
round(tab_normal, 2)
```

This gives the information for individuals for normal BMI. Let's now look at the corresponding proportion for the obeses:

```{r}
pred_obses <- predict(mod_menop_logit_gam, newdata = subset(data.for.pred, bmi == 30), se.fit = TRUE)
tab_obses <- cbind(
  predict = plogis(pred_obses$fit),
  lwr.CI = plogis(pred_obses$fit + qnorm(0.025)*pred_obses$se.fit),
  upr.CI = plogis(pred_obses$fit + qnorm(0.975)*pred_obses$se.fit)
)
rownames(tab_obses) <- c(40, 45, 50, 55, 60)
round(tab_obses, 2)
```

These tables confirm our expectation concerning the influenc of BMI upon the onset of menopause.

For the sake of comparion, let's compare the first table to what we would have predicted using the cauchit model or the logit GLM model:

```{r}
data.for.pred <- expand.grid(age = c(40, 45, 50, 55, 60), bmi = 21.75, smoked = TRUE)
pred_GAM <- predict(mod_menop_logit_gam, newdata = subset(data.for.pred, bmi == 21.75), se.fit = TRUE)
pred_logit <- predict(mod_menop_logit, newdata = subset(data.for.pred, bmi == 21.75), se.fit = TRUE)
pred_cauchit <- predict(mod_menop_cauchit, newdata = subset(data.for.pred, bmi == 21.75), se.fit = TRUE)
tab <- cbind(
  logit_GAM = plogis(pred_GAM$fit),
  logit_GLM = plogis(pred_logit$fit),
  cauchit_GLM = binomial(link = "cauchit")$linkinv(pred_cauchit$fit)
)
rownames(tab) <- c(40, 45, 50, 55, 60)
round(tab, 2)
```

Let's look at the same proportion in the raw data. We cannot be as strict otherwise we won't get much individuals, so let's enlarge a bit the selection and see that we get a few hundred individuals at least:
```{r}
tab <- with(HSE98women,
     rbind(sum(age >= 39 & age <= 41 & bmi > 18.5 & bmi < 25, na.rm = TRUE),
           sum(age >= 44 & age <= 46 & bmi > 18.5 & bmi < 25, na.rm = TRUE),
           sum(age >= 49 & age <= 51 & bmi > 18.5 & bmi < 25, na.rm = TRUE),
           sum(age >= 54 & age <= 56 & bmi > 18.5 & bmi < 25, na.rm = TRUE),
           sum(age >= 59 & age <= 61 & bmi > 18.5 & bmi < 25, na.rm = TRUE)
           )
)
rownames(tab) <- c(40, 45, 50, 55, 60)
tab
```
That seems a good selection process (+/- 1 year and the whole WHO normal BMI category), so let's check the proportion of menopaused individuals in each of these categories (keeping both smokers and non smokers as the effect is rather small, see later):

```{r}
tab <- with(HSE98women, 
     rbind(
       mean(menopause[age >= 39 & age <= 41 & bmi > 18.5 & bmi < 25], na.rm = TRUE),
       mean(menopause[age >= 44 & age <= 46 & bmi > 18.5 & bmi < 25], na.rm = TRUE),
       mean(menopause[age >= 49 & age <= 51 & bmi > 18.5 & bmi < 25], na.rm = TRUE),
       mean(menopause[age >= 54 & age <= 56 & bmi > 18.5 & bmi < 25], na.rm = TRUE),
       mean(menopause[age >= 59 & age <= 61 & bmi > 18.5 & bmi < 25], na.rm = TRUE)
     )
)
rownames(tab) <- c(40, 45, 50, 55, 60)
round(tab, 2)
```

These results are roughly consistent with the predictions from the GAM model. It confirms that the GLM logit and cauchit are underestimating the prevalence of menopause at old age. All models however may overestimate the prevalence of menaupose at young age. Depending on the goal of the study, other link functions, not implemented in R base, could be tried but it is quite advanced so we won't do that.

We can now turn to the exploration of the smoking effect by comparing predictions between smokers and non smokers at median BMI:

Selecting the best model, let's now look at the effect of BMI, still using smokers:

```{r}
data.for.pred.nosmoke <- expand.grid(
  age = age.for.pred,
  bmi = median(mod_menop_cauchit_gam$model$bmi),
  smoked = FALSE
  )

gam.p.nosmoke <- predict(mod_menop_logit_gam, newdata = data.for.pred.nosmoke, se.fit = TRUE)
gam.upr.nosmoke <- binomial(link = "logit")$linkinv(gam.p.nosmoke$fit + qnorm(0.975) * gam.p.nosmoke$se.fit)
gam.lwr.nosmoke <- binomial(link = "logit")$linkinv(gam.p.nosmoke$fit + qnorm(0.025) * gam.p.nosmoke$se.fit)
gam.fit.nosmoke <- binomial(link = "logit")$linkinv(gam.p.nosmoke$fit)


plot(gam.fit ~ age.for.pred, type = "l", las = 1, ylab = "Probability of being menopaused", xlab = "Age (yrs)", ylim = c(0, 1), col = "red")
points(gam.upr ~ age.for.pred, type = "l", lty = 2, col = "red")
points(gam.lwr ~ age.for.pred, type = "l", lty = 2, col = "red")

points(gam.fit.nosmoke ~ age.for.pred, type = "l", col = "green")
points(gam.upr.nosmoke ~ age.for.pred, type = "l", lty = 2, col = "green")
points(gam.lwr.nosmoke ~ age.for.pred, type = "l", lty = 2, col = "green")

legend("topleft", fill = c("green", "red"), legend = c("non-smoker", "smoker"), bty = "n")
```

Although significant, the effect of smoking is rather small. Because the interaction between the smoking status and the BMI was not significant, we won't draw predictions to explore that.
