---
title: "Solutions for the exercises with LM"
author: "Alexandre Courtiol"
date: "`r Sys.Date()`"
output: 
  html_vignette:
    toc: true
vignette: >
  %\VignetteIndexEntry{Exercise LM: Let's practice more}
  %\VignetteEncoding{UTF-8}
  %\VignetteEngine{knitr::rmarkdown}
---
```{r setup, include=FALSE}
library(LM2GLMM)
library(car)
knitr::opts_chunk$set(cache = FALSE, fig.align = "center", fig.width = 6, fig.height = 6)
```

## Disclamer

In this vignette I illustrate the key steps required to solve the exercises. By no means I am trying to provide a detailed report of the analysis of each dataset. Any such report would need additional analyses and results would have to be written in good English, not in telegraphic style or lines of codes.


# Dataset: chickwts (difficulty = 0/5)

## Goal

* compare the effectiveness of various feed supplements on the weights of six weeks old chickens
* predict the proportion of chicken over 300 grams for each feed supplements

## Exploring the data

We always start by looking at the data we have:

```{r, fig.width = 8}
str(chickwts)
plot(weight ~ feed, data = chickwts)
```

I don't know much about chicken but the values seem alright.


We can also check the structure of the experimental design using the cross product of the design matrix we are planing to use for the fitting procedure:

```{r}
crossprod(model.matrix( ~ feed, data = chickwts))
```

Here, because there is only one factor involved, we could also have directly used the following:

```{r}
table(chickwts$feed)
```

We check if there is any missing data:

```{r}
any(is.na(chickwts))
```

Great there is no missing data.

## Fitting the model

We fit the model:

```{r}
(mod_chick <- lm(weight ~ feed, data = chickwts))
```

## Checking the assumptions

* linearity cannot be an issue here as we only have one qualitative predictor.
* multicollinearity is not an issue either has each chick has received a different feeding treatment (see cross-product above).
* fixed values for the predictor: unless the diet have not been recorded correctly, there should not been any problem.

Let's have a look at the residuals of the model:

```{r}
par(mfrow = c(2, 2))
plot(mod_chick)
```

Everything looks very good. The errors should thus be independent, homoscedastic and normally distributed.


## Testing the effect of diet

```{r}
anova(mod_chick)
```

The diet seems to influence significantly the growth of chicken (F = ```r round(anova(mod_chick)$F[1], 2)```, df1 = ```r anova(mod_chick)$Df[1]```, df2 = ```r anova(mod_chick)$Df[2]```, p-value < 0.001).

## Predicting the mean effect of each diet

```{r}
pred <- predict(mod_chick, newdata = data.frame(feed = levels(chickwts$feed)),
        interval = "confidence")
rownames(pred) <- levels(chickwts$feed)
pred
```

Let's plot this:

```{r, results="hide"}
plot(pred[, "fit"] ~ I(1:nrow(pred) + 0.05), axes = FALSE, pch = 20,
     ylim = range(mod_chick$model$weight),
     ylab = "Chicken weight (gr)", xlab = "Diet")
axis(1, 1:nrow(pred), labels = rownames(pred))
axis(2)
box()
for (row in 1:nrow(pred)) {
  arrows(x0 = row + 0.05, y0 = pred[row, "lwr"], x1 = row + 0.05, y1 = pred[row, "upr"],
         code = 3, angle = 90, length = 0.1)
  diet <- levels(chickwts$feed)[row]
  weights.to.plot <- chickwts$weight[chickwts$feed == diet]
  points(rep(row, length(weights.to.plot)) - 0.05, weights.to.plot, col = "grey")
  }
```

The plot shows the predicted mean weight of chicken for each diet (black) with associated 95% confidence intervals. The observed weights of the chicken included in the analysis are represented by grey dots.

To make the plot even better we could have sorted the diet by increasing order of their effects.

## Testing which diet differs from each others

```{r, message = FALSE, fig.width = 8}
summary(posthoc <- multcomp::glht(mod_chick, linfct = multcomp::mcp(feed = "Tukey")))
par(oma = c(0, 8, 0, 0))
plot(posthoc)
```


## Predicting the proportion of chiken heavier than 300 grams

```{r}
pred2 <- predict(mod_chick,
                newdata = data.frame(feed = levels(chickwts$feed)),
                interval = "prediction",
                se.fit = TRUE)

se.pred <- sqrt(pred2$se.fit^2 + pred2$residual.scale^2)

freq_more300 <- pnorm(rep(300, 6),
                      mean = pred2$fit[, "fit"],
                      sd = se.pred, lower.tail = FALSE)

output <- cbind(pred2$fit, "%>300gr" = round(freq_more300, 2))

rownames(output) <- levels(chickwts$feed)

output
```

We can plot this: 

```{r, fig.width = 8}
barplot(sort(output[, "%>300gr"]), ylim = c(0, 1), ylab = "Predicted proportion of chicken larger than 300 gr")
```

We clean the session for the next exercise:

```{r}
rm(list = ls())
```


# Dataset: InsectSprays (difficulty = 2/5)

## Goal

* compare the effectiveness of various insecticides on the number of insects in agricultural experimental units
* what is the efficiency of spray C?

## Exploring the data

```{r}
str(InsectSprays)
range(InsectSprays$count)
table(InsectSprays$spray)
boxplot(count ~ spray, data = InsectSprays)
any(is.na(InsectSprays))
```


## Fitting the model

```{r}
(mod_insect <- lm(count ~ spray, data = InsectSprays))
```

## Checking the assumptions

As for the dataset ```chickwts```, because we only study the effect of one qualitative predictor, we expect linearity and multicollinearity, and that the values for the predictor are fixed as assumed. We can thus directly turn to the analysis of the prediction error.

```{r}
par(mfrow = c(2, 2))
plot(mod_insect)
```

That does not look so good. There seem to be some heteroscedasticity in the data. Let's test this possibility:

```{r}
lmtest::bptest(mod_insect)
```

The Breusch-Pagan test shows that we can reject the assumption of homoscedasticity. One option would be to try to use GLM but we have not yet seen this, so let's try to reduce the problems and apply a LM fit anyhow. The first thing top try to solve the problems is to use a power transformation of the response variable.

## Box Cox

For the Box Cox transformation to work, we need all the values to be strictly positive, but we saw that the minimum value is zero, so let's start by adding one to every count and let's refit the model.

```{r}
InsectSprays$count_bis <- InsectSprays$count + 1
mod_insect_bis <- update(mod_insect, count_bis ~ .)
```

We can now start the Box Cox analysis:

```{r}
car::boxCox(mod_insect_bis)
summary(bc <- car::powerTransform(mod_insect_bis))
```

The estimated value for lambda is close to 1/3 which corresponds to a cube root exponent to the observation but this is not particularly meaningfull so we will stick to the best estimate for lambda which is ```r bc$lambda``` and stored in the object bc$lambda.

Let's fit the model on the transformed data and display the new diagnostic plots:

```{r}
InsectSprays$count_bc <- car::bcPower(InsectSprays$count_bis, bc$lambda)
mod_insect_bc <- update(mod_insect_bis, count_bc ~ .)
par(mfrow = c(2, 2))
plot(mod_insect_bc)
```

That looks fantastic! The heteroscedasticity is gone:

```{r}
lmtest::bptest(mod_insect_bc)
```

The lack of dependence should not be an issue (although the original paper where these data have been published showed that the experiemental design was not as simple as the dataset here suggests), but let's check anyhow:

```{r}
lmtest::dwtest(mod_insect_bc)
car::durbinWatsonTest(mod_insect_bc)
```

Although surprising differences between the two implementation of this test, independence is not rejected.

The normality looked good on the plots but we could also test it:

```{r}
nortest::lillie.test(mod_insect_bc$residuals)
shapiro.test(mod_insect_bc$residuals)
```
Again, no issue here.

## Testing the effect of insecticides

```{r}
anova(mod_insect_bc)
```

The type of insecticide seems to influence significantly the number of insects counted (F = ```r round(anova(mod_insect_bc)$F[1], 2)```, df1 = ```r anova(mod_insect_bc)$Df[1]```, df2 = ```r anova(mod_insect_bc)$Df[2]```, p-value < 0.001).

## Predicting the effect of spray C

```{r}
(meanC_BC <- predict(mod_insect_bc, newdata = data.frame(spray = "C"), interval = "confidence"))
```
This corresponds to the number of insects + 1 on the Box Cox scale. Thus we need to transform this result back on the original scale:

```{r}
(meanC <- ((meanC_BC * bc$lambda) + 1)^(1/bc$lambda) - 1)
```
We thus expect to find on average (95%CI) ```r  round(meanC[1, "fit"], 2)``` (```r round(meanC[1, "lwr"], 2)``` -- ```r round(meanC[1, "upr"], 2)```) insects on agricultural experimental units treated with the insecticide C.


