---
title: "GLM: Residuals & Assumptions"
author: "Alexandre Courtiol"
date: "`r Sys.Date()`"
output:
  ioslides_presentation:
    widescreen: true
    smaller: true
vignette: >
  %\VignetteIndexEntry{3.2 Residuals & Assumptions}
  %\VignetteEncoding{UTF-8}
  %\VignetteEngine{knitr::rmarkdown}
---

```{r setup, include=FALSE}
library(LM2GLMM)
options(width = 120)
knitr::opts_chunk$set(cache = FALSE, fig.width = 5, fig.height = 5, fig.align = "center")
```

## You will learn in this session


## Our toy models

```{r}
set.seed(1L)
Aliens <- data.frame(humans_eaten = round(runif(n = 100, min = 0, max = 15)))
Aliens$size  <- rnorm( n = 100, mean = 50 + 1.5 * Aliens$humans_eaten, sd = 5)
Aliens$eggs  <- rpois( n = 100, lambda = exp(-1 + 0.1 * Aliens$humans_eaten))
Aliens$happy <- rbinom(n = 100, size = 1, prob = plogis(-3 + 0.3 * Aliens$humans_eaten))
Aliens$all_eyes  <- round(runif(nrow(Aliens), min = 1, max = 12))
Aliens$blue_eyes <- rbinom(n = nrow(Aliens), size = Aliens$all_eyes, prob = plogis(-2 + 0.5 * Aliens$humans_eaten))
Aliens$pink_eyes <- Aliens$all_eyes - Aliens$blue_eyes

mod_gauss <- glm(size  ~ humans_eaten, family = gaussian(), data = Aliens)
mod_poiss <- glm(eggs  ~ humans_eaten, family = poisson(),  data = Aliens)
mod_binar <- glm(happy ~ humans_eaten, family = binomial(), data = Aliens)
mod_binom <- glm(cbind(blue_eyes, pink_eyes) ~ humans_eaten, family = binomial(), data = Aliens)
```


# Residuals

## Many types of residuals can be computed

<div class="columns-2">

```{r}
rbind(residuals(mod_poiss, type = "deviance")[1:2],
      residuals(mod_poiss, type = "pearson")[1:2],
      residuals(mod_poiss, type = "working")[1:2],
      residuals(mod_poiss, type = "response")[1:2]
      )
```

```{r}
rbind(residuals(mod_binom, type = "deviance")[1:2],
      residuals(mod_binom, type = "pearson")[1:2],
      residuals(mod_binom, type = "working")[1:2],
      residuals(mod_binom, type = "response")[1:2]
      )
```

</div>

<br>

Note 1: we can also compute partial residuals, which we shall see later.

Note 2: this is also true for LM and Gaussian GLM, but it makes no difference in such case (except for partial residuals).

## Distribution

### They are not necessarily normaly distributed!

<div class="columns-2">

```{r, echo = FALSE}
par(las = 1)
plot(ecdf(residuals(mod_poiss, type = "working")), col = 3, main = "mod_poiss")
plot(ecdf(residuals(mod_poiss, type = "deviance")), col = 1, add = TRUE)
plot(ecdf(residuals(mod_poiss, type = "pearson")), col = 2, add = TRUE)
plot(ecdf(residuals(mod_poiss, type = "response")), col = 4, add = TRUE)
legend("bottomright", fill = 1:4, bty = "n", legend = c("deviance", "pearson", "working", "response"), title = "Type of residuals:")
```

```{r, echo = FALSE}
par(las = 1)
plot(ecdf(residuals(mod_binom, type = "working")), col = 3, main = "mod_binom")
plot(ecdf(residuals(mod_binom, type = "deviance")), col = 1, add = TRUE)
plot(ecdf(residuals(mod_binom, type = "pearson")), col = 2, add = TRUE)
plot(ecdf(residuals(mod_binom, type = "response")), col = 4, add = TRUE)
legend("topleft", fill = 1:4, bty = "n", legend = c("deviance", "pearson", "working", "response"), title = "Type of residuals:")
```

</div>


## Response residuals

These are the residuals that we saw in LM: observed - predicted. Simple but pretty much useless.

```{r}
rbind(residuals(mod_gauss, type = "response")[1:2],
      (mod_gauss$y - mod_gauss$fitted.values)[1:2])
rbind(residuals(mod_poiss, type = "response")[1:2],
      (mod_poiss$y - mod_poiss$fitted.values)[1:2])
rbind(residuals(mod_binom, type = "response")[1:2],
      (mod_binom$y - mod_binom$fitted.values)[1:2])
```


## Pearson residuals

These are response residuals scaled by the square root of the variance function and accounting for their prior weights. 

```{r}
variances <- poisson()$variance(mod_poiss$fitted.values)
rbind(residuals(mod_poiss, type = "pearson")[1:2],
      (residuals(mod_poiss, type = "response") * sqrt(mod_poiss$prior.weights / variances))[1:2])

variances <- binomial()$variance(mod_binom$fitted.values)
rbind(residuals(mod_binom, type = "pearson")[1:2],
      (residuals(mod_binom, type = "response") * sqrt(mod_binom$prior.weights / variances))[1:2])
```

These residuals are useful to check the quality of the fit.


## Deviance residuals

These are the residuals that are minimized after the fit.

Contrary to those computed by ```family()$dev.resids``` they are here not squared.

<br>

```{r}
residuals(mod_poiss, type = "deviance")[1:2]
rbind((residuals(mod_poiss, type = "deviance")[1:2])^2,
      with(mod_poiss, poisson()$dev.resids(y = y, mu = fitted.values, wt = prior.weights))[1:2])
```


## Deviance residuals

The sum of the squared deviance residuals gives the scaled residual deviance of the model fit:

```{r}
c(sum(residuals(mod_gauss, type = "deviance")^2), deviance(mod_gauss))
c(sum(residuals(mod_poiss, type = "deviance")^2), deviance(mod_poiss))
c(sum(residuals(mod_binom, type = "deviance")^2), deviance(mod_binom))
```

<br>

These residuals are also useful to check the quality of the fit.


## Working residuals

These are the residuals used during the last iteration of the iterative fitting procedure. Useless otherwise.

```{r}
rbind(mod_poiss$residuals[1:2],
      ((mod_poiss$y - mod_poiss$fitted.values)/poisson()$mu.eta(mod_poiss$linear.predictors))[1:2])

rbind(mod_binom$residuals[1:2],
      ((mod_binom$y - mod_binom$fitted.values)/binomial()$mu.eta(mod_binom$linear.predictors))[1:2])
```


## Partial residuals

These are residuals expressed at the level of each predictor.

```{r}
mod_UK <- glm(milk ~ drink + sex + cigarettes, data = UK[1:10, ], family = poisson())
residuals(mod_UK, type = "partial")
```

<br>

They are useful to check for departure from linearity.

Note: they can also be computed for LM.


## What you need to remember



# Table of content

## The Generalized Linear Model: GLM

* 3.0 [Introduction](./GLM_intro.html)
* 3.1 [Intervals & Tests](./GLM_intervals.html)
* 3.2 [Residuals & Assumptions](./GLM_assumptions.html)
* 3.3 [Overdispersion & Zero-inflation](./GLM_overdispersion.html)
* 3.4 [Let's practice more](./GLM_practice.html)

