---
title: "GLM: Let's practice"
author: "Alexandre Courtiol"
date: "`r Sys.Date()`"
output:
  ioslides_presentation:
    widescreen: true
    smaller: true
vignette: >
  %\VignetteIndexEntry{3.3 Practice}
  %\VignetteEncoding{UTF-8}
  %\VignetteEngine{knitr::rmarkdown}
---

```{r setup, include=FALSE}
library(LM2GLMM)
options(width = 100)
knitr::opts_chunk$set(cache = FALSE, fig.width = 5, fig.height = 5, fig.align = "center")
```

## Dataset: esoph

```{r}
head(esoph)
```

### Goal

* Find out whether you should rather limit alcohol consumption or tabacco consumption in order to avoid developing an oesophageal cancer? Does it depends on your age?


## Dataset: effects::TitanicSurvival

```{r}
data(TitanicSurvival, package = "effects")
head(TitanicSurvival)
```

### Goal

* Find out who mostly survived and who mostly died during the Titanic disaster of 1912.


## Dataset: LM2GLMM::UK

```{r}
UK[1:2, ]
```

### Goal

* Try to identify the determinant of the smoking behaviour of children.
* Try to identify the determinant of bronchitis in children.
* Try to identify the determinant of the variable ```backward```.


## Dataset: LM2GLMM::HSE98women

```{r}
head(HSE98women)
```

### Goal

* How does the probability of not having periods changes with age, smoking status and BMI?
* How does the probability of having been drunk recently changes with age?
* Does the number of drinking problems experienced during life depends on education?


# Table of content

## The Generalized Linear Model: GLM

* 3.0 [Introduction](./GLM_intro.html)
* 3.1 [Intervals & Tests](./GLM_intervals.html)
* 3.2 [Residuals & Assumptions](./GLM_assumptions.html)
* 3.3 [Let's practice more](./GLM_practice.html)
