---
title: "GLM: Let's practice!"
author: "Alexandre Courtiol"
date: "`r Sys.Date()`"
output:
  ioslides_presentation:
    widescreen: true
    smaller: true
vignette: >
  %\VignetteIndexEntry{3.1 GLM: Let's practice ----------------------------------------exercises}
  %\VignetteEncoding{UTF-8}
  %\VignetteEngine{knitr::rmarkdown}
---

```{r setup, include=FALSE}
library(LM2GLMM)
options(width = 100)
knitr::opts_chunk$set(cache = FALSE, fig.width = 5, fig.height = 5, fig.align = "center",
                      cache.path = "./cache_knitr/Exo_GLM/", fig.path = "./fig_knitr/Exo_GLM/")
```

## The Generalized Linear Model: GLM

* 3.0 [Introduction](./GLM_intro_cource.html)
* 3.1 [Let's practice](./GLM_practice_exercises.html)

<br>

<div align="right">
[Back to main menu](./Title.html#2)
</div>


## Dataset: InsectSprays (difficulty = 2/5) `r .emo("practice")`

```{r InsectSprays}
head(InsectSprays, n = 3)
levels(InsectSprays$spray)
```

<br>

`r .emo("goal")`  Test whether insecticide type significantly affects the number of insects in agricultural experimental units.

`r .emo("goal")` Compare the effect of spray C to all other spray types.

`r .emo("goal")` What is the mean number of insects we would predict to find on a plot treated with each insecticide spray. Compare results between the LM approach (with and without BoxCox transformation) to those from the GLM approach.


## Dataset: Surprise (difficulty = 0/5) `r .emo("practice")`

```{r}
head(Surprise)
```

<br>

`r .emo("goal")` Do children value more the type of the present, the cost of the present, or both?

## Dataset: esoph (difficulty = 3/5) `r .emo("practice")`

```{r}
head(esoph)
```

<br>

`r .emo("goal")` Would it be better to limit alcohol consumption or tobacco consumption in order to avoid developing an oesophageal cancer? Use a GLM to find out.


## Dataset: TitanicSurvival (difficulty = 3/5) `r .emo("practice")`

```{r}
head(TitanicSurvival)
```

<br>

`r .emo("goal")` Find out how the sex, age and passenger class influenced who died during the Titanic disaster of 1912.


## Dataset: Challenger (difficulty = 2/5) `r .emo("practice")`

```{r}
head(Challenger)
```

<br>

`r .emo("goal")` What was the probability of O-ring failure when the space shuttle Challenger took off on the 28th of January 1986 by 31 degrees F?


## Dataset: UK (difficulty = ?/5) `r .emo("practice")`

```{r}
UK[1:2, ]
```

<br>

`r .emo("goal")` Try to identify the influential determinants of the smoking behaviour of children.

`r .emo("goal")` Try to identify the influential determinants of bronchitis in children.

`r .emo("goal")` Try to identify the influential determinants of the variable ```backward```.


## Dataset: HSE98women (difficulty = ?/5) `r .emo("practice")`

```{r}
head(HSE98women)
```

<br>

`r .emo("goal")` Study the influences of age, body mass index and smoking status upon the probability of being menopaused.

`r .emo("goal")` Estimate the proportion of menopaused women in the population, depending on their age (40, 45, 50, 55, 60 yrs).

# Solutions

## Answers to Exercises

<br>

You can find the solution to all problems [here](./GLM_practice_solution.html).

Give it a good try before checking how I did it!


# Table of contents

## The Generalized Linear Model: GLM

* 3.0 [Introduction](./GLM_intro_cource.html)
* 3.1 [Let's practice](./GLM_practice_exercises.html)

<br>

<div align="right">
[Back to main menu](./Title.html#2)
</div>
