---
title: "GLM: Let's practice"
author: "Alexandre Courtiol"
date: "`r Sys.Date()`"
output:
  ioslides_presentation:
    widescreen: true
    smaller: true
vignette: >
  %\VignetteIndexEntry{3.3  Let's practice more GLMs}
  %\VignetteEncoding{UTF-8}
  %\VignetteEngine{knitr::rmarkdown}
---

```{r setup, include=FALSE}
library(LM2GLMM)
options(width = 100)
knitr::opts_chunk$set(cache = TRUE, fig.width = 5, fig.height = 5, fig.align = "center",
                      cache.path = "./cache_knitr/Exo_GLM/", fig.path = "./fig_knitr/Exo_GLM/")
```

## The Generalized Linear Model: GLM

* 3.0 [Introduction](./GLM_intro.html)
* 3.1 [Intervals & Tests](./GLM_intervals.html)
* 3.2 [Residuals & Assumptions](./GLM_assumptions.html)
* 3.3 [Let's practice more](./GLM_practice.html)

# Exercises

## Dataset: InsectSprays

```{r InsectSprays}
head(InsectSprays, n = 3)
levels(InsectSprays$spray)
```

### Goal

* Compare the effectiveness of various insecticides on the number of insects in agricultural experimental units.
* What is the efficiency of spray C?
* Compare results between the LM approach (with and without BoxCox transformation) and the GLM approach.


## Dataset: LM2GLMM::Surprise

```{r}
head(Surprise)
```

### Goal

* Do children value more the type of the present, the cost of the present, or both?

## Dataset: esoph

```{r}
head(esoph)
```

### Goal

* Find out whether you should rather limit alcohol consumption or tabacco consumption in order to avoid developing an oesophageal cancer?


## Dataset: carData::TitanicSurvival

```{r}
data(TitanicSurvival, package = "carData")
head(TitanicSurvival)
```

### Goal

* Find out who mostly survived and who mostly died during the Titanic disaster of 1912.
* Compute the odds-ratio for survival between the two most contrasting groups of individuals. 


## Dataset: LM2GLMM::Challenger

```{r}
head(Challenger)
```

### Goal

* What was the probability of O-ring failure when the space shuttle Challenger took off on the 28th of January 1986 by 31 degrees F?


## Dataset: LM2GLMM::UK

```{r}
UK[1:2, ]
```

### Goal

* Try to identify the influential determinants of the smoking behaviour of children.
* Try to identify the influential determinants of bronchitis in children.
* Try to identify the influential determinants of the variable ```backward```.


## Dataset: LM2GLMM::HSE98women

```{r}
head(HSE98women)
```

### Goal

* Study the influences of age, body mass index and smoking status upon the probability of being menopaused
* Estimate the proportion of menopaused women in the population, depending on their age (40, 45, 50, 55, 60 yrs)

## Dataset: MASS::mammals

```{r}
data(mammals, package = "MASS")
head(mammals)
```

### Goal

* What is the allometric exponent for the growth of brain size with body mass?
* Does this exponent agree with the usual expectation of 2/3?
* How large do we expect a 1kg animal to be?
* Rank organisms by relative brain size (i.e. controlled for body size)
* Repeat the analysis with the dataset ```MASS::Animals```


# Solutions

## Answers to Exercises

<br>

You can find the solution to all problems [here](./Exercise_Practice_GLM.html).

Give it a good try before checking how I did it!

# Table of contents

## The Generalized Linear Model: GLM

* 3.0 [Introduction](./GLM_intro.html)
* 3.1 [Intervals & Tests](./GLM_intervals.html)
* 3.2 [Residuals & Assumptions](./GLM_assumptions.html)
* 3.3 [Let's practice more](./GLM_practice.html)
