---
title: "GLM: Let's practice!"
subtitle: "Solutions"
author: "Alexandre Courtiol"
date: "`r Sys.Date()`"
output:
  html_vignette:
    toc: true
vignette: >
  %\VignetteEncoding{UTF-8}
  %\VignetteIndexEntry{3.1 GLM: Let's practice ----------------------------------------solutions}
  %\VignetteEngine{knitr::rmarkdown}
editor_options: 
  chunk_output_type: console
---
```{r setup, include=FALSE}
library(LM2GLMM)
knitr::opts_chunk$set(cache = TRUE, fig.align = "center", fig.width = 7, fig.height = 6,
                      cache.path = "./cache_knitr/Exo_GLM_solution/", fig.path = "./fig_knitr/Exo_GLM_solution/")
options(width = 100, rmarkdown.html_vignette.check_title = FALSE)
```

## Disclamer

In this vignette I illustrate the key steps required to solve the exercises. By no means I am trying to provide a detailed report of the analysis of each dataset. Any such report may need additional analyses and results would have to be written in good English, not in telegraphic style or lines of codes.

<br>

# Dataset: InsectSprays (difficulty = 2/5)

`r .emo("goal")`  Compare the effectiveness of various insecticides on the number of insects in agricultural experimental units.

`r .emo("goal")` What is the efficiency of spray C?

`r .emo("goal")` Compare results between the LM approach (with and without BoxCox transformation) and the GLM approach.

Note: I will not check the assumptions for LM, nor detail the dataset (and plotting variables) since we did it before (see [LM exercises](./LM_practice_solutions.html)).


## Fitting the models

We fit the LM:

```{r}
fit_insect <- lm(count ~ spray, data = InsectSprays)
```

We fit the LM on the BoxCox transformed response:

```{r}
InsectSprays$count_bis <- InsectSprays$count + 1
fit_insect_bis <- update(fit_insect, count_bis ~ .)
bc <- car::powerTransform(fit_insect_bis)
InsectSprays$count_bc <- car::bcPower(InsectSprays$count_bis, bc$lambda)
fit_insect_bc <- update(fit_insect_bis, count_bc ~ .)
```

We fit the GLM:

```{r}
fit_insect_glm <- glm(count ~ spray, data = InsectSprays, family = poisson())
```

## Checking the assumptions for GLM

All assumptions about model structure have been checked before (see exercises for LM).

We still have to test all assumptions about the errors.

```{r insect_glm_resid}
library(DHARMa)
insect_glm_resid <- simulateResiduals(fit_insect_glm, n = 1000)
plot(insect_glm_resid) ## many tests at once
```

This reveals one issue concerning the dispersion.
Let's thus look at this in more details:

```{r insect_glm_resid2}
testDispersion(insect_glm_resid)
```

The dispersion value is higher than expected, so there is slight overdispersion.

Let's attempt to solve this issue before testing other assumptions.

## Solving the overdispersion issue

Is the overdispersion caused by two many zeros?

```{r}
table(fit_insect_glm$model$count) ## we count the zeros and other counts
testZeroInflation(insect_glm_resid)
```

nope... the number of observed zeros matches the expectation of the model.

Let's try to fit an alternative model to account for the overdispersion:

```{r}
library(spaMM)
fit_insect_glm_nb <- fitme(count ~ spray, data = InsectSprays, family = spaMM::negbin())
rbind(deviance(fit_insect_glm), deviance(fit_insect_glm_nb))
```

The negative binomial model fits the data much better as shown by its smaller deviance.

Note that the prediction are the same under the 2 models:

```{r}
plot(predict(fit_insect_glm, type = "response"), fit_insect_glm$model$count, col = "red", lwd = 6)
abline(0, 1, col = "red", lwd = 6)
points(predict(fit_insect_glm_nb), fit_insect_glm_nb$data$count, col = "blue")
abline(0, 1, col = "blue")
```

The only thing that differs is the residual variance around those predictions (not shown).

Note also that the Gaussian models have a deviance expressed on another scale, which makes the comparison between models particularly difficult when using Gaussian models (but this is not the case here).

Let us stick to the negative binomial model as our best GLM!

## Re-testing assumptions for the fit of the negative binomial model

```{r insect_glm_nb_resid, warning = FALSE}
insect_glm_nb_resid   <- simulateResiduals(fit_insect_glm_nb, n = 1000)
testDispersion(insect_glm_nb_resid)
```

The dispersion issue is solved, but what about the other assumptions?

```{r insect_glm_nb_resid2}
plot(insect_glm_nb_resid)
testTemporalAutocorrelation(insect_glm_nb_resid, time = 1:nrow(model.frame(fit_insect_glm_nb)))
testTemporalAutocorrelation(insect_glm_nb_resid, time = order(fitted(fit_insect_glm_nb)))
testZeroInflation(insect_glm_nb_resid)
```

Everything looks great!

## Tests comparing the fitted models to their respective null models
```{r}
anova(fit_insect, update(fit_insect, . ~ 1))  ## F-test for LM
anova(fit_insect_bc, update(fit_insect_bc, . ~ 1))  ## F-test for LM with BoxCox transformation
lmtest::lrtest(fit_insect_glm, update(fit_insect_glm, . ~ 1)) ## LRT for GLM Poisson
anova(fit_insect_glm_nb, update(fit_insect_glm_nb, . ~ 1)) ## LRT for GLM NegBin
```

No matter the models, the effect of the insecticide is detected!

## Comparing effect of spray C to others

Let's illustrate how to do that for the two GLMs since they have been fitted with 2 different packages (**{stats}** and **{spaMM}** ):

### GLM Poisson

```{r}
library(multcomp)
comp_glm <- glht(fit_insect_glm, linfct = mcp(spray = c("C - A = 0",
                                                        "C - B = 0",
                                                        "C - D = 0",
                                                        "C - E = 0",
                                                        "C - F = 0")))
summary(comp_glm)
```

### GLM NegBin

```{r}
comp_glm_nb <- glht(fit_insect_glm_nb, linfct = mcp(spray = c("C - A = 0",
                                                              "C - B = 0",
                                                              "C - D = 0",
                                                              "C - E = 0",
                                                              "C - F = 0")),
                    coef. = fixef.HLfit) ## needed for glht to work with spaMM
summary(comp_glm_nb)
plot(comp_glm_nb)
```

The sprays C is the most effective one (fewer insects remaining) but we cannot exclude that E is equally effective. 

## Predictions and method comparison

We now compute the predictions and CI for the different models.

### LM

```{r}
data.for.pred <- data.frame(spray = levels(InsectSprays$spray))
pred_lm <- cbind(data.for.pred,
                 predict(fit_insect, newdata = data.for.pred, interval = "confidence"))
pred_lm
```

### LM with BoxCox

```{r}
pred_bc <- cbind(data.for.pred,
                 predict(fit_insect_bc, newdata = data.for.pred, interval = "confidence"))
f.inverse <- function(x) car::bcnPowerInverse(x, lambda = bc$lambda, gamma = 0) - 1
## Note: we removed 1 as we added 1 so to be able to do the BoxCox transformation
pred_bc$fit <- f.inverse(x = pred_bc$fit)
pred_bc$lwr <- f.inverse(x = pred_bc$lwr) 
pred_bc$upr <- f.inverse(x = pred_bc$upr)
pred_bc
```

### GLM Poisson

```{r}
pred_glm_poiss_eta <- predict(fit_insect_glm, newdata = data.for.pred, se.fit = TRUE)
pred_glm_poiss <- cbind(data.for.pred, data.frame(fit = exp(pred_glm_poiss_eta$fit)))
pred_glm_poiss$lwr <- exp(pred_glm_poiss_eta$fit + qnorm(0.025)*pred_glm_poiss_eta$se.fit)
pred_glm_poiss$upr <- exp(pred_glm_poiss_eta$fit + qnorm(0.975)*pred_glm_poiss_eta$se.fit)
pred_glm_poiss
```

### GLM NegBin

```{r}
pred_glm_nb_table <- cbind(data.for.pred,
                           predict(fit_insect_glm_nb, newdata = data.for.pred),
                           get_intervals(fit_insect_glm_nb, newdata = data.for.pred,
                                         intervals = "predVar"))
colnames(pred_glm_nb_table)[-1] <- c("fit", "lwr", "upr")
pred_glm_nb_table
```

### Combining predicitions

We combine the predictions so as to be able to plot them easily:

```{r}
predictions_all <- rbind(cbind(pred_lm, method = "LM"),
                         cbind(pred_bc, method = "LM BoxCox"),
                         cbind(pred_glm_poiss, method = "GLM Poisson"),
                         cbind(pred_glm_nb_table, method = "GLM NegBin"))

## declare method as factor for better display
predictions_all$method <- forcats::fct_inorder(predictions_all$method)
```

### Plot

```{r}
library(ggplot2)
ggplot() +
  geom_dotplot(aes(y = count, x = spray),
               data = InsectSprays,
               binwidth = 0.5,
               binaxis = "y", stackdir = "center",
               fill = "white") +
  geom_point(aes(y = fit, x = spray, colour = method),
             data = predictions_all,
             position = position_dodge(width = 0.5)) + ## for spreading things appart
  geom_errorbar(aes(y = fit, x = spray, colour = method, ymin = lwr, ymax = upr),
                data = predictions_all, 
                position = position_dodge(width = 0.5)) +
  theme_classic()
```

Conclusions:

- GLM and LM not BoxCoxed fits yield similar point predictions.
- GLM and LM BoxCoxed fits yield similar CI.

<br>


# Dataset: LM2GLMM::Surprise (difficulty = 0/5)

`r .emo("goal")` Do children value more the type of the present, the cost of the present, or both?

## Solution

```{r, Surprise}
# coplot(happiness ~ price | type, data = Surprise)
```

<br>

# Dataset: esoph (difficulty = 3/5)

`r .emo("goal")` Find out whether you should rather limit alcohol consumption or tabacco consumption in order to avoid developing an oesophageal cancer?


## Exploration of the data

```{r}
esoph
str(esoph)
```

The factors are ordered which -- by default -- will select for different contrasts (polynomial contrasts), so we set them as unordered (note: polynomial contrasts are interesting to identify how to best model ordinal variables as quantitative variables):

```{r}
esoph$agegp_f <- factor(esoph$agegp, ordered = FALSE)
esoph$alcgp_f <- factor(esoph$alcgp, ordered = FALSE)
esoph$tobgp_f <- factor(esoph$tobgp, ordered = FALSE)
```


```{r}
summary(esoph)
esoph$cancer_freq <- esoph$ncases / (esoph$ncases + esoph$ncontrols)
esoph$cancer_N <- esoph$ncases + esoph$ncontrols
summary(esoph$cancer_freq) ## proportion of cancer
table(esoph$alcgp_f, esoph$tobgp_f)
table(esoph$alcgp_f, esoph$tobgp_f, esoph$agegp_f)
```

We plot the data:

```{r}
coplot(cancer_freq ~ tobgp_f | agegp_f, data = esoph)
coplot(cancer_freq ~ alcgp_f | agegp_f, data = esoph)
```

## We fit the models

Since the outcome is binomial, the only possibilities here are to explore which of the possible link functions fit the data best:

```{r}
fit_esoph_logit <- glm(cbind(ncases, ncontrols) ~ agegp_f + alcgp_f + tobgp_f, data = esoph,
                       family = binomial(link = "logit"))
fit_esoph_cauchit <- glm(cbind(ncases, ncontrols) ~ agegp_f + alcgp_f + tobgp_f, data = esoph,
                         family = binomial(link = "cauchit"))
fit_esoph_probit <- glm(cbind(ncases, ncontrols) ~ agegp_f + alcgp_f + tobgp_f, data = esoph,
                        family = binomial(link = "probit"))
c(fit_esoph_logit = deviance(fit_esoph_logit),
  fit_esoph_cauchit = deviance(fit_esoph_cauchit),
  fit_esoph_probit = deviance(fit_esoph_probit)) ## you could also use AIC
```

The probit link function leads to a model fitting the data best, so we will use it.


## Comparison to null model

```{r}
fit_esoph_probit_H0 <- glm(cbind(ncases, ncontrols) ~ 1, data = esoph,
                           family = binomial(link = "probit"))
lmtest::lrtest(fit_esoph_probit_H0, fit_esoph_probit)
```

The full model fits the data much better than the null model!

## Testing the model assumptions

Linearity is not an issue since we considered all predictors as factorial variables.
Collinearity should also not be an issue due to the neat sampling design.

```{r resid_esoph_probit}
esoph_probit_resid <- simulateResiduals(fit_esoph_probit, n = 1000)
plot(esoph_probit_resid)
testTemporalAutocorrelation(esoph_probit_resid, time = 1:nrow(esoph))
testTemporalAutocorrelation(esoph_probit_resid, time = fitted(fit_esoph_probit))
testZeroInflation(esoph_probit_resid)
testDispersion(esoph_probit_resid)
```

Everything seems in order.

## Testing the model effects

```{r}
car::Anova(fit_esoph_probit)
```

## Plotting the effects

Let us first try to plot the effects using functions from packages:

```{r}
library(effects)
plot(predictorEffects(fit_esoph_probit)) ## response scale
plot(predictorEffects(fit_esoph_probit, transformation = list(link = NULL, inverse = NULL))) ## linear scale
```

```{r}
library(visreg)
visreg(fit_esoph_probit, scale = "response", ylim  = c(0, 1))
visreg2d(fit_esoph_probit, "alcgp_f", "tobgp_f", plot.type = "image", scale = "response")
visreg2d(fit_esoph_probit, "alcgp_f", "tobgp_f", plot.type = "persp", scale = "response")
```

```{r}
fit_esoph_probit_spaMM <- fitme(cbind(ncases, ncontrols) ~ agegp_f + alcgp_f + tobgp_f,
                                data = esoph,
                                family = binomial(link = "probit"))
#plot_effects(fit_esoph_probit_spaMM, focal_var = "agegp_f") ## buggy in current spaMM
#plot_effects(fit_esoph_probit_spaMM, focal_var = "alcgp_f")
#plot_effects(fit_esoph_probit_spaMM, focal_var = "tobgp_f")

plot_effects_manual <- function(data) { 
  ggplot(data) +
     aes(y = pointp, x = forcats::fct_inorder(focal_var), ymax = up, ymin = low) +
     geom_point() +
     geom_errorbar(width = 0.1) +
     scale_y_continuous(limits = c(0, 1)) +
     theme_classic()
}

plot_effects_manual(pdep_effects(fit_esoph_probit_spaMM, focal_var = "agegp_f"))
plot_effects_manual(pdep_effects(fit_esoph_probit_spaMM, focal_var = "alcgp_f"))
plot_effects_manual(pdep_effects(fit_esoph_probit_spaMM, focal_var = "tobgp_f"))
```


Note: the 3 packages differ in how they compute predictions!

This is not part of the exercise, but let us compare precisely the predictions for the age group 25-34 according to all 3 packages to better characterise and understand the differences.

We first need a little data wrangling to create a table comparing the predictions across methods:

```{r}
effects_pred <- predictorEffect(fit_esoph_probit, predictor = "agegp_f")
effects_pred <- as.data.frame(summary(effects_pred)[c("effect", "lower", "upper")])[1, ]
colnames(effects_pred) <- c("fit", "lwr", "upr")

visreg_pred <- visreg(fit_esoph_probit, scale = "response", plot = FALSE)
visreg_pred <- visreg_pred[[1]]$fit[1, c("visregFit", "visregLwr", "visregUpr")]
colnames(visreg_pred) <- c("fit", "lwr", "upr")

spaMM_pred <- pdep_effects(fit_esoph_probit_spaMM, focal_var = "agegp_f")
spaMM_pred <- spaMM_pred[1, c("pointp", "low", "up")]
colnames(spaMM_pred) <- c("fit", "lwr", "upr")

rbind(effects = effects_pred,
      visreg = visreg_pred,
      spaMM = spaMM_pred)
```

- **{effects}** (under its default settings) computes partial-dependence effects (i.e. consider all values for non focal predictors, irrespective of the focal predictor, and compute the mean effect) and compute the mean BEFORE transforming the data back to the scale of the response:

```{r}
newdata <- data.frame(agegp_f = "25-34",
                      alcgp_f = esoph$alcgp_f,
                      tobgp_f = esoph$tobgp_f)
fit_esoph_probit$family$linkinv(mean(predict(fit_esoph_probit, newdata = newdata)))
```

- **{visreg}** (under its default settings) selects the level with the highest number of observation for the non focal factor (and when there are equal group sizes, it seems to pick the first level that comes out of the candidate ones; I did not check if that follows the order of the levels or their position in the data...).

```{r}
table(esoph$alcgp_f)
table(esoph$tobgp_f)
newdata <- data.frame(agegp_f = "25-34",
                      alcgp_f = "0-39g/day",
                      tobgp_f = "0-9g/day")
predict(fit_esoph_probit, newdata = newdata, type = "response")
```

- the function `pdep_effects()` from **{spaMM}** (under its default settings) computes partial-dependence effects (i.e. consider all values for non focal predictors, irrespective of the focal predictor, and compute the mean effect) and compute the mean AFTER transforming the data back to the scale of the response:

```{r}
newdata <- data.frame(agegp_f = "25-34",
                      alcgp_f = esoph$alcgp_f,
                      tobgp_f = esoph$tobgp_f)
mean(predict(fit_esoph_probit, newdata = newdata, type = "response"))
```

Note: doing predictions yourself is perhaps the easiest way to know what is actually happening!

## Predictions by hand

Doing things by hand is tedious but it gives you more control.

For example, here, to answer the question we may want to compare extreme situations:

```{r}
newdata_esoph_young <- expand.grid(agegp_f = c("25-34"),
                                   alcgp_f = c("0-39g/day", "120+"),
                                   tobgp_f = c("0-9g/day", "30+"))
pred_esoph_young <- predict(fit_esoph_probit, newdata = newdata_esoph_young,
                            type = "link", se.fit = TRUE)
newdata_esoph_young$pred <- binomial(link = "probit")$linkinv(pred_esoph_young$fit)
newdata_esoph_young$lwr <- binomial(link = "probit")$linkinv(pred_esoph_young$fit +
                            qnorm(0.025)*pred_esoph_young$se.fit)
newdata_esoph_young$upr <- binomial(link = "probit")$linkinv(pred_esoph_young$fit +
                            qnorm(0.975)*pred_esoph_young$se.fit)
newdata_esoph_young
```

Same for old people:

```{r}
newdata_esoph_old <- expand.grid(agegp_f = c("75+"),
                                 alcgp_f = c("0-39g/day", "120+"),
                                 tobgp_f = c("0-9g/day", "30+"))
pred_esoph_old <- predict(fit_esoph_probit, newdata = newdata_esoph_old,
                          type = "link", se.fit = TRUE)
newdata_esoph_old$pred <- binomial(link = "probit")$linkinv(pred_esoph_old$fit)
newdata_esoph_old$lwr <- binomial(link = "probit")$linkinv(pred_esoph_old$fit +
                            qnorm(0.025)*pred_esoph_old$se.fit)
newdata_esoph_old$upr <- binomial(link = "probit")$linkinv(pred_esoph_old$fit +
                            qnorm(0.975)*pred_esoph_old$se.fit)
newdata_esoph_old
```

So again, drinking seems to be a worse than smoking but we are extrapolating a bit since very few individuals correspond to the predicted conditions! (see `table()` call above).

One thing is clear, not smoking and not drinking seems to be the way to avoid these cancers.

## Interaction alcohol - tobacco?

We could also consider the interaction between alcohol and tobacco.

```{r}
fit_esoph_probit2a <- glm(cbind(ncases, ncontrols) ~ agegp_f + alcgp_f * tobgp_f,
                          data = esoph,
                          family = binomial(link = "probit"))
car::Anova(fit_esoph_probit2a)
summary(fit_esoph_probit2a)
```

For the ease of interpretation, when you have 2 factors in interaction you can also directly code their combination:

```{r}
esoph$drink_smoke <- factor(paste(esoph$alcgp_f, esoph$tobgp_f, sep = "_"))
fit_esoph_probit2b <- glm(cbind(ncases, ncontrols) ~ agegp_f + drink_smoke, data = esoph,
                          family = binomial(link = "probit"))
summary(fit_esoph_probit2b)
```

The fits `fit_esoph_probit2a` and `fit_esoph_probit2b` have the same likelihood and number of degrees of freedom.
They are equivalent, but just use a different parameterisation of the same thing.

The summary table from `fit_esoph_probit2b` now directly compares all drink - alcohol combination to the reference group with no cigarette and minimum alcohol consumption.

Ideally we should again check all assumptions (you should do it!).

Let us do the minimum here:

```{r fit_esoph_probit2 resid, warning = FALSE}
plot(simulateResiduals(fit_esoph_probit2b, n = 1000))
```

Let us now compare if it is better to drink a lot and not smoke or the reverse: 

```{r, message = FALSE}
library(multcomp)
posthoc <- glht(fit_esoph_probit2b, linfct = c("`drink_smoke120+_0-9g/day` - `drink_smoke0-39g/day_30+` == 0"))
par(oma = c(1, 5, 1, 1))
plot(posthoc)
confint(posthoc)
summary(posthoc, test = Chisqtest()) ## choose tests the better suited to your case
```

We have compared two different situations, using a posthoc approach.
Drinking a lot seems to be worse than smoking a lot for this particular cancer.

## Alternative: effects considered as linear

We could also have considered the predictor as quantitative predictors:

```{r}
fit_esoph_probit3 <- glm(cbind(ncases, ncontrols) ~ agegp_f + alcgp + tobgp, data = esoph,
                         family = binomial(link = "probit"))
summary(fit_esoph_probit3)
```

The test on the polynomial contrasts (automatic due to the variable being ordered factors) suggest that linear (`.L`) effects matter but that quadratic (`.Q`) effects don't. For alcohol, there are hints of a slight cubic (`.C`) effect, but we will ignore this for the sake of simplicity (the effect is small and in the absence of a quadratic effect, it is a little strange to interpret).

So let's recode the variable as truly quantitative and let's fit linear effects:

```{r}
esoph$alc_num <- unclass(esoph$alcgp)
esoph$tob_num <- unclass(esoph$tobgp)
fit_esoph_probit4 <- glm(cbind(ncases, ncontrols) ~ agegp_f + alc_num + tob_num, data = esoph,
                         family = binomial(link = "probit"))
plot(simulateResiduals(fit_esoph_probit4, n = 1000))
car::crPlots(fit_esoph_probit4) ## partial residuals (to check linearity, albeit somewhat redundant with polynomial contrasts)
rbind(factors = AIC(fit_esoph_probit),
      interaction = AIC(fit_esoph_probit2b),
      poly = AIC(fit_esoph_probit3),
      linear = AIC(fit_esoph_probit4))
confint(fit_esoph_probit4)
```

This last model seems quite good.
Again, our results suggest that alcohol matters most.
Do but bear in mind that it now compares the effect of increasing by one category in each predictor.


## More?

* One could also do a model with only age and tobacco, one with only age and alcohol and compare those by parametric bootstrap to get a p-value. But let us stop here for this exercise.
* One could explore interactions with the quantitative predictors.

<br>


# Dataset: TitanicSurvival (difficulty = 3/5)

`r .emo("goal")` Find out how the sex, age and passenger class influenced who died during the Titanic disaster of 1912.

## Exploration of the data
```{r}
head(TitanicSurvival)
str(TitanicSurvival)
summary(TitanicSurvival)
with(data = TitanicSurvival,
     tapply(survived, sex, function(x) mean(x == "yes")))
with(data = TitanicSurvival,
     tapply(survived, passengerClass, function(x) mean(x == "yes")))
with(data = TitanicSurvival,
     table(cut(age, breaks = c(0, 10, 25, 35, 50, Inf), include.lowest = TRUE), ## make groups
           sex, passengerClass))
```

## Visualization of the data
```{r}
coplot(survived ~ age | sex * passengerClass, data = TitanicSurvival, panel = panel.smooth)
library(car)
scatterplot(survived == "yes" ~ age, data = TitanicSurvival)
```

The effect of age does not seem linear...

## We recode some variables
```{r}
TitanicSurvival$surv <-  TitanicSurvival$survived == "yes"
TitanicSurvival$age_f <- cut(TitanicSurvival$age,
                             breaks = c(0, 10, 25, 35, 50, Inf), include.lowest = TRUE)
table(TitanicSurvival$age_f)
```


## Fitting the models

We fit a series of models with different links and considering age as factor or continuous:

```{r}
fit_logit <-   glm(surv ~ age_f + passengerClass + sex, data = TitanicSurvival,
                   family = binomial(link = "logit"))
fit_probit <-  glm(surv ~ age_f + passengerClass + sex, data = TitanicSurvival,
                   family = binomial(link = "probit"))
fit_cauchit <- glm(surv ~ age_f + passengerClass + sex, data = TitanicSurvival,
                   family = binomial(link = "cauchit"))

fit_logit2 <-   glm(surv ~ age + passengerClass + sex, data = TitanicSurvival,
                   family = binomial(link = "logit"))
fit_probit2 <-  glm(surv ~ age + passengerClass + sex, data = TitanicSurvival,
                   family = binomial(link = "probit"))
fit_cauchit2 <- glm(surv ~ age + passengerClass + sex, data = TitanicSurvival,
                   family = binomial(link = "cauchit"))

AIC(fit_logit, fit_probit, fit_cauchit, fit_logit2, fit_probit2, fit_cauchit2)
```

The cauchit model with age as a factor provides a much better fit than other models, so we will select that one.


## Checking the assumptions

```{r resid_Titanic}
library(DHARMa)
resid_Titanic <- simulateResiduals(fit_cauchit, n = 1000)
plot(resid_Titanic)
testTemporalAutocorrelation(resid_Titanic, time = order(fitted(fit_cauchit)))
testTemporalAutocorrelation(resid_Titanic, time = 1:nrow(model.frame(fit_cauchit)))
```

There is a little bit of serial autocorrelation but it is low and hard to understand (observation are sorted by alphabetical order) so we will ignore that for now.

There is no need to test for overdispersion in a binary model!


## We look at predictions

```{r}
data.for.pred <- with(data = TitanicSurvival,
                 expand.grid(age_f = levels(age_f),
                             passengerClass = levels(passengerClass),
                             sex = levels(sex)))
pred <- predict(fit_cauchit, newdata = data.for.pred, type = "link")
data.to.plot <- cbind(data.for.pred, pred = pred)
coplot(pred ~ unclass(age_f) | passengerClass + sex, data = data.to.plot)
library(lattice) ## same, more fancy
xyplot(pred ~ unclass(age_f) | passengerClass + sex, data = data.to.plot)
```

## GAM?

Another way to cope with the non-linearity of the age effect would be to do a GAM model:

```{r, gam_fit}
library(mgcv)
fit_cauchit_gam <- gam(surv ~ s(age) + passengerClass + sex,
                       data = TitanicSurvival,
                       family = binomial(link = "cauchit"))
AIC(fit_cauchit_gam)
```

The model is a bit better, we could use it.

## GAM model assumptions:

```{r cauchit_gam_resid}
cauchit_gam_resid <- simulateResiduals(fit_cauchit_gam)
plot(cauchit_gam_resid)
```
The QQ plot is good (left), but the uniformity plot on the right suggests things are not quite perfect.

Two predictors are factors the third is already modelled quite flexibly using GAM, so the issue is probably not about how we modelled each variable but more about something important we are missing.

Let's see if an interaction between `passengerClass` & `sex` would improve things:

```{r, gam_fit2}
fit_cauchit_gam2 <- gam(surv ~ s(age) + passengerClass*sex,
                        data = TitanicSurvival,
                        family = binomial(link = "cauchit"))
AIC(fit_cauchit_gam2)
cauchit_gam_resid2 <- simulateResiduals(fit_cauchit_gam2)
plot(cauchit_gam_resid2)
```

Everything is much better!
Ideally we should re-examine what link function is best but let's stick to what we have for simplicity.


## Tests

For GAM models, we need to use `anova()` (which calls `mgcv::anova.gam()` and thus does perform an type I anova but a marginal anova -- documentation says type III, so not type II as `car::Anova()` unfortunately, but perhaps the documentation is wrong... I did not check):

```{r}
anova(fit_cauchit_gam2)
```

## Predictions

```{r}
data.for.pred2 <- with(data = TitanicSurvival,
                      expand.grid(age_f = levels(TitanicSurvival$age_f),
                                  passengerClass = levels(passengerClass),
                                  sex = levels(sex)))
data.for.pred2$age <- NA
data.for.pred2$age[data.for.pred2$age_f == "[0,10]"] <- (10 - 0)/2
data.for.pred2$age[data.for.pred2$age_f == "(10,25]"] <- (25 + 10)/2
data.for.pred2$age[data.for.pred2$age_f == "(25,35]"] <- (25 + 35)/2
data.for.pred2$age[data.for.pred2$age_f == "(35,50]"] <- (35 + 50)/2
data.for.pred2$age[data.for.pred2$age_f == "(50,Inf]"] <- (50 + 81)/2

pred_age_f <- cbind(data.for.pred2,
                    fit = predict(fit_cauchit, newdata = data.for.pred2,
                                  type = "response"))
pred_age_gam <- cbind(data.for.pred2,
                      fit = predict(fit_cauchit_gam2, newdata = data.for.pred2,
                                    type = "response"))
```

```{r}
ggplot(pred_age_f) +
  aes(y = fit, x = age, colour = passengerClass, linetype = sex, shape = sex) +
  geom_point() +
  geom_line() +
  scale_y_continuous(limits = c(0, 1)) +
  theme_classic() +
  labs(title = "Age as factor", y = "Predicted survival prob.") -> plot1

ggplot(pred_age_gam) +
  aes(y = fit, x = age, colour = passengerClass, linetype = sex, shape = sex) +
  geom_point() +
  geom_line() +
  scale_y_continuous(limits = c(0, 1)) +
  theme_classic() +
  labs(title = "Age as GAM", y = "Predicted survival prob.") -> plot2

library(patchwork) ## to combine plots

plot1 + plot2 & theme(legend.position = "bottom")
```


Note: as the model is not fitted with `link = logit`, the odd ratios are not constant with age, so we cannot quantify effects using them.

<br>


# Dataset: Challenger (difficulty = ?/5)

`r .emo("goal")` What was the probability of O-ring failure when the space shuttle Challenger took off on the 28th of January 1986 by 31 degrees F?


## Let's explore the data

```{r}
str(Challenger)
summary(Challenger)
table(nb_orings = Challenger$oring_tot, nb_pb_orings = Challenger$oring_dt)
```

## Let's visualize the data

```{r}
plot(jitter(oring_dt, factor = 0.3) ~ temp, data = Challenger)
```
Notice the use of jitter!

## We recode the data

```{r}
Challenger$oring_fine <- Challenger$oring_tot - Challenger$oring_dt
```

## We fit the models

```{r}
fit_binom_logit <- glm(cbind(oring_fine, oring_dt) ~ temp, data = Challenger,
                       family = binomial(link = "logit"))
fit_binom_probit <- glm(cbind(oring_fine, oring_dt) ~ temp, data = Challenger,
                        family = binomial(link = "probit"))
fit_binom_cauchit <- glm(cbind(oring_fine, oring_dt) ~ temp, data = Challenger,
                         family = binomial(link = "cauchit"))
AIC(fit_binom_logit)
AIC(fit_binom_probit)
AIC(fit_binom_cauchit)
```

We could choose the logit or the probit model. It would be easier with the logit to interprete the estimates but we do not need this here, so let us consider the probit model since it is best.

## Let's check assumptions
```{r}
r_probit <- simulateResiduals(fit_binom_probit, n = 1000)
plot(r_probit)
r_probit2 <- simulateResiduals(fit_binom_probit, n = 1000, seed = 2)
plot(r_probit2) ## replicate with other seed
testUniformity(r_probit)
testTemporalAutocorrelation(r_probit, time = 1:nrow(Challenger))
testTemporalAutocorrelation(r_probit, time = order(fitted(fit_binom_probit)))
testTemporalAutocorrelation(r_probit, time = order(Challenger$temp))
```
There are some problems about the autocorrelation.
One thing that may help is to consider non linear effects for the temperature, but we don't have much data and we want to extrapolate and polynomials are terrible for that, so we will stick to what we have.

## Tests

```{r}
Anova(fit_binom_probit)
summary(fit_binom_probit)
```

## Predictions

```{r}
pred <- data.frame(temp = 31:81)
pred_probit <- predict(fit_binom_probit, newdata = pred, se.fit = TRUE)
pred$pred <- binomial(link = "probit")$linkinv(pred_probit$fit)
pred$lwr <- binomial(link = "probit")$linkinv(
  pred_probit$fit + qnorm(0.025) * pred_probit$se.fit)
pred$upr <- binomial(link = "probit")$linkinv(
  pred_probit$fit + qnorm(0.975) * pred_probit$se.fit)

plot(1 - pred$pred ~ pred$temp, type = "l",
     ylim = c(0, 1), ylab = "Probability of failure",
     xlab = "Temp (F)")
points(1 - pred$lwr ~ pred$temp, type = "l", lty = 2)
points(1 - pred$upr ~ pred$temp, type = "l", lty = 2)
points(oring_dt / oring_tot ~ temp, data = Challenger, col = "red")

## Predict logit
pred2 <- data.frame(temp = 31:81)
pred_logit <- predict(fit_binom_logit, newdata = pred, se.fit = TRUE)
pred2$pred <- binomial(link = "logit")$linkinv(pred_logit$fit)
pred2$lwr <- binomial(link = "logit")$linkinv(
  pred_logit$fit + qnorm(0.025) * pred_logit$se.fit)
pred2$upr <- binomial(link = "logit")$linkinv(
  pred_logit$fit + qnorm(0.975) * pred_logit$se.fit)

points(1 - pred2$pred ~ pred2$temp, type = "l", col = 3)
points(1 - pred2$lwr ~ pred2$temp, type = "l", lty = 2, col = 3)
points(1 - pred2$upr ~ pred2$temp, type = "l", lty = 2, col = 3)
```

## More

Since a single failure could be fatal, perhaps a binary model would make more sense!

You could try to replicate the original study that excluded all the OK flights!

<br>


# Dataset: UK (difficulty = ?/5)

`r .emo("goal")` Try to identify the influential determinants of the smoking behaviour of children.

`r .emo("goal")` Try to identify the influential determinants of bronchitis in children.

`r .emo("goal")` Try to identify the influential determinants of the variable ```backward```.

## Smoking behaviour

### Looking at the data

```{r}
str(UK)
lapply(UK, summary)
UK$cigarettes_kid_bin <- UK$cigarettes_kid != "Never" 
table(UK$cigarettes_kid_bin, UK$cigarettes_friends)
scatterplot(cigarettes_kid_bin ~ age + sex, data = UK)
scatterplot(cigarettes_kid_bin ~ cigarettes + sex, data = UK)
scatter.smooth(UK$age, UK$cigarettes, lpars = list(col = "red", lwd = 2))
```

### Fiting the model

```{r}
fit_logit <- glm(cigarettes_kid_bin ~ sex*(cigarettes + age + cigarettes_friends),
                 family = binomial(link = "logit"), data = UK)
fit_probit <- glm(cigarettes_kid_bin ~ sex*(cigarettes + age + cigarettes_friends),
                  family = binomial(link = "probit"), data = UK)
fit_cauchit <- glm(cigarettes_kid_bin ~ sex*(cigarettes + age + cigarettes_friends),
                   family = binomial(link = "cauchit"), data = UK)
c(AIC(fit_logit), AIC(fit_probit), AIC(fit_cauchit))
```

### Comparison to null model

```{r}
anova(update(fit_logit, . ~ 1, data = fit_logit$model), fit_logit, test = "LR")
```

### Test of the model assumptions

```{r}
resid_fit_logit <- simulateResiduals(fit_logit, n = 1000)
plot(resid_fit_logit)
testUniformity(resid_fit_logit)
testTemporalAutocorrelation(resid_fit_logit, time = order(fitted(fit_logit)))
testTemporalAutocorrelation(resid_fit_logit, time = 1:nrow(fit_logit$model))
```

### Tests

```{r}
Anova(fit_logit)
```

### Predictions

Using the nice feature "cond" from visreg:

```{r}
visreg(fit_logit, "cigarettes", by = "sex", scale = "response",
       cond = list(cigarettes_friends = "None of them", age = mean(fit_logit$model$age)))
visreg(fit_logit, "cigarettes", by = "sex", scale = "response",
       cond = list(cigarettes_friends = "Most of them", age = mean(fit_logit$model$age)))
visreg(fit_logit, "age", by = "sex", scale = "response",
       cond = list(cigarettes = 0, cigarettes_friends = "None of them"))
visreg(fit_logit, "cigarettes_friends", by = "sex", scale = "response",
       cond = list(cigarettes = 0, age = mean(fit_logit$model$age)))
```

<br>

# Dataset: HSE98women, exercise menopause (difficulty = ?/5)

`r .emo("goal")` Study the influences of age, body mass index and smoking status upon the probability of being menopaused.

`r .emo("goal")` Estimate the proportion of menopaused women in the population, depending on their age (40, 45, 50, 55, 60 yrs).

Note: here, for the sake of simplicity we simply define menopause as the lack of periods in old ages.


## Data preparation and exploration

We create the response variable:
```{r}
HSE98women$menopause <-  HSE98women$period == FALSE
```

We fit a first model to help us easily exploring the data used for fitting and not all data (there are many missing values in the dataset).

```{r}
fit_menop_logit <- glm(menopause ~ age + bmi + smoked, data = HSE98women, family = binomial())
nrow(HSE98women) ## original number of observations
nrow(fit_menop_logit$model)  ## number of observations considered in the model
```

We compare the variables before and after filtering to see if any obvious biais is introduced by the observations we discared:
```{r}
summary(HSE98women[, c("menopause", "age", "bmi", "smoked")])
summary(fit_menop_logit$model)
```

The only big discrepancy between the two datasets is that the minimal age is now 18 yrs old while the original dataset also contains children. That suggests that girls too young to have periods are not being considered, which will greatly help us to model menopause.

Let's check the correlation between our two continuous predictor:

```{r}
cor.test(HSE98women$age, HSE98women$bmi)
cor.test(fit_menop_logit$model$age, fit_menop_logit$model$bmi)
```

The correlation is not very high, especially in the dataset that is used for the fit, so multicollinearity should not be an issue.

We visualise the relationship between the predictor and the response:
```{r}
scatterplot(menopause ~ age + smoked, data = fit_menop_logit$model)
scatterplot(menopause ~ bmi + smoked, data = fit_menop_logit$model)
```
There don't seem to be any obvious interaction between the age and smoked or between BMI and smoked, so we won't consider any. We will however consider that the effect of age may depend on the BMI.


## Fitting the models

```{r}
fit_menop_logit   <- glm(menopause ~ age * bmi + smoked, data = HSE98women, family = binomial(link = "logit"))
fit_menop_probit  <- glm(menopause ~ age * bmi + smoked, data = HSE98women, family = binomial(link = "probit"))
fit_menop_cauchit <- glm(menopause ~ age * bmi + smoked, data = HSE98women, family = binomial(link = "cauchit"))
```

Because it is very difficult to explore the shape of the relationship between the age and bmi upon the probability of being menopaused, we will also fit general additive models (GAM), which allows parameters to vary along the predictors.

```{r, message=FALSE}
fit_menop_logit_gam   <- gam(menopause ~ s(age, bmi) + smoked, data = HSE98women, family = binomial(link = "logit"))
fit_menop_probit_gam  <- gam(menopause ~ s(age, bmi) + smoked, data = HSE98women, family = binomial(link = "probit"))
fit_menop_cauchit_gam <- gam(menopause ~ s(age, bmi) + smoked, data = HSE98women, family = binomial(link = "cauchit"))
```

```{r}
rbind(logit = AIC(fit_menop_logit),
      probit = AIC(fit_menop_probit),
      cauchit = AIC(fit_menop_cauchit),
      logit_GAM = AIC(fit_menop_logit_gam),
      probit_GAM = AIC(fit_menop_probit_gam),
      cauchit_GAM = AIC(fit_menop_cauchit_gam)
      )
```

The GAM logit model seems to be the best model, which will prevent us to express simply the effect of coefficients compared to a usual GLM. That's too bad but as the model is much better, we will go for it.

## Tests

The best practices for testing effects in GAM are not quite established.

We are going to test for the effect of the interaction by Likelihood ratio test:
```{r}
fit_menop_logit_gam_no_int <- gam(menopause ~ s(age) + s(bmi) + smoked, family = binomial(), data = fit_menop_logit_gam$model)
anova(fit_menop_logit_gam_no_int, fit_menop_logit_gam, test = "Chisq")
```
Note that models with GAM have non-integer degrees of freedom. This is normal.

Let's now test the effect of BMI by Likelihood ratio test:

```{r}
fit_menop_logit_gam_no_bmi <- gam(menopause ~ s(age) + smoked, family = binomial(), data = fit_menop_logit_gam$model)
anova(fit_menop_logit_gam_no_bmi, fit_menop_logit_gam, test = "Chisq")
```

Let's now test the effect of age by Likelihood ratio test:

```{r}
fit_menop_logit_gam_no_age <- gam(menopause ~ s(bmi) + smoked, family = binomial(), data = fit_menop_logit_gam$model)
anova(fit_menop_logit_gam_no_age, fit_menop_logit_gam, test = "Chisq")
```

Let's now test the effect of smoking by Likelihood ratio test:

```{r}
fit_menop_logit_gam_never_smoked <- gam(menopause ~ s(age, bmi), family = binomial(), data = fit_menop_logit_gam$model)
anova(fit_menop_logit_gam_never_smoked, fit_menop_logit_gam, test = "Chisq")
```


## Predictions

Let's first explore the model output with the pacakge **{visreg}**, which also works on GAM models:

```{r}
visreg(fit_menop_logit_gam)
```

These plots show the patterns identified by the smoothing parameters in the GAM: the effect of age seems to change after 40 yrs and the effect of BMI seem to be changing between 23 and 32 yrs. It is difficult to account for such threshold effects by polynoms, so we will have to stick to GAM here...

Le't us do the same plots on the scale of the response variable:
```{r}
visreg(fit_menop_logit_gam, scale = "response")
visreg2d(fit_menop_logit_gam, "age", "bmi", scale = "response")
#visreg2d(fit_menop_logit_gam, "age", "bmi", scale = "response", plot.type = "rgl")
```

We will now plot the predictions for the model with the logit link, the model with the cauchit link and the model with the GAM model with the logit link to explore their differences. We will predict the influence of age and consider the median BMI and smokers.

```{r}
age.for.pred <- seq(min(fit_menop_cauchit_gam$model$age), max(fit_menop_cauchit_gam$model$age), length = 100)
data.for.pred <- expand.grid(
  age = age.for.pred,
  bmi = median(fit_menop_cauchit_gam$model$bmi),
  smoked = TRUE
  )

gam.p <- predict(fit_menop_logit_gam, newdata = data.for.pred, se.fit = TRUE)
gam.upr <- binomial(link = "logit")$linkinv(gam.p$fit + qnorm(0.975) * gam.p$se.fit)
gam.lwr <- binomial(link = "logit")$linkinv(gam.p$fit + qnorm(0.025) * gam.p$se.fit)
gam.fit <- binomial(link = "logit")$linkinv(gam.p$fit)

cauchit.p <- predict(fit_menop_cauchit, newdata = data.for.pred, se.fit = TRUE)
cauchit.upr <- binomial(link = "cauchit")$linkinv(cauchit.p$fit + qnorm(0.975) * cauchit.p$se.fit)
cauchit.lwr <- binomial(link = "cauchit")$linkinv(cauchit.p$fit + qnorm(0.025) * cauchit.p$se.fit)
cauchit.fit <- binomial(link = "cauchit")$linkinv(cauchit.p$fit)

logit.p <- predict(fit_menop_logit, newdata = data.for.pred, se.fit = TRUE)
logit.upr <- binomial(link = "logit")$linkinv(logit.p$fit + qnorm(0.975) * logit.p$se.fit)
logit.lwr <- binomial(link = "logit")$linkinv(logit.p$fit + qnorm(0.025) * logit.p$se.fit)
logit.fit <- binomial(link = "logit")$linkinv(logit.p$fit)

plot(gam.fit ~ age.for.pred, type = "l", las = 1, ylab = "Probability of being menopaused", xlab = "Age (yrs)", ylim = c(0, 1))
points(gam.upr ~ age.for.pred, type = "l", lty = 2)
points(gam.lwr ~ age.for.pred, type = "l", lty = 2)

points(cauchit.fit ~ age.for.pred, type = "l", lty = 1, col = "green")
points(cauchit.upr ~ age.for.pred, type = "l", lty = 2, col = "green")
points(cauchit.lwr ~ age.for.pred, type = "l", lty = 2, col = "green")

points(logit.fit ~ age.for.pred, type = "l", lty = 1, col = "red")
points(logit.upr ~ age.for.pred, type = "l", lty = 2, col = "red")
points(logit.lwr ~ age.for.pred, type = "l", lty = 2, col = "red")

legend("topleft", fill = c("black", "green", "red"), legend = c("logit GAM", "cauchit", "logit"), bty = "n")
```

Selecting the best model, let's now look at the effect of BMI, still using smokers:

```{r}
data.for.pred <- expand.grid(
  age = age.for.pred,
  bmi = c(18.5, 25, 30),
  smoked = TRUE
  )

gam.skiny.p <- predict(fit_menop_logit_gam, newdata = subset(data.for.pred, bmi == 18.5), se.fit = TRUE)
gam.skiny.upr <- binomial(link = "logit")$linkinv(gam.skiny.p$fit + qnorm(0.975) * gam.skiny.p$se.fit)
gam.skiny.lwr <- binomial(link = "logit")$linkinv(gam.skiny.p$fit + qnorm(0.025) * gam.skiny.p$se.fit)
gam.skiny.fit <- binomial(link = "logit")$linkinv(gam.skiny.p$fit)

gam.overweight.p <- predict(fit_menop_logit_gam, newdata = subset(data.for.pred, bmi == 25), se.fit = TRUE)
gam.overweight.upr <- binomial(link = "logit")$linkinv(gam.overweight.p$fit + qnorm(0.975) * gam.overweight.p$se.fit)
gam.overweight.lwr <- binomial(link = "logit")$linkinv(gam.overweight.p$fit + qnorm(0.025) * gam.overweight.p$se.fit)
gam.overweight.fit <- binomial(link = "logit")$linkinv(gam.overweight.p$fit)

gam.obese.p <- predict(fit_menop_logit_gam, newdata = subset(data.for.pred, bmi == 30), se.fit = TRUE)
gam.obese.upr <- binomial(link = "logit")$linkinv(gam.obese.p$fit + qnorm(0.975) * gam.obese.p$se.fit)
gam.obese.lwr <- binomial(link = "logit")$linkinv(gam.obese.p$fit + qnorm(0.025) * gam.obese.p$se.fit)
gam.obese.fit <- binomial(link = "logit")$linkinv(gam.obese.p$fit)

plot(gam.skiny.fit ~ age.for.pred, type = "l", las = 1, ylab = "Probability of being menopaused", xlab = "Age (yrs)", ylim = c(0, 1))
points(gam.skiny.upr ~ age.for.pred, type = "l", lty = 2)
points(gam.skiny.lwr ~ age.for.pred, type = "l", lty = 2)

points(gam.overweight.fit ~ age.for.pred, type = "l", lty = 1, col = "green")
points(gam.overweight.upr ~ age.for.pred, type = "l", lty = 2, col = "green")
points(gam.overweight.lwr ~ age.for.pred, type = "l", lty = 2, col = "green")

points(gam.obese.fit ~ age.for.pred, type = "l", lty = 1, col = "red")
points(gam.obese.upr ~ age.for.pred, type = "l", lty = 2, col = "red")
points(gam.obese.lwr ~ age.for.pred, type = "l", lty = 2, col = "red")

legend("topleft", fill = c("black", "green", "red"), legend = c("skiny", "overweight", "obese"), bty = "n")
```

We can see on this plot that obesity seems to accelerate the onset of menopause, but it does not seem to make any difference later on.

We will now predict the proportion of women in menopause at 40, 45, 50, 55 and 60 yrs. To illustrate the effect ofthe BMI will will perform prediction for a BMI of 21.75 (middle of normal BMI range according to WHO) and 30 (lower limit for obesity) Kg/(m^2). Again, we will perform the predictions considering smoker since we have data for this category. We will also compute the confidence intervals (prediction intervals make little sense in the context of the prediction of proportion...).

```{r}
data.for.pred <- expand.grid(
  age = c(40, 45, 50, 55, 60),
  bmi = c(21.75, 30),
  smoked = TRUE
  )
pred_normal <- predict(fit_menop_logit_gam, newdata = subset(data.for.pred, bmi == 21.75), se.fit = TRUE)
tab_normal <- cbind(
  predict = plogis(pred_normal$fit),
  lwr.CI = plogis(pred_normal$fit + qnorm(0.025)*pred_normal$se.fit),
  upr.CI = plogis(pred_normal$fit + qnorm(0.975)*pred_normal$se.fit)
)
rownames(tab_normal) <- c(40, 45, 50, 55, 60)
round(tab_normal, 2)
```

This gives the information for individuals for normal BMI. Let's now look at the corresponding proportion for the obeses:

```{r}
pred_obses <- predict(fit_menop_logit_gam, newdata = subset(data.for.pred, bmi == 30), se.fit = TRUE)
tab_obses <- cbind(
  predict = plogis(pred_obses$fit),
  lwr.CI = plogis(pred_obses$fit + qnorm(0.025)*pred_obses$se.fit),
  upr.CI = plogis(pred_obses$fit + qnorm(0.975)*pred_obses$se.fit)
)
rownames(tab_obses) <- c(40, 45, 50, 55, 60)
round(tab_obses, 2)
```

These tables confirm our expectation concerning the influenc of BMI upon the onset of menopause.

For the sake of comparion, let's compare the first table to what we would have predicted using the cauchit model or the logit GLM model:

```{r}
data.for.pred <- expand.grid(age = c(40, 45, 50, 55, 60), bmi = 21.75, smoked = TRUE)
pred_GAM <- predict(fit_menop_logit_gam, newdata = subset(data.for.pred, bmi == 21.75), se.fit = TRUE)
pred_logit <- predict(fit_menop_logit, newdata = subset(data.for.pred, bmi == 21.75), se.fit = TRUE)
pred_cauchit <- predict(fit_menop_cauchit, newdata = subset(data.for.pred, bmi == 21.75), se.fit = TRUE)
tab <- cbind(
  logit_GAM = plogis(pred_GAM$fit),
  logit_GLM = plogis(pred_logit$fit),
  cauchit_GLM = binomial(link = "cauchit")$linkinv(pred_cauchit$fit)
)
rownames(tab) <- c(40, 45, 50, 55, 60)
round(tab, 2)
```

Let's look at the same proportion in the raw data. We cannot be as strict otherwise we won't get much individuals, so let's enlarge a bit the selection and see that we get a few hundred individuals at least:
```{r}
tab <- with(HSE98women,
     rbind(sum(age >= 39 & age <= 41 & bmi > 18.5 & bmi < 25, na.rm = TRUE),
           sum(age >= 44 & age <= 46 & bmi > 18.5 & bmi < 25, na.rm = TRUE),
           sum(age >= 49 & age <= 51 & bmi > 18.5 & bmi < 25, na.rm = TRUE),
           sum(age >= 54 & age <= 56 & bmi > 18.5 & bmi < 25, na.rm = TRUE),
           sum(age >= 59 & age <= 61 & bmi > 18.5 & bmi < 25, na.rm = TRUE)
           )
)
rownames(tab) <- c(40, 45, 50, 55, 60)
tab
```
That seems a good selection process (+/- 1 year and the whole WHO normal BMI category), so let's check the proportion of menopaused individuals in each of these categories (keeping both smokers and non smokers as the effect is rather small, see later):

```{r}
tab <- with(HSE98women, 
     rbind(
       mean(menopause[age >= 39 & age <= 41 & bmi > 18.5 & bmi < 25], na.rm = TRUE),
       mean(menopause[age >= 44 & age <= 46 & bmi > 18.5 & bmi < 25], na.rm = TRUE),
       mean(menopause[age >= 49 & age <= 51 & bmi > 18.5 & bmi < 25], na.rm = TRUE),
       mean(menopause[age >= 54 & age <= 56 & bmi > 18.5 & bmi < 25], na.rm = TRUE),
       mean(menopause[age >= 59 & age <= 61 & bmi > 18.5 & bmi < 25], na.rm = TRUE)
     )
)
rownames(tab) <- c(40, 45, 50, 55, 60)
round(tab, 2)
```

These results are roughly consistent with the predictions from the GAM model. It confirms that the GLM logit and cauchit are underestimating the prevalence of menopause at old age. All models however may overestimate the prevalence of menaupose at young age. Depending on the goal of the study, other link functions, not implemented in R base, could be tried but it is quite advanced so we won't do that.

We can now turn to the exploration of the smoking effect by comparing predictions between smokers and non smokers at median BMI:

Selecting the best model, let's now look at the effect of BMI, still using smokers:

```{r}
data.for.pred.nosmoke <- expand.grid(
  age = age.for.pred,
  bmi = median(fit_menop_cauchit_gam$model$bmi),
  smoked = FALSE
  )

gam.p.nosmoke <- predict(fit_menop_logit_gam, newdata = data.for.pred.nosmoke, se.fit = TRUE)
gam.upr.nosmoke <- binomial(link = "logit")$linkinv(gam.p.nosmoke$fit + qnorm(0.975) * gam.p.nosmoke$se.fit)
gam.lwr.nosmoke <- binomial(link = "logit")$linkinv(gam.p.nosmoke$fit + qnorm(0.025) * gam.p.nosmoke$se.fit)
gam.fit.nosmoke <- binomial(link = "logit")$linkinv(gam.p.nosmoke$fit)


plot(gam.fit ~ age.for.pred, type = "l", las = 1, ylab = "Probability of being menopaused", xlab = "Age (yrs)", ylim = c(0, 1), col = "red")
points(gam.upr ~ age.for.pred, type = "l", lty = 2, col = "red")
points(gam.lwr ~ age.for.pred, type = "l", lty = 2, col = "red")

points(gam.fit.nosmoke ~ age.for.pred, type = "l", col = "green")
points(gam.upr.nosmoke ~ age.for.pred, type = "l", lty = 2, col = "green")
points(gam.lwr.nosmoke ~ age.for.pred, type = "l", lty = 2, col = "green")

legend("topleft", fill = c("green", "red"), legend = c("non-smoker", "smoker"), bty = "n")
```

Although significant, the effect of smoking is rather small. Because the interaction between the smoking status and the BMI was not significant, we won't draw predictions to explore that.

