---
title: "GLM: Introduction"
author: "Alexandre Courtiol"
date: "`r Sys.Date()`"
output:
  ioslides_presentation:
    widescreen: true
    smaller: true
vignette: >
  %\VignetteIndexEntry{3.0 Generalized Linear Models}
  %\VignetteEncoding{UTF-8}
  %\VignetteEngine{knitr::rmarkdown}
editor_options: 
  chunk_output_type: console
---
```{r setup, include=FALSE}
library(LM2GLMM)
library(doSNOW) ## for load balancing in spaMM
options(width = 110)
knitr::opts_chunk$set(cache = FALSE, cache.path = "./cache_knitr/GLM_intro/", fig.path = "./fig_knitr/GLM_intro/", fig.width = 5, fig.height = 5, fig.align = "center", error = TRUE)
spaMM::spaMM.options(nb_cores = parallel::detectCores() - 1L)
```

## The Generalized Linear Model: GLM

* 3.0 [Introduction](./GLM_intro_course.html)
* 3.1 [Let's practice](./GLM_practice_exercises.html)

<br>

<div align="right">
[Back to main menu](./Title.html#2)
</div>

## You will learn in this session `r .emo("goal")`

* that GLM can be fitted on different types of responses
* how to convert values between the response and linear predictor scales
* how to interpret parameter estimates from GLM Poisson and Binomial
* that LR tests are best for family with known dispersion
* that intervals based on likelihood are best for family with known dispersion
* that tests by parametric bootstrap are best and easy to do with **{spaMM}**
* that most assumptions behind GLM are similar to those for LM (but harder to test)
* that General Additive Models (GAM) can be useful to improve linearity
* how to diagnose and tackle overdispersion, zero-augmentation and separation


# Definition and notations

## Mathematical notations of GLM `r .emo("info")`

<font size="5"> $\eta_i = \beta_0 + \beta_1 \times x_{1,i} + \beta_2 \times x_{2,i} + \dots + \beta_{p} \times x_{p,i}$ </font> or in matrix notation <font size="5"> $\eta = \text{X}\beta$ </font>,

with:

* $\text{E}(\text{Y}) = \mu = g^{-1}(\eta)$
* $\text{Var}(\text{Y}) = \phi\text{V}(\mu)$ 

<br>

We call:

* $\eta$ the linear predictor
* $g$ the link function ($g^{-1}$ is the inverse link function)
* $\text{V}$ the variance function
* $\phi$ is the dispersion parameter

The GLM fit leads to <font size="5"> $\text{Y} = g^{-1}(\eta) + \epsilon = g^{-1}(\text{X}\beta) + \epsilon = g^{-1}(\widehat{\eta})+ \varepsilon = g^{-1}(\text{X}\widehat{\beta}) + \varepsilon$ </font>

Here the errors can have a distribution other than Gaussian.

## LM is a particular case of GLM `r .emo("info")`

<font size="5"> $\eta_i = \beta_0 + \beta_1 \times x_{1,i} + \beta_2 \times x_{2,i} + \dots + \beta_{p} \times x_{p,i}$ </font> or in matrix notation <font size="5"> $\eta = \text{X}\beta$ </font>,

with:

* $\text{E}(\text{Y}) = \mu = g^{-1}(\eta)$
* $\text{Var}(\text{Y}) = \phi\text{V}(\mu)$ 

<br>

This is identical to the LM if:

* $\mu = g^{-1}(\eta) = \eta$, where $g$ is the *identity* function
* $\phi = \sigma^2$, the dispersion parameter equals the error variance
* $\text{V}(\mu) = 1$, the variance function is constant

<br>

So in GLM, the probability distribution of $\text{Y}$ can be Gaussian.

## GLM can do more than LM! `r .emo("info")`

### The response

* one dimension: $n \times 1$
* continuous or categorical
* member of the univariate linear exponential family: $f(y; \theta; \phi) = \text{exp}\left( \frac{y\theta-b(\theta)}{a(\phi)} + c(y; \phi)\right)$ `r .emo("nerd")`

This includes:

* the normal distribution (for many continuous outcomes)
* the binomial distribution (for binary or binomial outcomes)
* the Poisson distribution (for counts)
* the gamma distribution (for continuous positive outcomes such as survival times)
* other less mainsteam distributions


## What do we need to fit a GLM? `r .emo("info")`

* data = response variable + design matrix, as for LM
* the probability distribution (e.g. Gaussian, Poisson...) which sets the variance function
* the link function

<br>

The probability distribution and the link function are obtained in R by calling the `family` functions:

```{r family, eval = FALSE}
gaussian(link = "identity")
binomial(link = "logit")
poisson(link = "log")
```

<br>

Note: the objects of class `family` created by these functions also contain other useful information needed for the fitting procedure.


# Data and model fits for this session

## Simulating data for GLM `r .emo("alien")`

```{r}
set.seed(1L)
Aliens <- data.frame(humans_eaten = round(runif(n = 20, min = 0, max = 15)))
Aliens$size  <- rnorm( n  = 20, mean = 50 + 1.5 * Aliens$humans_eaten, sd = 5)
Aliens$eggs  <- rpois( n  = 20, lambda = exp(-1 + 0.1 * Aliens$humans_eaten))
Aliens$happy <- rbinom(n  = 20, size = 1, prob = plogis(-3 + 0.3 * Aliens$humans_eaten))
Aliens$all_eyes  <- round(runif(nrow(Aliens), min = 1, max = 12))
Aliens$blue_eyes <- rbinom(n = nrow(Aliens), size = Aliens$all_eyes,
                           prob = plogis(-2 + 0.5 * Aliens$humans_eaten))
Aliens$pink_eyes <- Aliens$all_eyes - Aliens$blue_eyes
head(Aliens)
```


## Simulating data for GLM `r .emo("alien")`

Or, using the **{LM2GLMM}** function:
```{r data}
set.seed(1L)
Aliens <- simulate_Aliens_GLM(N = 20)
head(Aliens)
```

The true parameter values are stored as attributes if you need them `r .emo("nerd")`
```{r attributes}
attributes(Aliens)$param.eta$blue_eyes  ## to get back the parameters
```


## Fitting the models `r .emo("practice")`

```{r pv, message=FALSE}
fit_gauss_lm <- lm(size  ~ humans_eaten, data = Aliens)

fit_gauss <- glm(size  ~ humans_eaten, family = gaussian(), data = Aliens)
fit_poiss <- glm(eggs  ~ humans_eaten, family = poisson(),  data = Aliens)
fit_binar <- glm(happy ~ humans_eaten, family = binomial(), data = Aliens)
fit_binom <- glm(cbind(blue_eyes, pink_eyes) ~ humans_eaten, family = binomial(),  data = Aliens)

library(spaMM)
fit_gauss_spaMM_ML <- fitme(size  ~ humans_eaten, family = gaussian(), data = Aliens)
fit_poiss_spaMM_ML <- fitme(eggs  ~ humans_eaten, family = poisson(),  data = Aliens)
fit_binar_spaMM_ML <- fitme(happy ~ humans_eaten, family = binomial(), data = Aliens)
fit_binom_spaMM_ML <- fitme(cbind(blue_eyes, pink_eyes) ~ humans_eaten, family = binomial(),  data = Aliens)

fit_gauss_spaMM_REML <- fitme(size  ~ humans_eaten, family = gaussian(), data = Aliens, method = "REML")
fit_poiss_spaMM_REML <- fitme(eggs  ~ humans_eaten, family = poisson(),  data = Aliens, method = "REML")
fit_binar_spaMM_REML <- fitme(happy ~ humans_eaten, family = binomial(), data = Aliens, method = "REML")
fit_binom_spaMM_REML <- fitme(cbind(blue_eyes, pink_eyes) ~ humans_eaten, family = binomial(),  data = Aliens,
                              method = "REML")
```


## The fitted GLM from `stats::glm()` `r .emo("practice")`

```{r fitted glm stats}
fit_poiss
```


## The fitted GLM from `spaMM::fitme()` `r .emo("practice")`

```{r fitted glm spaMM}
fit_poiss_spaMM_ML
```

<!--
# The GLM object

## Components stored in GLM objects

```{r}
names(fit_gauss)
names(fit_binom)
```

The components are the same no matter the family.

We will only detail the interesting ones.

## Coefficients (```coefficients```)

```{r}
fit_gauss$coefficients
fit_poiss$coefficients
fit_binom$coefficients
```

Same idea as for LM, but here they are expressed in the scale of the linear predictor!!!

## The residuals (```residuals```)

As we shall see later, those are the working residuals:

```{r}
rbind(fit_poiss$residuals[1:2], residuals(fit_poiss, type = "working")[1:2])
rbind(fit_binom$residuals[1:2], residuals(fit_binom, type = "working")[1:2])
```

<br>

Although useful during the fitting procedure, these residuals are not useful to us.

We will see that we can extract other types of residuals that are more useful with ```residuals()```.


## Linear predictors (```linear.predictors```)

This contains the predicted values for the observations at the scale of the linear predictor.

```{r}
rbind(summary(fit_gauss$linear.predictors),
      summary(fit_poiss$linear.predictors),
      summary(fit_binar$linear.predictors),
      summary(fit_binom$linear.predictors)
      )
```


## Linear predictors (```linear.predictors```)

They are computed as in LM:

```{r}
rbind(fit_gauss$linear.predictors[1:2], (model.matrix(fit_gauss) %*% fit_gauss$coefficients)[1:2])
rbind(fit_poiss$linear.predictors[1:2], (model.matrix(fit_poiss) %*% fit_poiss$coefficients)[1:2])
rbind(fit_binom$linear.predictors[1:2], (model.matrix(fit_binom) %*% fit_binom$coefficients)[1:2])
```


## Fitted values (```fitted.values```)

This contains the predicted values for the observations at the scale of the response:

```{r}
rbind(summary(fit_gauss$fitted.values),
      summary(fit_poiss$fitted.values))
```

<br>

For binomial family, the fitted values is a probability:

```{r}
rbind(summary(fit_binar$fitted.values),
      summary(fit_binom$fitted.values))
```


## Fitted values (```fitted.values```)

The fitted values are obtained from the linear predictors:

```{r}
rbind(fit_gauss$fitted.values[1:2], gaussian()$linkinv(fit_gauss$linear.predictors[1:2]))
rbind(fit_poiss$fitted.values[1:2], poisson()$linkinv(fit_poiss$linear.predictors[1:2]))
rbind(fit_binom$fitted.values[1:2], binomial()$linkinv(fit_binom$linear.predictors[1:2]))
```

## Deviances (```deviance``` & ```null.deviance```)

The GLM object contains:

* the unscaled residual deviance of the model fit (deviance between the fit and the fit of a saturated model)
* the unscaled null deviance (deviance between the null fit and the fit of a saturated model)

```{r}
sigma2_error <- deviance(fit_gauss) / fit_gauss$df.residual
sigma2_resid <- sigma2_error *  fit_gauss$df.residual / nrow(fit_gauss$model)
fit_gauss_sat <- spaMM::fitme(formula = size ~ -1 + offset(size), data = Aliens, fixed = list(phi = sigma2_resid))

c(fit_gauss$deviance, deviance(fit_gauss), -2*(logLik(fit_gauss) - logLik(fit_gauss_sat))*sigma2_resid)
fit_gauss_null <- update(fit_gauss, . ~ 1)
c(fit_gauss$null.deviance, fit_gauss_null$deviance)
```


## Deviances (```deviance``` & ```null.deviance```)

The GLM object contains:

* the unscaled residual deviance of the model fit (deviance between the fit and the fit of a saturated model)
* the unscaled null deviance (deviance between the null fit and the fit of a saturated model)

```{r}
fit_poiss_sat <- update(fit_poiss, . ~ as.factor(fit_poiss$y))
c(fit_poiss$deviance, -2*(logLik(fit_poiss) - logLik(fit_poiss_sat)))
fit_poiss_null <- update(fit_poiss, . ~ 1)
c(fit_poiss$null.deviance, fit_poiss_null$deviance)
```

## Deviances (```deviance``` & ```null.deviance```)

The GLM object contains:

* the unscaled residual deviance of the model fit (deviance between the fit and the fit of a saturated model)
* the unscaled null deviance (deviance between the null fit and the fit of a saturated model)

```{r}
fit_binom_sat <- update(fit_binom, . ~ as.factor(fit_binom$y))
c(fit_binom$deviance, -2*(logLik(fit_binom) - logLik(fit_binom_sat)))
fit_binom_null <- update(fit_binom, . ~ 1)
c(fit_binom$null.deviance, fit_binom_null$deviance)
```

## Akaike Information Criterion (```aic```)

Contrary to ```family()$aic``` here we directly have the correct AIC and not just part of it:

<br>

```{r}
c(fit_gauss$aic, -2*logLik(fit_gauss) + 2*(fit_gauss$rank + 1))  ## + 1 for the estimate of the variance
c(fit_poiss$aic, -2*logLik(fit_poiss) + 2*(fit_poiss$rank))
c(fit_binom$aic, -2*logLik(fit_binom) + 2*(fit_binom$rank))
```


## Prior weights (```prior.weights```)

The prior weights are weights for the variance function often imputed by the user:

```{r}
logLik(fit_poiss)
fit_poiss_double <- glm(eggs  ~ humans_eaten, family = poisson(), weights = rep(2, nrow(Aliens)), data = Aliens)

Aliens_double <- Aliens[rep(1:nrow(Aliens), each = 2), ]
fit_poiss_double_bis <- glm(eggs  ~ humans_eaten, family = poisson(),  data = Aliens_double)
c(logLik(fit_poiss_double), logLik(fit_poiss_double_bis))

fit_poiss_double$prior.weights[1:4]
```

Don't get confused: the prior weights must be specify via the argument ```weights``` in ```lm()``` and ```glm()```.


## Prior weights (```prior.weights```)

The prior weights are especially useful for binomial models with more than one trial.

They are computed implicitly when you use the syntax ```cbind(sucess, failure)```:

```{r}
fit_binom$prior.weights[1:6]
```

but they can also be set manually to directly use the probability of success as the response variable:

```{r}
Aliens$prop_blue <- Aliens$blue_eyes / Aliens$all_eyes
fit_binom_bis  <- glm(prop_blue ~ humans_eaten, weights = all_eyes, data = Aliens, family = binomial())
c(logLik(fit_binom), logLik(fit_binom_bis))
```

<br>

If you use ```cbind``` AND prior weights, the prior weights considered in the model will be the product of both priors!
-->

# The family object

## What does the family object contain? `r .emo("info")`

A lot of functions that will be used by `glm()` and friends:

```{r}
str(binomial(link = "logit"))
```

Note: we will explore the 2 most useful components.


## The link function (```link``` & ```linkfun```) `r .emo("practice")`

The link function converts the values from the scale of the response to the one of the linear predictor.

```{r, results='hold'}
binomial(link = "logit")$linkfun(0.9)  ## the default
log(0.9/(1 - 0.9))
```
```{r, results='hold'}
binomial(link = "probit")$linkfun(0.9)
qnorm(0.9)
```

<br>

Note: when things get complex, to convert the values from the scale of the response to the one of the linear predictor, use `linkfun()` from the family object instead of trying to do it by hand.

<!--
c(binomial(link = "cauchit")$linkfun(0.9), qcauchy(0.9))
c(binomial(link = "log")$linkfun(0.9), log(0.9))
c(binomial(link = "cloglog")$linkfun(0.9), log(-log(1 - 0.9)))
-->

## The inverse link function (`linkinv`) `r .emo("practice")`

The inverse link function converts the values of the linear predictor back into the scale of the response.

<br>

```{r, results = "hold"}
binomial(link = "logit")$linkinv(2.197225)
plogis(2.197225)
1/(1 + exp(-2.197225))
```

<br>

Note: when things get complex, to convert the values from the scale of the linear predictor to the one of the response, use `linkinv()` from the family object instead of trying to do it by hand.

<!--
## The variance function (```variance```)

This function gives the value for $\text{V}(\mu)$ for any $\mu$:

<br>

```{r}
gaussian(link = "identity")$variance(1:10)  ## = constant
poisson(link = "log")$variance(1:10) ## = lambda
binomial(link = "logit")$variance(seq(0, 1, 0.1))  ## = p x q
```


## Deviance residuals (```dev.resids```)

This function computes the squared deviance residuals (their sum equals the unscaled deviance).

It is need because the deviance is a key statistics for GLM: it is the LRT statistics between a fit and the fit of a *saturated model*:

```{r}
fit_binom$deviance
-2*(logLik(fit_binom) - logLik(update(fit_binom, . ~ as.factor(fit_binom$y))))
```


```{r}
sum(with(fit_binom, binomial()$dev.resids(y = y, mu = fitted.values, wt = prior.weights)))
```


## Other components from ```family()```

* aic: compute part of the AIC (not the whole thing!)
* initialize: creates the initial values for mu and recomputes Y and weights if binomial with cbind
* mu.eta: compute the derivative $\frac{\text{d}\mu}{\text{d}\eta}$ (for the iterative fitting procedure)
* validmu: test if the values of mu are valid
* valideta: test if the values of eta are valid
* simulate: simulation function (for e.g. parametric bootstrap)
-->

# Fitting procedure

## How to fit a GLM? `r .emo("info")`

### By maximum likelihood

<br>

We want to find the $\widehat{\beta}$ maximizing the likelihood.

This is the same as finding the $\widehat{\beta}$ minimizing the residual deviance.

<br>

Residual deviance = $-2 \times (\text{logLik}_\text{focal fit} - \text{logLik}_\text{staturated fit})$


## A simple function to fit GLM numerically `r .emo("info")`

```{r}
my_glm1 <- function(formula, data, family, weights = NULL) {

  useful.data <- model.frame(formula = formula, data = data)  ## keep only useful rows and columns
  X <- model.matrix(object = formula, data = useful.data)     ## build design matrix
  y <- model.response(useful.data)                            ## extract response
  nobs <- NROW(y) ## count the rows of observations (not nrow() as y can be a vector or a matrix)
  if (is.null(weights)) weights <- rep(1, nobs)  ## set the prior.weights if NULL

  etastart <- start <- mustart <- NULL  ## required for family$initialize
  eval(family$initialize)               ## recomputes y and prior weights if binomial + cbind
  
  get_resid_deviance <- function(coef) {  ## define function computing the unscaled residual deviance
    eta <- as.numeric(X %*% coef)   ## compute the linear predictors
    mu  <- family$linkinv(eta)      ## express the linear predictor in the scale of the response
    sum(family$dev.resids(y, mu, weights))  ## compute the deviance. Here: prior weights are used!
  }
  
  fit <- nlminb(rep(0, ncol(X)), get_resid_deviance) ## optimisation; same as optim() but seems to work better
  if (fit$convergence != 0) warning("the algorithm did not converge")  ## test if convergence has been reached
  fit$par  ## returns estimates
}
```

Note: it should handle all families but it does not consider offsets, convergence issues and maybe other sources of complexity.


## No analytic solution... `r .emo("info")`

### Problem

unlike LM there is no analytic solution to directly obtain maximum likelihood estimates (MLE)!

### Solution to avoid the full numerical optimisation

The function ```glm.fit()``` fits GLM using an iterative algorithm based on weighted least squares to update parameter estimates.

The *working weights* (not to be mixed up with *prior weights*) are inversely proportional to the variance of the response, but the variance depends on the mean which, in turn, depends on the parameters.

The algorithm fixes these weights, determines the parameter values that minimize the weighted sum of squared residuals, then updates the weights and repeats the process until the weights stabilize (it usually happens in very few iterations).

This algorithm is thus called IRLS for iteratively re-weighted least squares, it is also sometimes called Fisher's scoring which has been proposed by Fisher (in a context more general than GLM).


## A toy function using IRLS `r .emo("nerd")`

```{r}
my_glm2 <- function(formula, data, family, weights = NULL) {
  useful.data <- model.frame(formula = formula, data = data)  ## keep only useful rows and columns
  X <- model.matrix(object = formula, data = useful.data)     ## build design matrix
  y <- model.response(useful.data)                            ## extract response
  nobs <- NROW(y) ## count the rows of observations (not nrow() as y can be a vector or a matrix)
  if (is.null(weights)) weights <- rep(1, nobs)  ## set the prior.weights if NULL

  etastart <- start <- mustart <- NULL  ## required for family$initialize
  eval(family$initialize)         ## creates mustart -- the initial values for mu -- and 
                                  ## recomputes y and weights if binomial + cbind
  eta <- family$linkfun(mustart)  ## compute initial values for eta from mustart
  
  for (iter in 1:25L) {
    mu <- family$linkinv(eta)    ## compute the fitted values
    deriv <- family$mu.eta(eta)  ## compute the derivative d mu/d eta
    w <- (y - mu)/deriv          ## compute the working residuals
    Z <- eta + w                 ## compute the adjusted response
    W <- weights*deriv^2/family$variance(mu)  ## compute the working weights
    lm.fit <- lm(Z ~ X - 1, weights = W)  ## lm with weights
    beta <- lm.fit$coef  ## extract the estimates
    eta <- X %*% beta    ## compute the linear predictor
  }
  beta  ## returns estimates
}
```


## Iterations in action `r .emo("practice")`

The algorithm usually converges quickly:

```{r iterations}
foo <- glm(size ~ humans_eaten, data = Aliens, family = gaussian(), control = list(trace = TRUE))
foo <- glm(eggs ~ humans_eaten, data = Aliens, family = poisson(), control = list(trace = TRUE))
```

Note: check other options for the iterative procedure in `?glm.control`


## A `glm.fit()` mockup `r .emo("nerd")`  `r .emo("nerd")`  `r .emo("nerd")`

```{r}
my_glm3 <- function(formula, data, family, weights = NULL) {
  useful.data <- model.frame(formula = formula, data = data)  ## keep only useful rows and columns
  X <- model.matrix(object = formula, data = useful.data)     ## build design matrix
  y <- model.response(useful.data)                            ## extract response
  nobs <- NROW(y) ## count the rows of observations (not nrow() as y can be a vector or a matrix)
  if (is.null(weights)) weights <- rep(1, nobs)  ## set the prior.weights if NULL
  etastart <- start <- mustart <- NULL  ## required for family$initialize
  eval(family$initialize)         ## creates mustart -- the initial values for mu -- and 
                                  ## recomputes y and weights if binomial + cbind
  eta <- family$linkfun(mustart)  ## compute initial values for eta from mustart
  devold <- Inf  ## set old residual deviance | glm.fit does something more clever here
  for (iter in 1:25L) {
    mu <- family$linkinv(eta)    ## compute the fitted values
    deriv <- family$mu.eta(eta)  ## compute the derivative d mu/d eta
    w <- (y - mu)/deriv          ## compute the working residuals
    Z <- eta + w                 ## compute the adjusted response
    W <- weights*deriv^2/family$variance(mu)  ## compute the working weights
    dev <- sum(family$dev.resids(y, mu, weights))        ## compute residual deviance
    if (abs(dev - devold)/(0.1 + abs(dev)) < 1e-7) break ## leave loop if it has converged
    devold <- dev  ## update residual deviance
    QR <- qr(X*sqrt(W))  ## perform the QR decomposition | glm.fit uses C_Cdqrls directly.
    beta <- qr.coef(QR, Z*sqrt(W))  ## compute coefficients | glm.fit uses C_Cdqrls directly.
    eta <- as.numeric(X %*% beta)  ## compute the predicted values on the scale of the linear predictor
  }
  if (iter == 25L) warning("the algorithm did not converge after 25 iterations...")
  beta  ## returns estimates
}
```

<!--
## Testing our toy functions

```{r}
rbind(coef(fit_gauss),  ## from glm()
      my_glm1(size ~ humans_eaten, data = Aliens, family = gaussian()),
      my_glm2(size ~ humans_eaten, data = Aliens, family = gaussian()),
      my_glm3(size ~ humans_eaten, data = Aliens, family = gaussian())
      )
```

```{r}
rbind(coef(fit_poiss),  ## from glm()
      my_glm1(eggs ~ humans_eaten, data = Aliens, family = poisson()),
      my_glm2(eggs ~ humans_eaten, data = Aliens, family = poisson()),
      my_glm3(eggs ~ humans_eaten, data = Aliens, family = poisson())
      )
```


## Testing our toy functions

```{r}
rbind(fit_binom$coefficients,
      my_glm1(cbind(blue_eyes, pink_eyes) ~ humans_eaten, data = Aliens, family = binomial()),
      my_glm2(cbind(blue_eyes, pink_eyes) ~ humans_eaten, data = Aliens, family = binomial()),
      my_glm3(cbind(blue_eyes, pink_eyes) ~ humans_eaten, data = Aliens, family = binomial())
      )
```

```{r, warning=FALSE}
Aliens$prop_blue <- Aliens$blue_eyes / (Aliens$blue_eyes + Aliens$pink_eyes)
rbind(fit_binom$coefficients,
      my_glm1(prop_blue ~ humans_eaten, data = Aliens, family = binomial(), weights = Aliens$all_eyes),
      my_glm2(prop_blue ~ humans_eaten, data = Aliens, family = binomial(), weights = Aliens$all_eyes),
      my_glm3(prop_blue ~ humans_eaten, data = Aliens, family = binomial(), weights = Aliens$all_eyes)
      )
```
-->


# Fitting troubles

## The iterative algorithm can sometimes fail `r .emo("broken")`

```{r, error = TRUE}
test <- data.frame(x = c(8.752, 20.27, 24.71, 32.88, 27.27, 19.09),
                   y = c(5254, 35.92, 84.14, 641.8, 1.21, 47.2))
glm(y ~ x, data = test, family = Gamma(link = "log")) 
```

## Coping with iterative algorithm failures `r .emo("practice")`

You may try to play with parameters of the iterative algorithm...

```{r}
glm(y ~ x, data = test, family = Gamma(link = "log"), start = c(10, -0.1), control = list(epsilon = 1e-3))
```

... but unless you really know what you are doing (& I don't), you risk to make the optimisation algorithm stop prematurely without warning.


## Coping with iterative algorithm failures `r .emo("practice")`

You may also try other fitting methods (safer & easier):

```{r, error = TRUE, message = FALSE}
library(spaMM) ## you need to load spaMM explicitly for that
glm(y ~ x, data = test, family = Gamma(link = "log"), method = "spaMM_glm.fit")
```

## Coping with iterative algorithm failures `r .emo("practice")`

You may also try other fitting functions (which is equivalent to selecting alternative fitting methods):

```{r, error = TRUE, results='hold'}
fitme(y ~ x, data = test, family = Gamma(link = "log"))
```

<!--
## The iterative algorithm can sometimes fail

```{r, error = TRUE}
test <- data.frame(x = c(8.752, 20.27, 24.71, 32.88, 27.27, 19.09),
                   y = c(5254, 35.92, 84.14, 641.8, 1.21, 47.2))
glm(y ~ x, data = test, family = Gamma(link = "log")) 
```

In such cases, you may also try other methods (much easier):

```{r, error = TRUE, message = FALSE, warning = FALSE}
library(glm2)   ## seems less robust than spaMM in at least some other cases, so try both in case of troubles
glm2(y ~ x, data = test, family = Gamma(link = "log"))
```
-->

## Coping with iterative algorithm failures `r .emo("practice")`

Just for fun let's try our toy functions too:

```{r, error = TRUE}
my_glm1(y ~ x, data = test, family = Gamma(link = "log"))  ## Yay! It works
my_glm2(y ~ x, data = test, family = Gamma(link = "log"))  ## Silly results
my_glm3(y ~ x, data = test, family = Gamma(link = "log"))  ## Crash
```


# GLM vs LM with transformation

## The example of the GLM Poisson `r .emo("info")`

If the process generating the data is not Gaussian fits will be more difficult...

```{r, error = TRUE}
manyAliens <- simulate_Aliens_GLM(N = 1000)
fit_lm <- lm(log(eggs) ~ humans_eaten, data = manyAliens) ## fails due to log(0)
```

<br>

...and parameter estimates may be substantially biased:

```{r}
fit_lm2 <- lm(log(eggs+1) ~ humans_eaten, data = manyAliens) ## runs since no longer 0s
fit_glm <- glm(eggs ~ humans_eaten, family = poisson(), data = manyAliens)
rbind(ground_truth = attr(manyAliens, "param.eta")$eggs,
      GLM_Poisson = fit_glm$coefficients,
      LM = fit_lm2$coefficients)
```


## The example of the GLM Poisson `r .emo("info")`

If the process generating the data is not Gaussian the assumptions for LM will also often not be met:

```{r, fig.width = 10, fig.height = 3}
par(mfrow = c(1, 4))
plot(fit_lm2)
```


## But, if the process is Gaussian... `r .emo("info")`

```{r gauss vs log, warning = FALSE}
set.seed(1L)
x <- seq(1, 2, length = 100)
simu <- replicate(100, {
  y <- exp(rnorm(n = length(x), mean = 2 + 1 * x, sd = 2)) ## exp(mean_resp + error)
  fit_lm   <- lm(log(y) ~ x)
  fit_glm  <- glm(y ~ x, family = gaussian(link = "log"))
  c(coef(fit_lm), coef(fit_glm))
})

round(rbind(
  ground_truth = c("mean int" = 2, "sd int" = NA,
                   "mean slope" = 1, "sd slope" = NA),
  LM = c("mean int" = mean(simu[1, ]), "sd int" = sd(simu[1, ]),
         "mean slope" = mean(simu[2, ]), "sd slope" = sd(simu[2, ])),
  GLM = c(mean(simu[3, ]), sd(simu[3, ]), mean(simu[4, ]), sd(simu[4, ]))), 3)
```

Note: in `fit_glm` the expected values are log transformed, not the residuals! So the estimates produced by this fit are biased.


## But, if the process is Gaussian... `r .emo("info")`

```{r gauss vs log 2}
set.seed(1L)
x <- seq(1, 2, length = 100)
simu <- replicate(100, {
  y <- rnorm(n = length(x), mean = exp(2 + 1 * x), sd = 2) ## exp(mean_resp) + error
  fit_lm   <- lm(log(y) ~ x)
  fit_glm  <- glm(y ~ x, family = gaussian(link = "log"))
  c(coef(fit_lm), coef(fit_glm))
})

round(rbind(
  ground_truth = c("mean int" = 2, "sd int" = NA,
                   "mean slope" = 1, "sd slope" = NA),
  LM = c("mean int" = mean(simu[1, ]), "sd int" = sd(simu[1, ]),
         "mean slope" = mean(simu[2, ]), "sd slope" = sd(simu[2, ])),
  GLM = c(mean(simu[3, ]), sd(simu[3, ]), mean(simu[4, ]), sd(simu[4, ]))), 3)
```

Note: here `fit_glm` corresponds to the simulation model, so the estimates produced are better than the alternatives.


## How to compare LM and GLM? `r .emo("practice")`

We can compare deviances (smaller = better), but not directly:

```{r}
set.seed(1L)
x <- seq(1, 2, length = 100)
y <- rnorm(n = length(x), mean = exp(2 + 1 * x), sd = 5)

fit_glm  <- glm(y ~ x, family = gaussian(link = "log"))
fit_lm   <- lm(log(y) ~ x)

rbind("deviance GLM"          = deviance(fit_glm),
      "raw deviance LM"       = deviance(fit_lm),  ## not comparable as response = log(y)
      "corrected deviance LM" = sum((y - exp(predict(fit_lm)))^2))  ## comparable deviance for lm
```

<br>

Note: you cannot use AIC in this case since the responses differ!

# Predictions

## 2 scales for predictions `r .emo("practice")`

### You can predict outcomes for 2 different scales

- for the scale of the **linear predictor** (default for glm), e.g.:

```{r, predict_link_scale}
predict(fit_binom, newdata = data.frame(humans_eaten = 1:5), type = "link")
```

Note: the numbers correspond here to logits since it is the link function used in this fit.

<br>
 
- for the scale of the **response**, e.g.:

```{r, predict_resp_scale}
predict(fit_binom, newdata = data.frame(humans_eaten = 1:5), type = "response")
```

Note: the numbers correspond here to probabilities since it is a binomial fit.

## 2 scales for predictions `r .emo("info")`

```{r, predict_both_scales, echo = FALSE, fig.width = 10}
library(ggplot2)
p_link <- predict(fit_binom, newdata = data.frame(humans_eaten = 1:5), type = "link")
p_resp <- predict(fit_binom, newdata = data.frame(humans_eaten = 1:5), type = "response")

logit_data <- data.frame(x = x <- seq(-4, 4, length = 100), y = plogis(x))
pred_data <- as.data.frame(cbind(humans_eaten = 1:5, link = p_link, response = p_resp))
pred_data_expanded <- data.frame(x = x <- seq(0, 10, length = 100),
                                 y = predict(fit_binom, newdata = data.frame(humans_eaten = x), type = "response"))


ggplot() +
  geom_abline(slope = coef(fit_binom)["humans_eaten"][[1]], intercept = coef(fit_binom)["(Intercept)"][[1]], linetype = "dashed") +
  geom_point(aes(y = link, x = humans_eaten, colour = factor(humans_eaten)), data = pred_data) +
  scale_x_continuous(limits = c(0, 10), breaks = 0:10) +
  scale_y_continuous(limits = c(-4, 4), breaks = seq(-4, 4, by = 2)) +
  labs(y = "prediction (link scale)", x = "humans eaten", colour = "humans eaten") +
  theme_bw() +
  theme(legend.position = "top") -> p1

ggplot() +
  geom_line(aes(y = y, x = x), data = pred_data_expanded, linetype = "dashed") +
  geom_point(aes(y = response, x = humans_eaten, colour = factor(humans_eaten)), data = pred_data) +
  scale_x_continuous(limits = c(0, 10), breaks = 0:10) +
  labs(y = "prediction (response scale)", x = "humans eaten", colour = "humans eaten") +
  scale_y_continuous(limits = c(0, 1)) +
  theme_bw() +
  theme(legend.position = "top") -> p2

ggplot() +
    geom_line(aes(y = y, x = x), data = logit_data, linetype = "dashed") +
    geom_point(aes(y = response, x = link, colour = factor(humans_eaten)), data = pred_data) +
    labs(y = "prediction (response scale)", x = "prediction (link scale)", colour = "humans eaten") +
    theme_bw() +
    theme(legend.position = "top") -> p3

library(patchwork)
p1 + p2 + p3
```

## Predictions on the linear scale `r .emo("recap")`

The predictions on the linear scale work just as we saw it for LMs.

How do we interpret the coefficients?

```{r}
summary(fit_binom)
```

\pause

```{r, predict_both_scales, echo = FALSE, fig.width = 3.5}
ggplot() +
  geom_abline(slope = coef(fit_binom)["humans_eaten"][[1]], intercept = coef(fit_binom)["(Intercept)"][[1]], linetype = "dashed") +
  geom_point(aes(y = link, x = humans_eaten, colour = factor(humans_eaten)), data = pred_data) +
  scale_x_continuous(limits = c(0, 10), breaks = 0:10) +
  scale_y_continuous(limits = c(-4, 4), breaks = seq(-4, 4, by = 2)) +
  labs(y = "prediction (link scale)", x = "humans eaten", colour = "humans eaten") +
  theme_bw() +
  theme(legend.position = "top")
```

## Predictions on the response scale `r .emo("info")`

Doing the right thing can be tricky (even when a single predictor is involved)...

### Why?

Because on the response scale the prediction associated with a mean value is not equal to the mean of predictions associated with each values.

```{r, prediction_of_means_data_prep}
Aliens2       <- data.frame(humans_eaten = rpois(n = 1000, lambda = 4)) ## new data making difference obvious
Aliens2$happy <- rbinom(n  = 1000, size = 1, prob = plogis(-3 + 0.3 * Aliens2$humans_eaten))
fit_binar2    <- glm(happy ~ humans_eaten, family = binomial(), data = Aliens2)
```

```{r, prediction_of_means}
predict(fit_binar2, newdata = data.frame(humans_eaten = mean(Aliens2$humans_eaten)), type = "response")[[1]]
mean(predict(fit_binar2, newdata = data.frame(humans_eaten = Aliens2$humans_eaten), type = "response"))
```

Note: considering a predictor at its mean no longer gives the mean expectation.


## Predictions on the response scale `r .emo("practice")`

Let's illustrate the issue on a more realistic model fit:

```{r, fit_titanic}
fit_titanic <- glm(survived ~ sex*age*passengerClass, data = TitanicSurvival, family = binomial())
```

- Naive (incorrect) computation of the mean survival for all sex - class combinations:

```{r,  prediction_of_means2}
newdata1 <- expand.grid(sex = levels(model.frame(fit_titanic)$sex),
                        passengerClass = levels(model.frame(fit_titanic)$passengerClass),
                        age = mean(model.frame(fit_titanic)$age))
preds1 <- predict(fit_titanic, newdata = newdata1, type = "response")
cbind(newdata1, survival = preds1)
```

## Predictions on the response scale `r .emo("practice")`

Let's illustrate the issue on a more realistic model fit:

```{r, fit_titanic2, eval=FALSE}
fit_titanic <- glm(survived ~ sex*age*passengerClass, data = TitanicSurvival, family = binomial())
```

- More accurate (marginal) computation of the mean survival for all sex - class combinations:

```{r,  prediction_of_means3, message=FALSE}
newdata2 <- expand.grid(sex = levels(model.frame(fit_titanic)$sex),
                        passengerClass = levels(model.frame(fit_titanic)$passengerClass),
                        age = model.frame(fit_titanic)$age)
preds2 <- cbind(newdata2, survival = predict(fit_titanic, newdata = newdata2, type = "response"))
output <- aggregate(survival ~ sex + passengerClass, data = preds2, FUN = mean)
output$wrong_survival <- preds1
output
```

## Predictions on the response scale `r .emo("practice")`

Let's illustrate the issue on a more realistic model fit:

```{r, fit_titanic3, eval=FALSE}
fit_titanic <- glm(survived ~ sex*age*passengerClass, data = TitanicSurvival, family = binomial())
```

- More accurate (conditional) computation of the mean survival for all sex - class combinations:

```{r,  prediction_of_means4}
wrong_data <- aggregate(age ~ sex + passengerClass, data = model.frame(fit_titanic), FUN = mean) #Predict at the mean of predictors. Incorrect
wrong_pred <- predict(fit_titanic, newdata = wrong_data, type = "response")
newdata3 <- aggregate(age ~ sex + passengerClass, data = model.frame(fit_titanic), FUN = identity) #Mean prediction across predictor values. Correct
newdata3 <- tidyr::unnest(newdata3, cols = age) ## unnest the list column storing ages
preds3 <- cbind(newdata3, survival = predict(fit_titanic, newdata = newdata3, type = "response"))
output <- aggregate(survival ~ sex + passengerClass, data = preds3, FUN = mean)
output$wrong_survival <- wrong_pred
output
```


## Predictions on the response scale `r .emo("practice")`

Things can be easier with **{spaMM}** provided that:

- you focus on one focal predictor
- you want partial-dependence effects

```{r,  fit_titanic4}
fit_titanic_spaMM <- fitme(survived ~ sex*age*passengerClass, data = TitanicSurvival, family = binomial())
```

```{r,  prediction_of_means7}
pdep_effects(fit_titanic_spaMM, focal_var = "passengerClass")
```


## Predictions on the response scale `r .emo("practice")`

Things can be easier with **{spaMM}** provided that:

- you focus on one focal predictor
- you want partial-dependence effects

```{r,  fit_titanic5, eval=FALSE}
fit_titanic_spaMM <- fitme(survived ~ sex*age*passengerClass, data = TitanicSurvival, family = binomial())
```

Doing the same by hand:
```{r,  prediction_of_means6}
newdata1 <- newdata2 <- newdata3 <- model.frame(fit_titanic_spaMM)
newdata1["passengerClass"] <- "1st"
newdata2["passengerClass"] <- "2nd"
newdata3["passengerClass"] <- "3rd"
newdata <- rbind(newdata1, newdata2, newdata3)
preds <- cbind(newdata, survival = predict(fit_titanic_spaMM, newdata = newdata))
aggregate(survival ~ passengerClass, data = preds, FUN = mean)
```


# Interpreting regression coefficients

## Estimates with the Gaussian family `r .emo("recap")`

<br>

<center> Same as LM (when link = identity)! </center>


## Estimates with the Poisson family with log link `r .emo("info")`

### Predictions for $\eta$ and $\mu$ differ!

```{r}
coef(fit_poiss)
predict(fit_poiss, newdata = data.frame(humans_eaten = 10)) #Linear predictor scale by default
coef(fit_poiss)[1] + 10 * coef(fit_poiss)[2]
```

Because we used the link function ```log```, this is a log number of eggs!


## Estimates with the Poisson family with log link `r .emo("info")`

### Predictions for $\eta$ and $\mu$ differ!

```{r}
coef(fit_poiss)
predict(fit_poiss, newdata = data.frame(humans_eaten = 10), type = "response")
exp(coef(fit_poiss)[1] + 10 * coef(fit_poiss)[2])
```

This is the average number of eggs predicted for aliens eating 10 humans.


## Estimates with the Poisson family with log link `r .emo("info")`

The model parameters are linear for $\eta$, not $\mu$!

What happens when different aliens eat one additional human? `r .emo("")`

```{r}
(p.eta <- predict(fit_poiss, newdata = data.frame(humans_eaten = 1:5))) #Linear predictor scale by default
diff(p.eta)
```

<br>

Note: the effect is constant at the scale of the linear predictor.


## Estimates with the Poisson family with log link `r .emo("info")`

The model parameters are linear for $\eta$, not $\mu$!

What happens when different aliens eat one additional human?

```{r}
(p.mu <- predict(fit_poiss, newdata = data.frame(humans_eaten = 1:5), type = "response"))
diff(p.mu)
```

<br>

Note: the effect is NOT constant at the scale of the response.


## Estimates with the Poisson family with log link `r .emo("practice")`

Ratios are the way out:

```{r}
p.mu[-1] / p.mu[-5]
exp(coef(fit_poiss)[2])
```

Eating one additional human increases the average number of eggs produced by aliens by ```r exp(fit_poiss$coefficients[2])``` times, which is equivalent to ```r 100*exp(fit_poiss$coefficients[2])-100```%.

Note: if you change the link function, the recipe will be different!


## Estimates and the Binomial with logit link `r .emo("info")`

```{r}
(p.mu <- predict(fit_binom, newdata = data.frame(humans_eaten = 1:5), type = "response"))
diff(p.mu)
```

<br>

Note: the effect is also not constant at the scale of the response!


## Estimates and the Binomial with logit link `r .emo("practice")`

Odds ratios are the way out:

```{r}
odds <- p.mu / (1 - p.mu)
odds[-1] / odds[-5]
```

The effect is constant on the scale of the odds ratios!

<br>

And the odds ratios can also be directly deduced from the parameter estimates:

```{r}
exp(coef(fit_binom)[2])
```

Eating one additional human increases the **odd** of aliens having blue eyes by ```r round(exp(fit_binom$coefficients[2]), 2)``` times.


## Estimates and the Binomial with logit link `r .emo("practice")`

What happens when different aliens eat 10 additional humans?

```{r}
p.mu <- predict(fit_binom, newdata = data.frame(humans_eaten = c(0, 10, 20, 30, 40)), type = "response")
odd <- p.mu / (1 - p.mu)
odd[-1] / odd[-5]
```

<br>

It is also constant and predicted from:

```{r}
exp(fit_binom$coefficients[2] * 10)
```

<br>

Note: as odds-ratios are ratios, a value of 0.1 means a 10 (=1/0.1) fold reduction in odds.


# Testing regression coefficients

<!--
## Covariances between parameter estimates

As for LM, you can extract covariance matrices using `vcov()`:

```{r vcov gaussian}
vcov(fit_gauss)
vcov(fit_poiss)
```

### gaussian family

<br>

```{r vcov gaussian bis}
X <- model.matrix(fit_gauss)
W <- matrix(0, ncol = nrow(fit_gauss$model), nrow = nrow(fit_gauss$model))
diag(W) <- fit_gauss$weights
XTWX <- t(X) %*% W %*% X
phi <- fit_gauss$deviance / fit_gauss$df.residual
phi*solve(XTWX)
vcov(fit_gauss)
```

## Covariances between parameter estimates

### gaussian family

<br>

Same using the QR decomposition:

```{r vcov gaussian 2}
(XTWX <- crossprod(fit_gauss$R))
phi <- fit_gauss$deviance / fit_gauss$df.residual
phi*solve(XTWX)
```


## Covariances between parameter estimates

### Poisson family

<br>

```{r vcov poisson}
vcov(fit_poiss)
XTWX <- crossprod(fit_poiss$R)
phi <- 1
phi*solve(XTWX)
```


## Covariances between parameter estimates

### Binomial family (binary case)

<br>

```{r vcov binar}
vcov(fit_binar)
XTWX <- crossprod(fit_binar$R)
phi <- 1
phi*solve(XTWX)
```

## Covariances between parameter estimates

### Binomial family (general case)

<br>

```{r vcov binom}
vcov(fit_binom)
XTWX <- crossprod(fit_binom$R)
phi <- 1
phi*solve(XTWX)
```

-->


## Summary: gaussian `r .emo("info")`

```{r summary gauss2}
summary(fit_gauss)
```


## Summary: gaussian `r .emo("recap")`

### Just like LM!
```{r  summary lm2 gauss2}
summary(fit_gauss)$coefficients
summary(fit_gauss_lm)$coefficients
```


## Summary: other families `r .emo("practice")`

Example: Poisson

```{r summary poiss}
summary(fit_poiss)$coefficients
z <- (coef(fit_poiss) - 0) / sqrt(diag(vcov(fit_poiss)))
pvalues <- 2 * pnorm(abs(z), lower.tail = FALSE)
cbind(z, pvalues)
```

```{r summary poiss spaMM, eval=FALSE}
summary(fit_poiss_spaMM_ML, details = c(p_value = "Wald"), verbose = FALSE)$beta_table ## same with spaMM
```

Note: when the dispersion parameter is not estimated, `summary` uses $z$, not $t$. Yet, these so-called "Wald tests" are not very reliable and it is better to rely on CI build using likelihood profiling or parametric bootstrap.


## Distribution of parameter estimates `r .emo("info")`

The parameter estimates are asymptotically normally distributed as in LM, but for non-Gaussian families small sample sizes deviate strongly from the asymptotic case. Example with large N:

```{r simu binar}
new_betas <- t(replicate(1000, update(fit_binar, data = simulate_Aliens_GLM(N = 1000))$coef))
```

<div class="columns-2">
```{r qq binar 1, fig.width=3, fig.height=3}
qqnorm(new_betas[, 1])
qqline(new_betas[, 1], col = "red", lwd = 2)
```

```{r qq binar 2, fig.width=3, fig.height=3}
qqnorm(new_betas[, 2])
qqline(new_betas[, 2], col = "red", lwd = 2)
```
</div>


## Distribution of parameter estimates `r .emo("warn")`

The parameter estimates are asymptotically normally distributed as in LM, but for non-Gaussian families small sample sizes deviate strongly from the asymptotic case. Example with small N:

```{r simu binar bis, warning = FALSE}
new_betas <- t(replicate(1000, update(fit_binar, data = simulate_Aliens_GLM(N = 10))$coef))
```

<div class="columns-2">
```{r qq binar bis 1, fig.width=3, fig.height=3}
qqnorm(new_betas[, 1])
qqline(new_betas[, 1], col = "red", lwd = 2)
```

```{r qq binar bis 2, fig.width=3, fig.height=3}
qqnorm(new_betas[, 2])
qqline(new_betas[, 2], col = "red", lwd = 2)
```
</div>

Note: this is why the so-called Wald tests shown earlier are not great...


## Anova: gaussian `r .emo("info")`

For `gaussian()` (and `quasiXXX()`) the F-test is the one you should use (as we have seen it with LM):

```{r Anova gauss}
car::Anova(fit_gauss, test = "F")
```


<!--
## Anova: gaussian

```{r LR gauss}
car::Anova(fit_gauss, test = "LR")  ## Does not do true LRT here!!
anova(fit_gauss_spaMM_ML, fit_gauss_spaMM0)  ## True LRT, "same" with spaMM
c(-2 * (logLik(update(fit_gauss, . ~ 1)) - logLik(fit_gauss)), 
  (deviance(update(fit_gauss, . ~ 1)) - deviance(fit_gauss)) / summary(fit_gauss)$dispersion)
```

Note: the computation will differ between methods when dispersion is not equals to 1...
-->


## Anova: Poisson and Binomial `r .emo("info")` `r .emo("proof")`

What about a F-test?

```{r comp 2, warning = FALSE}
set.seed(1L)
N <- 20
pv <- replicate(1000, {
  d <- data.frame(eggs = rpois(N, lambda = 0.5),
                  mood = factor(sample(c("depr.", "happy", "upset", "tired", "anxious"), size = N, replace = TRUE)))
  fit_poiss_1      <- glm(eggs ~ mood, family = poisson(), data = d)
  c(F =  car::Anova(fit_poiss_1, test = "F")["mood", "Pr(>F)"],
    LR = car::Anova(fit_poiss_1, test = "LR")["mood", "Pr(>Chisq)"])
})
rbind(mean(pv["F", ] <= 0.05),
      mean(pv["LR", ] <= 0.05))
```

Note: the F-test performs very badly here, but the LRT is better.


## Anova by LRT `r .emo("practice")`

```{r Anova Poiss}
anova(update(fit_poiss, . ~ 1), fit_poiss, test = "LRT")
car::Anova(fit_poiss, test = "LR")
LR <- -2 * (logLik(update(fit_poiss, . ~ 1)) - logLik(fit_poiss))
pvalue <- pchisq(LR, df = 1, lower.tail = FALSE)
cbind(LR, pvalue)  ## the same
```


## Anova: LRT by parametric bootstrap `r .emo("practice")`

You can compare model fits this way using **{spaMM}**:

```{r lrt boot 2, results='hide', cache=TRUE}
fit_poiss_spaMM0 <- update(fit_poiss_spaMM_ML, . ~ 1)
res <- anova(fit_poiss_spaMM_ML, fit_poiss_spaMM0, boot.repl = 1000)
```
```{r lrt boot 2 show}
res 
```

## Anova: LRT by parametric bootstrap `r .emo("practice")`

You can also compare model fits this way using **{pbkrtest}** for `stats::glm()` fits:

```{r lrt boot pbkrtest, cache=TRUE}
fit_poiss_0 <- update(fit_poiss, . ~ 1)
res <- pbkrtest::PBmodcomp(fit_poiss, fit_poiss_0, nsim = 1000, seed = 123)
```
```{r lrt boot pbkrtest show}
res 
```

## Anova: LRT by parametric bootstrap `r .emo("nerd")`

Let us compare the 2 versions of the LRT using parametric bootstrap `r .emo("slow")` `r .emo("slow")` `r .emo("slow")`

```{r test by boot, warning = FALSE, results='hide', cache=TRUE}
set.seed(1L)
N <- 20
pv_boot <- replicate(1000, {
  d <- data.frame(eggs = rpois(N, lambda = 0.5),
                  mood = factor(sample(c("depr.", "happy", "upset", "tired", "anxious"), size = N, replace = TRUE)))
  fit_poiss_1      <- fitme(eggs ~ mood, family = poisson(), data = d)
  fit_poiss_1_null <- fitme(eggs ~ 1, family = poisson(), data = d)
  boot <- anova(fit_poiss_1, fit_poiss_1_null, boot.repl = 200)
  c(raw = boot$rawBootLRT$p_value,
    Bart = boot$BartBootLRT$p_value)
})
```
```{r}
rbind(mean(pv_boot["Bart", ] <= 0.05), mean(pv_boot["raw", ] <= 0.05))
```

Note: applying the Bartlett correction is not as accurate (but it should become best if `boot.repl` is small).


## Comparing the four possible test `r .emo("info")`

Let us compare the performances of all 4 tests (a perfect test would follow the dashed red line)

```{r ecfd spaMM 1, echo = FALSE, fig.width = 7, fig.height = 5}
par(las = 1)
plot(ecdf(pv["F", ]), col = "blue", cex = 0.5, main = "")
plot(ecdf(pv["LR", ]), col = "green", add = TRUE, cex = 0.5)
plot(ecdf(pv_boot["Bart", ]), col = "orange", add = TRUE, cex = 0.5)
plot(ecdf(pv_boot["raw", ]), col = "red", add = TRUE, cex = 0.5)
legend("topleft", fill = c("blue", "green", "orange", "red"), legend = c("F", "LR", "Param boot Bartlett", "Param boot raw"), bty = "n")
abline(0, 1, col = "black", lty = 2, lwd = 2)
```


## Comparing the four possible test `r .emo("info")`

Let us compare the performances of all 4 tests (a perfect test would follow the dashed red line)

```{r ecfd spaMM 2, echo = FALSE, fig.width = 7, fig.height = 5}
par(las = 1)
plot(ecdf(pv["F", ]), xlim = c(0, 0.1), ylim = c(0, 0.1), col = "blue", cex = 0.5, main = "")
plot(ecdf(pv["LR", ]), col = "green", add = TRUE, cex = 0.5)
plot(ecdf(pv_boot["Bart", ]), col = "orange", add = TRUE, cex = 0.5)
plot(ecdf(pv_boot["raw", ]), col = "red", add = TRUE, cex = 0.5)
legend("topleft", fill = c("blue", "green", "orange", "red"), legend = c("F", "LR", "Param boot Bartlett", "Param boot raw"), bty = "n")
abline(0, 1, col = "black", lty = 2, lwd = 2)
```
Note: here the F-test and asymptotic LRT are **anticonservative** `r .emo("warn")`

# Confidence intervals of parameter estimates

## Which methods for computing CIs? `r .emo("practice")`

For the Gaussian family, do as in LM! (Use the quantiles from the Student's distribution or `confint()`).

For families with non-estimated variances (e.g. binomial, poisson...), CI based on the Gaussian or the Student's t distributions do not work well and instead you should build CI by likelihood profiling or parametric bootstrap.

If **{MASS}** is installed, calling `confint()` on a model fitted with `stats::glm()` directly does build CI by likelihood profiling:

```{r profiling}
confint(fit_poiss)
```


## Which methods for computing CIs? `r .emo("practice")`

For the Gaussian family, do as in LM! (Use the quantiles from the Student's distribution or `confint()`).

For families with non-estimated variances (e.g. binomial, poisson...), CI based on the Gaussian or the Student's t distributions do not work well and instead you should build CI by likelihood profiling or parametric bootstrap.

For model fitted with `spaMM::fitme()`, calling `confint()` also performs likelihood profiling:

```{r profiling spaMM confint}
confint(fit_poiss_spaMM_ML, parm = "(Intercept)")
confint(fit_poiss_spaMM_ML, parm = "humans_eaten")
```


## Which methods for computing CIs? `r .emo("practice")`

For the Gaussian family, do as in LM! (Use the quantiles from the Student's distribution or `confint()`).

For families with non-estimated variances (e.g. binomial, poisson...), CI based on the Gaussian or the Student's t distributions do not work well and instead you should build CI by likelihood profiling or parametric bootstrap.

For model fitted with `spaMM::fitme()`, calling `confint()` can also be used to compute CI using parametric bootstrap if the argument `boot_args` is correctly set:

```{r profiling spaMM confint boot, cache=TRUE, results='hide'}
res_boot <- confint(fit_poiss_spaMM_ML, parm = "humans_eaten", boot_args = list(nsim = 1000))
```


## Which methods for computing CIs? `r .emo("practice")`
```{r, profiling spaMM confint boot out}
res_boot
res_boot$normal  ## do not use the other types here!
```


## Which methods for computing CIs? `r .emo("practice")`

Which method is best depends on the particular circumstances, but you can (as always) test for yourself using simulations `r .emo("slow")` `r .emo("slow")` `r .emo("slow")`
```{r test coverage confint, results='hide', cache=TRUE, warning=FALSE}
set.seed(1L)
N <- 20
coverage_small_list <- replicate(1000, {
  d <- simulate_Aliens_GLM(N = N)
  fit_poiss_spaMM_ML <- fitme(eggs  ~ humans_eaten, family = poisson(),  data = d)
  ref <- attr(d, "param.eta")$eggs["slope"] ## 0.1
  prof <- confint(fit_poiss_spaMM_ML, parm = "humans_eaten")
  boot <- confint(fit_poiss_spaMM_ML, parm = "humans_eaten", boot_args = list(nsim = 200))
  c(prof = sum(findInterval(ref, prof$interval)) == 1,
    boot_norm = sum(findInterval(ref, boot$normal[2:3])) == 1)
}, simplify = FALSE)
```
```{r test coverage confint res}
apply((do.call("rbind", coverage_small_list)), 2, mean, na.rm = TRUE)
```
Note: here the construction of CI by likelihood profiling behaves best.


# Intervals for new observations

## CI using Gaussian approximation `r .emo("practice")`

There is no longer a direct method offered by `predict()` for models fitted with `stats::glm()`:
```{r conf fitted poiss}
newdata <- data.frame(humans_eaten = 3:4)
pred <- predict(fit_poiss, newdata = newdata, se.fit = TRUE)
lwr_eta <- pred$fit + qnorm(0.025) * pred$se.fit
upr_eta <- pred$fit + qnorm(0.975) * pred$se.fit
cbind(pred = poisson(link = "log")$linkinv(pred$fit),
      lwr = poisson(link = "log")$linkinv(lwr_eta),
      upr = poisson(link = "log")$linkinv(upr_eta))
```
Notes:

- don't use `type = "response"` in `predict()` because transformations need to happen after CI are computed!

- for proper predictions involving more predictors, better use `pdep_effects()` on a model fitted with **{spaMM}** `r .emo("recap")`

## CI using Gaussian approximation `r .emo("practice")`

CIs are assumed symmetrical (i.e. Gaussian) on the **linear predictor scale** but not on the **response scale**.

```{r conf fitted poiss}
newdata <- data.frame(humans_eaten = 0:1)
pred_wrong <- predict(fit_poiss, newdata = newdata, se.fit = TRUE, type = "response")
lwr_eta <- pred_wrong$fit + qnorm(0.025) * pred_wrong$se.fit
upr_eta <- pred_wrong$fit + qnorm(0.975) * pred_wrong$se.fit
cbind(pred = pred_wrong$fit,
      lwr = lwr_eta,
      upr = upr_eta)
```

```{r, fig.width = 4, echo = FALSE}
newdata <- data.frame(humans_eaten = 0:1)
#Correct method. CIs on linear predictor scale
pred <- predict(fit_poiss, newdata = newdata, se.fit = TRUE)
lwr_eta <- pred$fit + qnorm(0.025) * pred$se.fit
upr_eta <- pred$fit + qnorm(0.975) * pred$se.fit
plot_data <- data.frame(humans_eaten = 0:1,
                        eggs = poisson(link = "log")$linkinv(pred$fit),
                        lwr = poisson(link = "log")$linkinv(lwr_eta),
                        upr = poisson(link = "log")$linkinv(upr_eta),
                        method = "link")
#Wrong method. CIs on response scale
pred_wrong <- predict(fit_poiss, newdata = newdata, se.fit = TRUE, type = "response")
lwr_eta <- pred_wrong$fit + qnorm(0.025) * pred_wrong$se.fit
upr_eta <- pred_wrong$fit + qnorm(0.975) * pred_wrong$se.fit
plot_data <- rbind(plot_data,
                   data.frame(humans_eaten = 0:1,
                              eggs = pred_wrong$fit,
                              lwr = lwr_eta,
                              upr = upr_eta,
                              method = "response"))

ggplot(data = plot_data) +
  geom_hline(yintercept = 0) +
  geom_errorbar(aes(x = humans_eaten, ymin = lwr, ymax = upr, group = method),
                width = 0.25, position = position_dodge(width = 0.25)) +
  geom_point(aes(x = humans_eaten, y = eggs, group = method, shape = method),
             size = 3, fill = "grey",
             position = position_dodge(width = 0.25)) +
  scale_shape_discrete(labels = c("Linear predictor scale", "Response scale"), name = "") +
  scale_x_continuous(breaks = c(3, 4)) +
  theme_classic()

```

## CI using Gaussian approximation `r .emo("practice")`

It is much easier with **{spaMM}**:

```{r conf fitted poiss spaMM}
get_intervals(fit_poiss_spaMM_ML, newdata = newdata, intervals = "fixefVar")
pdep_effects(fit_poiss_spaMM_ML, focal_var = "humans_eaten", focal_values = 3:4)
```

Note: here the 2 methods are equivalent because only one predictor is considered in the model fit.


# Comparing non-nested models

## Comparing different links or different families `r .emo("info")`

Since the response variable does not change but models are not nested, you can use:

* likelihood (larger = better)
* AIC (smaller = better)

For actual testing, the only method available is the parametric bootstrap!

Note: using the deviance is not recommended because some family return a scaled deviance while other do not `r .emo("broken")`

<!--
## Comparing LM vs GLM

This is a special case because the responses differ! So only parametric bootstrap is possible:
```{r LM vs GLM}
set.seed(1L)
x <- seq(1, 2, length = 100)
y <- exp(rnorm(n = length(x), mean = 2 + 1 * x, sd = 0.5))

fit_lm   <- lm(log(y) ~ x)
fit_glm  <- glm(y ~ x, family = gaussian(link = "log"))
logLik_diff_H0 <- replicate(1000, {
  new.y <- as.matrix(exp(simulate(fit_lm)))
  #new.y <- as.matrix(simulate(fit_glm)) ## alternative
  logLik(lm(log(new.y) ~ x)) - logLik(glm(new.y ~ x, family = gaussian(link = "log")))
  })
logLik_diff_obs <- logLik(fit_lm) - logLik(fit_glm)
(sum(logLik_diff_obs < logLik_diff_H0) + 1) / (length(logLik_diff_H0) + 1)
```

Here we used `fit_lm` as the null model. In principle, we could used `fit_glm`, but in practice the function `simulate()` bugs from time to time for GLM with `gaussian()$link != "identity"`.
-->


# Residuals (an interlude before checking assumptions)

## `plot(model)` is useless for GLM `r .emo("info")`

```{r plot poiss, fig.width = 5, fig.height = 4}
par(mfrow = c(2, 2))
plot(fit_poiss)
```

Note: what we would expect to see here is no longer the same as for LM.

## Many types of residuals can be computed for GLM `r .emo("info")`

```{r residuals poiss}
rbind(
  residuals(fit_poiss, type = "deviance")[1:2],
  residuals(fit_poiss, type = "pearson")[1:2],
  residuals(fit_poiss, type = "working")[1:2],
  residuals(fit_poiss, type = "response")[1:2])
```

Note: we can compute even more types of residuals, such as the partial residuals...

<br>

Unfortunately, none of these residuals are particularly useful to check the model assumptions!


<!--
## Distribution

### They are not necessarily normaly distributed!

<div class="columns-2">

```{r ecdf resid poiss, echo = FALSE}
par(las = 1)
plot(ecdf(residuals(fit_poiss, type = "working")), col = 3, main = "fit_poiss")
plot(ecdf(residuals(fit_poiss, type = "deviance")), col = 1, add = TRUE)
plot(ecdf(residuals(fit_poiss, type = "pearson")), col = 2, add = TRUE)
plot(ecdf(residuals(fit_poiss, type = "response")), col = 4, add = TRUE)
legend("bottomright", fill = 1:4, bty = "n", legend = c("deviance", "pearson", "working", "response"), title = "Type of residuals:")
```

```{r ecdf resid binom, echo = FALSE}
par(las = 1)
plot(ecdf(residuals(fit_binom, type = "working")), col = 3, main = "fit_binom")
plot(ecdf(residuals(fit_binom, type = "deviance")), col = 1, add = TRUE)
plot(ecdf(residuals(fit_binom, type = "pearson")), col = 2, add = TRUE)
plot(ecdf(residuals(fit_binom, type = "response")), col = 4, add = TRUE)
legend("topleft", fill = 1:4, bty = "n", legend = c("deviance", "pearson", "working", "response"), title = "Type of residuals:")
```

</div>


## Response residuals

These are the residuals that we saw in LM: observed - predicted. Simple but pretty much useless.

```{r response res}
rbind(residuals(fit_gauss, type = "response")[1:2],
      (fit_gauss$y - fit_gauss$fitted.values)[1:2])
rbind(residuals(fit_poiss, type = "response")[1:2],
      (fit_poiss$y - fit_poiss$fitted.values)[1:2])
rbind(residuals(fit_binom, type = "response")[1:2],
      (fit_binom$y - fit_binom$fitted.values)[1:2])
```


## Pearson residuals

These are response residuals scaled by the square root of the variance function and accounting for their prior weights.

```{r pearson res}
variances <- poisson()$variance(fit_poiss$fitted.values)
rbind(residuals(fit_poiss, type = "pearson")[1:2],
      (residuals(fit_poiss, type = "response") * sqrt(fit_poiss$prior.weights / variances))[1:2])

variances <- binomial()$variance(fit_binom$fitted.values)
rbind(residuals(fit_binom, type = "pearson")[1:2],
      (residuals(fit_binom, type = "response") * sqrt(fit_binom$prior.weights / variances))[1:2])
```

These residuals are often use to check the quality of the fit.


## Pearson residuals

The sum of the squared Pearson residuals gives the Pearson $\chi^2$ statistic of goodness of fit:

```{r pearson dev binom}
(X <- sum(residuals(fit_binom, type = "pearson")^2))
```

This statistic is sometimes used to test the goodness of fit of the model (= assess the ability of a model to fit the data with no alternative in mind) or to test for overdispersion (= is the residual variance compatible with the variance function), but I would not recommend you to use such test because its usage is often not optimal (it requires a lot of replications within each grouping, the power is poor, and the coverage is often poor).

```{r pearson dev pois}
pchisq(X, fit_poiss$df.residual, lower.tail = FALSE)  ## Pearson goodness of fit test
X / fit_poiss$df.residual  ## measure of overdispersion
```


## Deviance residuals

These are the residuals that are minimized after the fit.

Contrary to those computed by ```family()$dev.resids``` they are here not squared.

<br>

```{r dev resid}
residuals(fit_poiss, type = "deviance")[1:2]
rbind((residuals(fit_poiss, type = "deviance")[1:2])^2,
      with(fit_poiss, poisson()$dev.resids(y = y, mu = fitted.values, wt = prior.weights))[1:2])
```


## Deviance residuals

The sum of the squared deviance residuals gives the residual (unscaled) deviance of the model fit:

```{r deviance}
c(sum(residuals(fit_gauss, type = "deviance")^2), deviance(fit_gauss))
c(sum(residuals(fit_poiss, type = "deviance")^2), deviance(fit_poiss))
c(sum(residuals(fit_binom, type = "deviance")^2), deviance(fit_binom))
```

<br>

These residuals are sometimes also used to check the quality of the fit, in the exact same fashion as the Pearson's residuals, but again (and for the same reasons), I would not recommend you to do that.


## Testing false positives in the goodness of fit tests

For ```fit_binom```

<div class = "columns-2">

```{r gof 5, fig.width = 3, fig.height = 3}
p <- replicate(1000, {
  d <- simulate_Aliens_GLM()
  m <- update(fit_binom, data = d)
  X <- sum(residuals(m, type = "pearson")^2)
  pchisq(X, m$df.residual, lower.tail = FALSE)
})
plot(ecdf(p))
abline(0, 1, lty = 2, col = "green")
```

```{r gof 6, fig.width = 3, fig.height = 3}
p <- replicate(1000, {
  d <- simulate_Aliens_GLM()
  m <- update(fit_binom, data = d)
  dev <- deviance(m)
  pchisq(dev, m$df.residual, lower.tail = FALSE)
})
plot(ecdf(p))
abline(0, 1, lty = 2, col = "green")
```

</div>


## Working residuals

These are the residuals used during the last iteration of the iterative fitting procedure. Useless otherwise.

```{r work res}
rbind(fit_poiss$residuals[1:2],
      ((fit_poiss$y - fit_poiss$fitted.values)/poisson()$mu.eta(fit_poiss$linear.predictors))[1:2])

rbind(fit_binom$residuals[1:2],
      ((fit_binom$y - fit_binom$fitted.values)/binomial()$mu.eta(fit_binom$linear.predictors))[1:2])
```


## Partial residuals

These are residuals expressed at the level of each predictor.

```{r partial res}
fit_UK_small <- glm(milk ~ sex + mother_weight + cigarettes, data = UK[1:10, ], family = poisson())
residuals(fit_UK_small, type = "partial")
```


## Partial residuals

These are residuals expressed at the level of each predictor.

Computation:
```{r compute partial res}
(p <- predict(fit_UK_small, type = "terms")) ## prediction expressed per predictor
c(sum(p[1, ]) + attr(p, "constant"), predict(fit_UK_small, type = "link")[1])
```


## Partial residuals

These are residuals expressed at the level of each predictor.

Computation:
```{r compute partial res 2}
rbind((p + residuals(fit_UK_small, type = "working"))[1, ],
      residuals(fit_UK_small, type = "partial")[1, ])
```

<br>

Partial residuals can be use to check for departure from linearity (concerns quantitative predictors).


## Partial residuals

```{r plot partial res, fig.width = 10, fig.height = 4}
library(car)
par(mfrow = c(1, 3))
crPlots(fit_UK_small, terms = ~ sex)
crPlots(fit_UK_small, terms = ~ mother_weight)
crPlots(fit_UK_small, terms = ~ cigarettes)
```


## Partial residuals

```{r plot partial res 2, fig.width = 10, fig.height = 4}
fit_UK <- glm(milk ~ sex + mother_weight + cigarettes, data = UK, family = poisson())
par(mfrow = c(1, 3))
crPlots(fit_UK, terms = ~ sex)
crPlots(fit_UK, terms = ~ mother_weight)
crPlots(fit_UK, terms = ~ cigarettes)
```
-->


## Computing residuals by simulation `r .emo("nerd")`

We can deduce useful residuals from the comparison of observations to values expected under the model assumption...

```{r bootstrap, fig.height = 3, fig.width = 3}
set.seed(1L)
simulated_data <- simulate(fit_poiss, 1000)
simulated_data_cont <- simulated_data + runif(nrow(simulated_data)*ncol(simulated_data), min = -0.5, max = 0.5)
hist(as.numeric(simulated_data_cont[1, ]), main = "distrib of first fitted value", nclass = 30)
abline(v = fit_poiss$y[1] + runif(1, min = -0.5, max = 0.5), col = "red", lwd = 2, lty = 2)
```
Note: here we compare the first observation (+/- a little noise) to 1000 response values simulated under the model fit.


## Computing residuals by simulation `r .emo("nerd")`

The idea of is to consider as residuals the quantile of each observation in its expected (simulated) distribution:
```{r bootstrap 2, fig.height = 3, fig.width = 3}
plot(ecdf_simulated_1 <- ecdf(unlist(simulated_data_cont[1, ])), main = "cdf of first fitted value")
noise <- runif(1, min = -0.5, max = 0.5)
observed_quantile_1_cont <-  ecdf_simulated_1(fit_poiss$y[1] + noise)
segments(x0 = fit_poiss$y[1] + noise, y0 = 0, y1 =  observed_quantile_1_cont, col = "red", lwd = 2)
arrows(x0 = fit_poiss$y[1] + noise, x1 = -1, y0 = observed_quantile_1_cont, col = "red", lwd = 2)
```


## Computing residuals by simulation `r .emo("nerd")`

We can compute them all at once using the same process:
```{r simu resid, fig.height = 4, fig.width = 4}
simulated_residuals <- rep(NA, nrow(fit_poiss$model))
for (i in 1:nrow(fit_poiss$model)) {
  ecdf_fn <- ecdf(unlist(simulated_data_cont[i, ]))
  simulated_residuals[i] <- ecdf_fn(fit_poiss$y[i] + runif(1, min = -0.5, max = 0.5))
}
simulated_residuals
```

## Computing residuals by simulation `r .emo("nerd")`

By construction, these residuals should be uniformly distributed.

<div class="columns-2">

```{r plot sim res, fig.height = 4, fig.width = 4}
plot(simulated_residuals)
```

```{r plot ecdf sim res, fig.height = 4, fig.width = 4}
plot(ecdf(simulated_residuals))
```

</div>


## Computing residuals by simulation `r .emo("practice")`

All of this can be done easily with the package **{DHARMa}**:
```{r simres poiss, fig.width = 7, fig.height = 4, message=FALSE}
library(DHARMa)
fit_poiss_simres <- simulateResiduals(fit_poiss)
plot(fit_poiss_simres)
```

<!--
## Simulating residuals by parametric bootstrap

A variant based on the parametric bootstrap takes more time to run (and is more fragile to convergence issues), but that is needed for some tests.

In this case, the residuals (instead of observations) are compared to those produced by models refitted on the simulated response (instead of being directly compared to the simulated response).

```{r simres UK param boot, fig.width = 7, fig.height = 4}
fit_poiss_simres_boot <- simulateResiduals(fit_poiss, refit = TRUE, n = 1000)
```
-->

# Assumptions

## The main assumptions `r .emo("info")`

### Model structure

* linearity (for the linear predictor)
* lack of perfect multicollinearity
* predictor variables have fixed values

### Errors

* independence (no serial autocorrelation)
* lack of overdispersion and underdispersion

# Assumptions about the model structure

## How to test for linearity? `r .emo("info")`

It is difficult, but we may try to:

* plot simulated residuals
* plot *partial* residuals (a special type of residuals associated to each predictor)
* plot the predictions from a GAM model


## Example of non-linearity `r .emo("alien")`

```{r Aliens}
set.seed(1L)
Aliens <- data.frame(humans_eaten = runif(100, min = 0, max = 15))
Aliens$eggs <- rpois(n  = 100, lambda = exp(1 + 0.2 * Aliens$humans_eaten - 0.02 *  Aliens$humans_eaten^2))
fit_poiss_bad <- glm(eggs ~ humans_eaten, data = Aliens, family = poisson()) ## mispecified model
(fit_poiss_good <- glm(eggs ~ poly(humans_eaten, 2, raw = TRUE), data = Aliens, family = poisson())) ## good model
```

<!--
## Example of non-linearity
```{r plot Aliens,  fig.width = 4, fig.height = 4}
plot(I(1 + 0.2 * Aliens$humans_eaten - 0.02 *  Aliens$humans_eaten^2) ~ Aliens$humans_eaten,
     ylab = "eta", ylim = c(-1, 2))
data.for.pred <- data.frame(humans_eaten = 0:15)
points(predict(fit_poiss_good, newdata = data.for.pred) ~ I(0:15), col = "blue")
points(predict(fit_poiss_bad, newdata = data.for.pred) ~ I(0:15), col = "red")
```


## Example of non-linearity
```{r plot Aliens 2,  fig.width = 4, fig.height = 4}
plot(exp(1 + 0.2 * Aliens$humans_eaten - 0.02 *  Aliens$humans_eaten^2) ~ Aliens$humans_eaten,
     ylab = "predicted number of eggs", ylim = c(0, 6))
points(predict(fit_poiss_good, newdata = data.for.pred, type = "response") ~ I(0:15), col = "blue")
points(predict(fit_poiss_bad, newdata = data.for.pred, type = "response") ~ I(0:15), col = "red")
```
-->


## Example of non-linearity `r .emo("practice")`

```{r plot res Aliens, fig.width = 7, fig.height = 4}
plot(s_bad <- simulateResiduals(fit_poiss_bad, n = 1000)) ## simulated residuals from the bad fit
```


## Example of non-linearity `r .emo("practice")`

```{r plot res Aliens 2, fig.width = 7, fig.height = 4}
plot(s_good <- simulateResiduals(fit_poiss_good, n = 1000)) ## simulated residuals from the good fit
```


## Example of non-linearity `r .emo("practice")`

```{r plot partial Aliens}
car::crPlots(fit_poiss_bad, terms = ~ humans_eaten) ## partial residuals from the bad fit
```

## Example of non-linearity `r .emo("practice")`

```{r plot partial Aliens 2}
car::crPlots(fit_poiss_good, terms = ~ poly(humans_eaten, 2, raw = TRUE)) ## partial residuals from the good fit
```

## Example of non-linearity `r .emo("practice")`

```{r GAM, fig.width = 4, fig.height = 4, message = FALSE}
library(mgcv) ## package to fit GAMs
fit_poissGAM <- gam(eggs ~ s(humans_eaten), data = Aliens, family = poisson())
plot(fit_poissGAM, shade = TRUE, residuals = FALSE, shade.col = "red")  ## on the scale of the linear predictor!
```


## Multicollinearity and fixed-values `r .emo("recap")`

### Same as for LM!


# Assumptions about the errors

## How to test for independence? `r .emo("practice")`

You can run the Durbin Watson test on the simulated residuals using **{DHARMa}**:

```{r autocorr}
testTemporalAutocorrelation(s_good, time = fit_poiss_good$fitted.values, plot = FALSE)
```


## How to test for independence? `r .emo("practice")`

You can run the Durbin Watson test on the simulated residuals using **{DHARMa}**:

```{r autocorr_2}
testTemporalAutocorrelation(s_good, time = Aliens$humans_eaten, plot = FALSE)
```

<br>

Note: As for LM it is good practice to test for the lack of serial autocorrelation along all your predictors and not just along the fitted values.

<!--
## A particular legitimate case of dependence

Survival times series that are discrete and complete can be analysed using a binomial (binary!) GLM.

Example: the study of the influence of rainfall on mortality (here A dies at 5 yrs old and B at 3 yrs old).

```{r eles}
data.frame(age = c(1:5, 1:5),
      id = c(rep("A", 5), rep("B", 5)),
      death = c(0, 0, 0, 0, 1, 0, 0, 1, NA, NA),
      annual_rain = c(100, 120, 310, 50, 200, 45, 100, 320, 100, 120))
```

In such a case, a random effect for the identity of the individual is not needed! So there is no need for mixed model if individuals are all equally independent from each other.
-->


## Overdispersion / Underdispersion `r .emo("info")`

A GLM assumes a particular relationship between mu and var(mu):

```{r varmu, fig.width = 10, fig.height = 3}
p <- seq(0, 1, 0.1)
lambda <- 0:10
theta <- 0:10
v_b <- binomial()$variance(p)
v_p <- poisson()$variance(lambda)
v_G <- Gamma()$variance(theta)
par(mfrow = c(1, 3), las = 2)
plot(v_b ~ p, type = "o"); plot(v_p ~ lambda, type = "o"); plot(v_G ~ theta, type = "o")
```


## Overdispersion / Underdispersion `r .emo("info")`

### Overdispersion = more variance than expected

* very common $\rightarrow$ increase in false positives
* specially relevant for Poisson and Binomial
* irrelevant for the binary case! (don't look for it)

Usual suspects:

* lack of linearity
* unobserved heterogeneity
* zero-augmentation

<br>

### Underdispersion = less variance than expected

* rather rare  $\rightarrow$ increase in false negatives


## Overdispersion / Underdispersion `r .emo("alien")`

### A toy example

```{r overdisp}
set.seed(1L)
popA <- data.frame(humans_eaten = runif(50, min = 0, max = 15))
popA$eggs <- rpois(n = 50, lambda = exp(-1 + 0.05 * popA$humans_eaten))
popA$pop <- "A"
popB <- data.frame(humans_eaten = runif(50, min = 0, max = 15))
popB$eggs <- rpois(n = 50, lambda = exp(-3 + 0.4 * popB$humans_eaten))
popB$pop <- "B"
AliensMix <- rbind(popA, popB)
(fit_poissMix <- glm(eggs ~ humans_eaten, family = poisson(), data = AliensMix))
```


## Overdispersion / Underdispersion `r .emo("info")`

Two measures are traditionally used to quantify dispersion:

```{r deviance overdisp}
fit_poissMix$deviance / fit_poissMix$df.residual
sum(residuals(fit_poissMix, type = "pearson")^2) / fit_poissMix$df.residual
```


## Overdispersion / Underdispersion `r .emo("info")`

Asymptotic tests based on these measures do not always behave well `r .emo("proof")`

<div class = "columns-2">

```{r gof 3, fig.width = 3, fig.height = 3}
p <- replicate(1000, {
  d <- simulate_Aliens_GLM()
  m <- update(fit_binar, data = d)
  X <- sum(residuals(m, type = "pearson")^2)
  pchisq(X, m$df.residual, lower.tail = FALSE)
})
plot(ecdf(p))
abline(0, 1, lty = 2, col = "green")
```

```{r gof 4, fig.width = 3, fig.height = 3}
p <- replicate(1000, {
  d <- simulate_Aliens_GLM()
  m <- update(fit_binar, data = d)
  dev <- sum(residuals(m, type = "deviance")^2)
  pchisq(dev, m$df.residual, lower.tail = FALSE)
})
plot(ecdf(p))
abline(0, 1, lty = 2, col = "green")
```

</div>

Note: this is far from the uniform distribution we aim at...


## Overdispersion / Underdispersion `r .emo("practice")`

We can again use **{DHARMa}** for that:

```{r sim overdisp}
r <- simulateResiduals(fit_poissMix, n = 1000)
testDispersion(r) # stat = squared pearson residuals compared to that obtained by simulation
```


# Solving dispersion problems

## Potential solutions `r .emo("practice")`

* fix linearity issues
* fix heterogeneity issues (if you have the data)

```{r inter}
fit_poissMix2 <- glm(eggs ~ pop*humans_eaten, family = poisson(), data = AliensMix)
r2 <- simulateResiduals(fit_poissMix2, n = 1000)
testDispersion(r2, plot = FALSE)  ## you can change options to test for underdispersion
```

## Potential solutions `r .emo("info")`

* fix linearity issues
* fix heterogeneity issues (if you have the data)
* model the overdispersion
* try another probability distribution
* there are specific solutions if the origin is zero-augmentation


## Modeling simple overdispersion

### For Poisson

Variance = $k \times \mu$

```{r quasipoiss}
fit_poissMixQ <- glm(eggs ~ humans_eaten, family = quasipoisson(), data = AliensMix)
summary(fit_poissMixQ)$coef
car::Anova(fit_poissMixQ, test = "F")  ## 'F' not 'LR' because dispersion is estimated!
```


## Simple modeling of overdispersion `r .emo("practice")`

### For Poisson

Variance = $k \times \mu$

It kind of works (if the overdispersion is not too strong), but we cannot do everything with these types of fit since they have no likelihood:
```{r confint quasipoiss}
confint(fit_poissMixQ) ## works (using deviance)
c(logLik(fit_poissMixQ), AIC(fit_poissMixQ)) ## does not work
```


## Simple modeling of overdispersion `r .emo("info")`

### For binomial

* irrelevant for the binary case!
* there is also ```quasibinomial()```, with the same limit (no likelihood)
* can happen in the general case

<br>

### Solution

* add a random effect with one level per individual (see GLMMs)


## Modelling departure from expected dispersion with other probability distributions `r .emo("info")`

For count data, you have 3 options:

* Poisson ($V(\mu) = \mu$)
* Negative binomial ($V(\mu) = \mu + \frac{\mu^2}{\theta}$, if $\theta = 1$ this is the geometric model)
* Conway-Maxwell-Poisson ($V(\mu) =$ something complex)

Note: only the Conway-Maxwell-Poisson can handle underdispersion.


## Negative binomial with **{spaMM}** `r .emo("practice")`

```{r negbin spaMM}
fit_poissMix_spaMM   <- fitme(eggs ~ humans_eaten, family = poisson(), data = AliensMix)
fit_poissMixNB_spaMM <- fitme(eggs ~ humans_eaten, family = spaMM::negbin(), data = AliensMix) ## namespace needed!
summary(fit_poissMixNB_spaMM)
c(AIC(fit_poissMix_spaMM), AIC(fit_poissMixNB_spaMM))
```

Note: the negative binomial fit is much better than the Poisson one.

<!--
## Conway-Maxwell-Poisson (from **{ spaMM}**)

### It can fit both over- and under- dispersion!

* overdispersion: nu < 1 (nu = 0 corresponds to the geometric distribution)
* expected dispersion (i.e. Poisson): nu = 1
* underdispersion: nu > 1


## Conway-Maxwell-Poisson

### It can fit both over- and under- dispersion!

```{r COMPoisson2, warning = FALSE, eval=FALSE}
fit_poissMixCP_spaMM <- fitme(eggs ~ humans_eaten, family = COMPoisson(), data = AliensMix)
summary(fit_poissMixCP_spaMM)
c(AIC(fit_poissMix_spaMM), AIC(fit_poissMixNB_spaMM), AIC(fit_poissMixCP), AIC(fit_poissMixCP_spaMM))
```

Note: here we estimate nu, so it can take (much) longer time to fit!


## Comparison

```{r comp COMPoiss, echo = FALSE, warning = FALSE, fig.height = 6, fig.width = 6}
d <- data.frame(humans_eaten = seq(0, 15, length = 1000))
p <- predict(fit_poissMix_spaMM, newdata = d, intervals = "fixefVar")
plot(p ~ d$humans_eaten, ylim = range(fit_poissMix_spaMM$data$eggs), type = "l", lwd = 3)
points(attr(p, "interval")[, 1] ~ d$humans_eaten, lty = 2, type = "l")
points(attr(p, "interval")[, 2] ~ d$humans_eaten, lty = 2, type = "l")
p <- predict(fit_poissMixNB_spaMM, newdata = d, intervals = "fixefVar")
points(p ~ d$humans_eaten, type = "l", col = "red", lwd = 3)
points(attr(p, "interval")[, 1] ~ d$humans_eaten, lty = 2, type = "l", col = "red")
points(attr(p, "interval")[, 2] ~ d$humans_eaten, lty = 2, type = "l", col = "red")
p <- predict(fit_poissMixCP_spaMM, newdata = d, intervals = "fixefVar")
points(p ~ d$humans_eaten, type = "l", col = "blue", lwd = 3)
points(attr(p, "interval")[, 1] ~ d$humans_eaten, lty = 2, type = "l", col = "blue")
points(attr(p, "interval")[, 2] ~ d$humans_eaten, lty = 2, type = "l", col = "blue")
legend("topleft", fill = c("black", "red", "blue"), legend = c("Poisson", "Negative Binomial", "Conway-Maxwell-Poisson "), bty = "n")
```
-->


# Zero-augmentation

## Zero-augmentation in brief `r .emo("info")`

### What is it?

It occurs for any event when there are too many zeros.

### Why does it occur?

It can occur when the response results from a 2 (or more) steps process

Examples:

* detection issue (low counts are less detected, e.g. counting cells on microscope)
* biological (e.g. infection, then spread of microbes)


## How to tackle zero-augmentation? `r .emo("info")`

### We have 2 main options for this:

Fit an **hurdle model**

* e.g. binomial + truncated Poisson or truncated negative binomial
* a single source of zeros (modelled by binomial)
* e.g. of response variable: number of offspring

<br>

Fit a **zero-inflation model**

* e.g. binomial + Poisson or negative binomial
* two sources of zeros (modelled by binomial AND Poisson/negative binomial)
* e.g. of response variable: number of viruses in individuals (0 for unexposed, 0 for exposed with strong immune system)

Note: if you are not sure where the zeros come from, try both!


## Generating zero-augmented data `r .emo("alien")`

```{r zeroinfl, fig.height=4}
set.seed(1L)
AliensZ <- simulate_Aliens_GLM(N = 1000)
AliensZ$eggs <- AliensZ$eggs * AliensZ$happy ## unhappy Aliens loose their eggs :-(
barplot(table(AliensZ$eggs))
```


## Poisson fit of zero-augmented data `r .emo("practice")`

```{r zeroinfl poiss}
fit_Zpoiss <- glm(eggs ~ humans_eaten, data = AliensZ, family = poisson())
r <- simulateResiduals(fit_Zpoiss, n = 1000)
testZeroInflation(r, plot = FALSE)
```
Note: we reject the null hypothesis which is that the observed number of zeros is compatible with the one expected if those data would have be produced by a Poisson model.

```{r zeroinfl poiss2}
mean(sum(AliensZ$eggs == 0) / sum(dpois(0, fitted(fit_Zpoiss)))) ## = ratioObsSim
```

## Poisson fit of zero-augmented data `r .emo("practice")`

```{r test zeroinfl, fig.width=7}
testZeroInflation(r, plot = TRUE) ## plot the number of 0s simulated under the model and observed
```

<!--
## Negative binomial fit of zero-augmented data

```{r test zeroinfl negbin}
fit_Znb <- glm.nb(eggs ~ humans_eaten, data = AliensZ)
r <- simulateResiduals(fit_Znb, refit = TRUE, n = 1000)
testZeroInflation(r, plot = FALSE)
```


## Negative binomial fit of zero-augmented data

```{r plot test zeroinfl}
testZeroInflation(r, plot = TRUE)
```
-->

## Hurdle fit of zero-augmented data `r .emo("practice")`

You can fit hurdle models with `pscl::hurdle()`:
```{r hurdle, message = FALSE}
library(pscl)
fit_Zhurd1 <- hurdle(eggs ~ humans_eaten | humans_eaten, dist = "poisson", zero.dist = "binomial",
                     data = AliensZ)
fit_Zhurd2 <- hurdle(eggs ~ humans_eaten | 1, dist = "poisson", zero.dist = "binomial",
                     data = AliensZ)
lmtest::lrtest(fit_Zhurd1, fit_Zhurd2)
```

Note: here we tested whether `humans_eaten` influenced the production of all 0s (since they all come from the binary process).

## Zero-inflation fit of zero-augmented data `r .emo("practice")`

You can fit zero-inflation models with `pscl::zeroinfl()`:
```{r zeroinfl fit, message = FALSE}
fit_Zzi1 <- zeroinfl(eggs ~ humans_eaten | humans_eaten, dist = "poisson", data = AliensZ)
fit_Zzi2 <- zeroinfl(eggs ~ humans_eaten | 1, dist = "poisson", data = AliensZ)
lmtest::lrtest(fit_Zzi1, fit_Zzi2)
```

Note: here we tested whether `humans_eaten` influenced the production of some of the 0s (since they come both from the binary process and the Poisson process).



## Comparison of alternative fits `r .emo("practice")`

```{r comp zeroinfl}
cbind(Poisson = AIC(fit_Zpoiss), ## poor fit
      Hurdle  = AIC(fit_Zhurd1), ## close to best
      ZeroInf = AIC(fit_Zzi1))   ## best but not by much
```

Let's also compare the model parameter estimates:
```{r comp zeroinfl2}
tab <- rbind(Poisson = c(fit_Zpoiss$coefficients, NA, NA),
             Hurdle = unlist(fit_Zhurd1$coefficients),  ## NOTE: the hurdle part predicts positive counts, not zeros!!
             ZeroInfl = unlist(fit_Zzi1$coefficients),  ## NOTE: the binary part predicts zeros, not counts!!
             Truth = c(attr(AliensZ, "param.eta")$eggs, attr(AliensZ, "param.eta")$happy))
colnames(tab) <- c("Int.count", "Slope.count", "Int.bin", "Slope.bin")
tab
```

Note: parameter estimates confirm that both the hurdle and the zero-inflation fits are (here) accurate.


# One last trap: separation

## The problem of separation in Binomial `r .emo("info")`

### What is it?

Separation occurs when a level (or combination of levels) for categorical predictor, or a particular threshold along a continuous predictor, predicts the outcomes perfectly.

### Complete or quasi separation?

From Wikipedia:

*For example, if the predictor X is continuous, and the outcome y = 1 for all observed x > 2. If the outcome values are perfectly determined by the predictor (e.g., y = 0 when x  2) then the condition "complete separation" is said to occur. If instead there is some overlap (e.g., y = 0 when x < 2, but y has observed values of 0 and 1 when x = 2) then "quasi-complete separation" occurs. A 2  2 table with an empty cell is an example of quasi-complete separation.*

### Consequences

* sometimes the model cannot be fitted
* when it does fit, the estimates and standard errors are usually wrong!


## The problem of separation in Binomial `r .emo("proof")`

```{r separ}
set.seed(1L)
n <- 50
test <- data.frame(happy = rbinom(2*n, prob = c(rep(0, n), rep(0.75, n)), size = 1), 
                   sp = factor(c(rep("sp1", n), rep("sp2", n))))
table(test$happy, test$sp)
```

```{r fit separ}
fit_separated <- glm(happy ~ sp, data = test, family = binomial())
summary(fit_separated)$coefficients
```

Note: despite the very clear effect, the test does not detect it.

## The problem of separation in Binomial `r .emo("practice")`

### Detection for `glm()` fits

```{r safebin, error = TRUE, message = FALSE, warning = FALSE}
library(safeBinaryRegression)  ## overload the glm function
fit_separated <- glm(happy ~ sp, data = test, family = binomial()) ## don't mind warnings here
detach(package:safeBinaryRegression) ## stop masking original glm()
```

## The problem of separation in Binomial `r .emo("practice")`

### Detection for `fitme()` fits

```{r spaMM separ, error = TRUE}
fit_separated_spaMM <- fitme(happy ~ sp, data = test, family = binomial())
```

Note: you may need to install an additional package, but **{spaMM}** will inform you about it.


## The problem of separation in Binomial `r .emo("practice")`

### Let's tweak the data in a conservative way
```{r tweak separ}
test$happy[1] <- 1
table(test$happy, test$sp)
fit_separated2 <- glm(happy ~ sp, data = test, family = binomial())
summary(fit_separated2)$coefficients
```

Note: this little tweak of the data is sufficient to recover significance.


## The problem of separation in Binomial `r .emo("practice")`

A better solution for the binary case (as well as for binomial in general) is to use a specific fitting method:

```{r brglm2, message = FALSE}
test$happy[1] <- 0  ## restore the original data
library(brglm2)
fit_separated3 <- glm(happy ~ sp, data = test, family = binomial(), method = "brglmFit")
summary(fit_separated3)$coefficients
```


## What you need to remember `r .emo("goal")`

* that GLM can be fitted on different types of responses
* how to convert values between the response and linear predictor scales
* how to interpret parameter estimates from GLM Poisson and Binomial
* that LR tests are best for family with known dispersion
* that intervals based on likelihood are best for family with known dispersion
* that tests by parametric bootstrap are best and easy to do with **{spaMM}**
* that most assumptions behind GLM are similar to those for LM (but harder to test)
* that General Additive Models (GAM) can be useful to improve linearity
* how to diagnose and tackle overdispersion, zero-augmentation and separation



# Table of contents

## The Generalized Linear Model: GLM

* 3.0 [Introduction](./GLM_intro_course.html)
* 3.1 [Let's practice](./GLM_practice_exercises.html)

<br>

<div align="right">
[Back to main menu](./Title.html#2)
</div>
