---
title: "Introduction"
author: "Alexandre Courtiol"
date: "`r Sys.Date()`"
output:
  ioslides_presentation:
    widescreen: true
    smaller: true
vignette: >
  %\VignetteIndexEntry{1 Introduction}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}

---
# General information

## What do you need?
* a computer with internet connection
* a web browser (up-to-date)
* R version $\geq$ 3.3
* RStudio version $\geq$ 1.0 (recommended)

## How to access the course?

The course is an **R package** called **LM2GLMM**.

### Installation

identify the correct `path`, then type:

```{r install_package, eval = FALSE}
install.packages(pkgs = "path/LM2GLMM")
```
<br>

### Loading

```{r load_package, eval = FALSE}
library(package = "LM2GLMM")
```


## You should get that:

```
 #############################################
 #                                           #
 #     This is the package for the course    #
 #                                           #
 #    Advanced Statistical Applications:     #
 #          from LM to GLMM using R          #
 #                                           #
 #   You have the version X.X.X installed!   #
 #                                           #
 # Type browseVignettes(package = 'LM2GLMM') #
 #    to access the slides of the course     #
 #                                           #
 #############################################
```

### !!! Always check the version number !!!

# Refreshing your R skills

## Vectors

```{r vectors}
foo <- c(1, 5, 10, 23)
foo[2]
foo2 <- c(a = 1, b = 5, c = 10, d = 23)
foo2[2]
foo2["b"]
foo2["b"][[1]]
```

## Vectors

```{r vectors2}
foo3 <- c(foo, as.numeric(foo2))
foo3
foo3[foo3 > 3]
summary(foo3)
table(foo3)
```

## Vectors

```{r vectors3}
class(foo3)
(foo4 <- names(foo2)) ## Tip: the brackets are used for immediate display
class(foo4)
foo5 <- factor(foo4)
class(foo5)
```

