---
title: "Introduction"
subtitle: "exercises"
author: "Alexandre Courtiol"
date: "`r Sys.Date()`"
output:
  ioslides_presentation:
    widescreen: true
    smaller: true
vignette: >
  %\VignetteIndexEntry{1.0 z-------------------------------------------------------------exercises}
  %\VignetteEncoding{UTF-8}
  %\VignetteEngine{knitr::rmarkdown}
---

```{r setup, include=FALSE}
library(LM2GLMM)
knitr::opts_chunk$set(cache = FALSE,
                      fig.align = "center",
                      fig.width = 4,
                      fig.height = 4,
                      cache.path = "./cache_knitr/LM_intro_ex/",
                      fig.path = "./fig_knitr/LM_intro_ex/")
options(width = 200)
set.seed(1L)
```

## Computing with R

Compute the following equation:

<br>

<center><font size="8"> $\sqrt{\frac{2^{3+1}}{\frac{4}{5\times{6}}}-20}$ </font></center>

<br>

<div align="right">
[Back to main menu](./Title.html#2)
</div>


## Data exploration

Using the dataset `TitanicSurvival`, figure out: 

- how many rows and columns are they in this dataset?
- how many columns correspond to factors?
- how many females below 20 years old survived in 2nd class, and how many did not?

```{r echo = FALSE, results='hide'}
dim(TitanicSurvival)
str(TitanicSurvival)
apply(TitanicSurvival, 2, function(x) any(is.na(x)))
focus <- with(TitanicSurvival, survived[sex == "female" & !is.na(age) & age < 20 & passengerClass == "2nd"])
table(focus, useNA = "always")
```


## An example of programming

The following function can be used to investigate the statistical power of a t-test (i.e. the probability to detect a significant effect when they really is one = rate of true positive):

```{r example, fig.show='hide'}
compare_heights <- function(n_group = 10, height_difference = 5) {
  male   <- rnorm(n = n_group, mean = 180, sd = 6)
  female <- rnorm(n = n_group, mean = 180 - height_difference, sd = 6)
  t_test_res <- t.test(male, female)
  t_test_res$p.value
}

N <- seq(from = 10, to = 60, by = 2)
power <- sapply(N, function(n) mean(replicate(100, compare_heights(n_group = n)) <= 0.05))
plot(power ~ N)
```

<br>

Can you run, read and understand this code?


## Plots

Using the dataset `TitanicSurvival`, find a good way to show at once what influenced the survival of passengers.


##

<div align="right">
[Back to main menu](./Title.html#2)
</div>
