---
title: "Introduction"
subtitle: "Solutions"
author: "Alexandre Courtiol"
date: "`r Sys.Date()`"
output:
  html_vignette:
    toc: true
vignette: >
  %\VignetteEncoding{UTF-8}
  %\VignetteIndexEntry{1.0 z-------------------------------------------------------------solutions}
  %\VignetteEngine{knitr::rmarkdown}
editor_options:
  chunk_output_type: console
---
```{r setup, include=FALSE}
library(LM2GLMM)
knitr::opts_chunk$set(cache = TRUE, fig.align = "center", fig.width = 6, fig.height = 6,
                      cache.path = "./cache_knitr/Exo_Intro_solution/", fig.path = "./fig_knitr/Exo_Intro_solution/")
options(width = 90)
```

## Computing with R

Compute the following equation:

<br>

<center><font size="8"> $\sqrt{\frac{2^{3+1}}{\frac{4}{5\times{6}}}-20}$ </font></center>

<br>

The idea is to do things step by step and always check that things are OK before they get too complex:

```{r equation}
3+1
(3+1)
2^(3+1)
(2^(3+1)) / 4
(2^(3+1)) / (4 / (5*6))
((2^(3+1)) / (4 / (5*6))) - 20
sqrt(((2^(3+1)) / (4 / (5*6))) - 20)
```


## Data exploration

Using the dataset `TitanicSurvival`, figure out: 

- how many rows and columns are they in this dataset?


```{r}
dim(TitanicSurvival)
```
We see that there are `r nrow(TitanicSurvival)` rows and `r ncol(TitanicSurvival)` columns.

<br>

- how many columns correspond to factors?

```{r}
str(TitanicSurvival)
```
We see that there are 3 factors.

<br>

- how many females below 20 years old survived in 2nd class, and how many did not?

```{r}
test <- TitanicSurvival$survived == "yes" & 
        TitanicSurvival$sex == "female" &
        TitanicSurvival$age < 20 &
        TitanicSurvival$passengerClass == "2nd"
sum(test, na.rm = TRUE)
```
Many other ways are possible, such as:

```{r, message=FALSE}
library(dplyr)
TitanicSurvival |>
  filter(survived == "yes", sex == "female", age < 20, passengerClass == "2nd") |>
  count()
```



## An example of programming

The following function can be used to investigate the statistical power of a t-test (i.e. the probability to detect a significant effect when they really is one = rate of true positive):

```{r example, fig.show='hide'}
compare_heights <- function(n_group = 10, height_difference = 5) {
  male   <- rnorm(n = n_group, mean = 180, sd = 6)
  female <- rnorm(n = n_group, mean = 180 - height_difference, sd = 6)
  t_test_res <- t.test(male, female)
  t_test_res$p.value
}

N <- seq(from = 10, to = 60, by = 2)
power <- sapply(N, function(n) mean(replicate(100, compare_heights(n_group = n)) <= 0.05))
plot(power ~ N)
```

<br>

Can you run, read and understand this code?

I cannot really correct that one here. Ask if you don't get it.
