---
title: "Introduction"
author: "Alexandre Courtiol"
date: "`r Sys.Date()`"
output:
  ioslides_presentation:
    widescreen: true
    smaller: true
vignette: >
  %\VignetteIndexEntry{1 Introduction}
  %\VignetteEncoding{UTF-8}
  %\VignetteEngine{knitr::rmarkdown}
editor_options: 
  chunk_output_type: console
---
```{r setup, include=FALSE}
library(LM2GLMM)
set.seed(1L)
knitr::opts_chunk$set(fig.align = "center")
```

## Table of content

### 1. [General information](./Introduction.html#3)
### 2. [The exam](./Introduction.html#7)
### 3. [A catalogue of most useful {base} R commands for this course](./Introduction.html#10)

<div align="right">
[Back to main menu](./Title.html#2)
</div>

# General information

## Course philosophy

### What I want you to become:
* autonomous (capable of doing a little, capable of learning easily)
* capable of tackling analyses of real dataset (and not artificial problem free dataset)

### How to get there?
* I will get you to experiment things in your R console as a mean to understanding and learning
* I will put a particular emphasis on programming -- not on mathematics

### What I will not cover in this course:
* how to use the {dplyr} and friends ({tidyr}, {tidyselect}, {forcats}...) to put data in shape
* which packages can do X or Y in one single click



## What do you need for the course?
* a laptop
* a computer with internet connection
* a web browser (up-to-date)
* R version $\geq$ 3.6
* RStudio version $\geq$ 1.2 (recommended)

## How to access the course?

The course is an **R package** called **LM2GLMM**.

To download and install this package, visit: https://github.com/courtiol/LM2GLMM/

To load the package, do:

```{r load_package, eval = FALSE}
library(package = "LM2GLMM") ## !!! Always check the version with me because I will update it often !!!
```

```
 #############################################
 #                                           #
 #    This is the package for the course     #
 #                                           #
 #    Advanced Statistical Applications:     #
 #          from LM to GLMM using R          #
 #                                           #
 #       Version XX.XX.XX.X installed!       #
 #                                           #
 #    To access the slides, type either      #
 #   browseVignettes(package = 'LM2GLMM')    #
 #                    or                     #
 #             get_vignettes()               #
 #                                           #
 #############################################
```

# The exam

## Procedure

### What?

* biological questions for which the answers will require you to build liner models
* you will answer by writing text and R code producing results including figures and/or tables 

### When?

* half a day suiting everyone, once the course will be over
* duration for the examination: **one hour**

(It should be sufficient, but you will be able to stay longer if you need to!)


### Condition?

* individually
* you are free to look at your course and R-help files (no internet)


## Grading

### I will grade:

* the correctness and clarity of the R code
* the correctness and clarity of the answers

### I will give bonuses for the quality/elegance of:

* the writing
* the code
* the illustrations

### I will help you:

* but only if you really are stuck


# A catalogue of most useful {base} R commands for this course

## Vectors: basics

```{r vectors basics}
foo <- c(1, 5, 5, 10, 23, NA, NA)
foo[2]
foo > 3
foo[foo > 3]
foo2 <- c(a = 1, b = 5, c = 10, d = 23)
foo2["b"]
```

## Vectors: some useful functions

```{r vectors function}
length(foo)
summary(foo)
any(is.na(foo))
table(foo, useNA = "always")
```

## Vectors: classes

```{r vectors classes}
foo1 <- c(1, 5, 5, NA)
foo2 <- c(a = 1, b = 3)
foo3 <- c(TRUE, FALSE)
foo4 <- c("a", "b", "c")
foo5 <- factor(foo3)
foo6 <- c(1L, 3L, 5L)
```

```{r vectors class}
class(foo1)
```

```{r vectors class table, echo = FALSE}
classes <- unlist(lapply(as.list(names <- paste0("foo", 1:6)), function(x) class(get(x))))
knitr::kable(rbind(names, classes))
```

## Vectors: factors

```{r vetors factors}
(foo <-  factor(c("a", "b", "c"))) ## Tip: the brackets force the display
```

```{r vetors factors2, results='hide'}
relevel(foo, ref = "c")                      ## change first level
factor(foo, levels = c("c", "b", "a"))       ## control order of levels
droplevels(foo[foo != "b"])                  ## drop empty levels
c(factor(c("a", "b")), factor(c("b", "c")))  ## do not coerce factors with different levels!
factor(c("a", "b"), c("b", "c"))             ## you can coerce characters
as.character(foo)                            ## you can turn factors into characters
```


## Attributes

```{r attributes}
foo <- c(a = 1, b = 5, c = 10, d = 23)
attributes(foo)
attr(foo, "names")  ## Tip: here a shortcut would be 'names(foo)', but this is not general
foo <-  factor(c("a", "b", "c"))
attributes(foo)
```

## Dataframes: basics

```{r dataframe}
foo <- data.frame(
  x = c(1, 3, 5),
  z = factor(c("a", "b", "c"))
  )
foo
dim(foo) ## Tip: try nrow() and ncol()
dimnames(foo) ## Tip: try rownames() and colnames()
```

## Dataframes: indexing

```{r dataframe indexing}
foo[2, ]
foo[, 2]
foo$x
foo[, "x"]
```

## Dataframes: combining rows

```{r dataframe foo2}
foo2 <- data.frame(
  x = c(2, 4),
  z = factor(c("b", "d"))
  )
```

```{r dataframe rbind}
rbind(foo, foo2) ## Note that the factor levels are redefined!
```

```{r vector joining factor levels}
c(foo$z, foo2$z) ## but this is not the case for vectors!!
```

## Dataframes: combining columns

```{r dataframe bar}
bar <- data.frame(
  x = c(1, 2, 3),
  w = c(2, 4, 3),
  s = factor(c("b", "d", "e"))
  )
```

```{r dataframe cbind}
cbind(foo, bar) ## Note that columns are not merged
```


## Dataframes: {dplyr} <---> {base} translation

```{r dataframe dplyr, message=FALSE}
library(dplyr)
```

```{r dataframe tibble}
foo_tbl <- tibble(x = c(1, 3, 5),
                  z = factor(c("a", "b", "c")))
foo_tbl ## the display is different
```

```{r dataframe dplyr select columns, results='hide'}
select(foo_tbl, x)          ## same as foo_tbl[, "x", drop = FALSE]
slice(foo_tbl, 2)           ## same as foo_tbl[2, ]
filter(foo_tbl, x == 3)     ## same as foo_tbl[foo_tbl$x == 3, ]
```


```{r dataframe dplyr select columns piped, results='hide'}
foo_tbl %>% select(x)       ## same as foo_tbl[, "x", drop = FALSE]
foo_tbl %>% slice(2)        ## same as foo_tbl[2, ]
foo_tbl %>% filter(x == 3)  ## same as foo_tbl[foo_tbl$x == 3, ]
```


## Matrices: basics

```{r matrix}
foo <- matrix(data = 1:4, nrow = 2, ncol = 2) ## Tip: try with byrow = TRUE
colnames(foo) <- c("a", "b"); rownames(foo) <- c("A", "B")
foo
```
Indexing similar to dataframes (but you cannot use $):
<div  class="columns-2">
```{r matrix indexing}
foo[, 2]
foo[2, ]
```
</div>

```{r matrix indexing 2}
foo[1, 2]
```

## Matrices: transposition, inversion

```{r matrix transpose}
foo
t(foo)  ## transpose of foo
solve(foo)  ## inverse of foo
```

## Matrices: multiplication

```{r matrix multiplication}
foo %*% matrix(c(-1, 1))  ## matrix multiplication; same as foo %*% c(0, 1, -1)
foo %*% foo  ## matrix multiplication
foo * foo  ## NOT MATRIX MULTIPLICATION!
```

## Lists

```{r lists}
foo <- list("foo1" = c(1:10), "foo2" = factor(c("a", "b")))
foo
foo[["foo2"]]
foo$foo2
```

## Sequences

```{r seq}
seq(from = 1, to = 5, by = 0.5)
seq(from = 1, by = 0.5, length.out = 9)
seq(from = 1, to = 5, length.out = 9)
```


## Pseudo-random numbers in R

```{r rnorm}
rnorm(5, mean = 2, sd = 0.1)  ## Tip: check ?Distributions for other distributions
rnorm(5, mean = 2, sd = 0.1)
set.seed(14353L)
rnorm(5, mean = 2, sd = 0.1)
set.seed(14353L)
rnorm(5, mean = 2, sd = 0.1)
```


## Functions

```{r function}
addA_B <- function(a = 0, b = 0) {
  c <- a + b
  return(c)
  }
```

```{r function test}
addA_B(a = 5, b = 7)
addA_B(a = 5)
```

```{r function short}
addA_B_bis <- function(a, b) a + b
addA_B_bis(2, 3)
```


## Repeated operations

```{r hello}
hello <- function(who = "alex") paste("hello", who)
```

```{r replicate}
replicate(10, hello())  ## returns a vector when it can
```

```{r sapply}
sapply(c("alex", "olivia"), function(i) hello(who = i)) ## returns a vector when it can
```

```{r lapply}
lapply(c("alex", "olivia"), function(i) hello(who = i)) ## returns a list
```


## An example combining all of the above

### Investigating the statistical power of a t-test

```{r example, fig.show='hide'}
compare_heights <- function(n_group = 10, height_difference = 5) {
  male   <- rnorm(n = n_group, mean = 180, sd = 6)
  female <- rnorm(n = n_group, mean = 180 - height_difference, sd = 6)
  t_test_res <- t.test(male, female)
  t_test_res$p.value
}

N <- seq(from = 10, to = 60, by = 2)
power <- sapply(N, function(n) mean(replicate(100, compare_heights(n_group = n)) <= 0.05))
plot(power ~ N)
```

Can you understand that?


<div align="right">
[Back to main menu](./Title.html#2)
</div>
