---
title: "LM: fitting procedures"
author: "Alexandre Courtiol"
date: "`r Sys.Date()`"
output:
  ioslides_presentation:
    widescreen: true
    smaller: true
vignette: >
  %\VignetteIndexEntry{2.1 Fitting procedures}
  %\VignetteEncoding{UTF-8}
  %\VignetteEngine{knitr::rmarkdown}
editor_options: 
  chunk_output_type: console
---
```{r setup, include=FALSE}
library(LM2GLMM)
options(width = 110)
knitr::opts_chunk$set(error = TRUE,
                      cache = TRUE,
                      cache.path = "./cache_knitr/LM_fitting/",
                      fig.path = "./fig_knitr/LM_fitting/")
```

## The Linear Model: LM

* 2.0 [Introduction](./LM_intro.html)
* 2.1 [Fitting procedure](./LM_fitting.html)
* 2.2 [Tests & Intervals](./LM_test_intervals.html)
* 2.3 [Assumptions & Outliers](./LM_assumptions.html)
* 2.4 [Let's practice](./LM_practice.html)

<br>

<div align="right">
[Back to main menu](./Title.html#2)
</div>

## You will learn in this session

* that the variance of residuals is a biased estimator of the error variance
* how to extract key statistics for models fitted with `stats::lm()` & `spaMM::fitme()`
* how to compute estimates numerically in a few lines of code
* how ```lm()``` works
* what the components stored within ```lm()``` objects are

<!--
<div class="columns-2">

```{r alien data 2 eval, fig.align='center', fig.asp=1, fig.width=4, echo = FALSE, fig.cap="True relationship"}
#plot(size ~ humans_eaten, data = Alien, ylab = "Alien size (cm)", xlab = "No. of humans eaten")
#abline(a = 50, b = 1.5, col = "green", lwd = 2)
```

```{r alien data 3 eval, fig.align='center', fig.asp=1, fig.width=4, echo = FALSE, fig.cap="Inferred relationship"}
#plot(size ~ humans_eaten, data = Alien, ylab = "Alien size (cm)", xlab = "No. of humans eaten")
#text(x = nrow(Alien)/2, y = 82, labels = "?", cex = 5, col = "blue")
#abline(fit_alien_lm, col = "blue", lwd = 2, lty = 2)
```
</div>
-->

# High-school reminiscence

## Sample and population variance

Do you remember that?

### Population variance

<center><font size = 5>
$$\frac{\displaystyle\sum^n_{i = 1}(x_i - \bar{x})^2}{n}$$
</font></center>

### Sample variance

<center><font size = 5>
$$\frac{\displaystyle\sum^n_{i = 1}(x_i - \bar{x})^2}{n\color{red}{-1}}$$
</font></center>


## Sample and population variance

```{r vars}
set.seed(1)
bias_vars_list <- replicate(100000, {
  observations <- rnorm(n = 20, mean = 0, sd = sqrt(5))
  numerator <- sum((observations - mean(observations))^2)
  c(bias_pop_var   = (numerator/20) - 5,
    bias_sample_var = (numerator/19) - 5)
}, simplify = FALSE)
bias_vars_df <- data.frame(do.call("rbind", bias_vars_list)) ## to turn the output as a clean dataframe
```

```{r vars2}
apply(bias_vars_df, 2, mean)
```

The bias corresponds to an underestimation stemming from looking at the variation around the mean estimate: the variation around the mean estimate is always smaller that the variation around the true mean in the population. This is because by construction the estimate of the mean is the value generally the closest to all **sampled** points.

The bias can be removed by considering $n - p$ in the denominator to correct for the bias when you estimate $p$ parameters ($p = 1$, when there is only one mean). 



# Data

## Our toy dataset

Let's simulate the size of 12 aliens using the function integrated in this LM2GLMM package:

```{r Aliens}
set.seed(123)
(Alien <- simulate_Aliens()) ## by default, N = 12, intercept = 50, slope = 1.5 and sigma2 = 25
```


# Fitting LM with `stats::lm()` & `spaMM::fitme()`

## A note on `package::fn`

In R, when you call a function it looks for it in the global environment (i.e. where your object are stored) and in all the packages attached to the so-called search list:

```{r search list}
search()
```

The packages there are those that have been attached by the system or by you (e.g. using `library()`).

Using the syntax `package::fn()` allows for you to access a function from an installed package that has not been attached which can be useful if you don't want to load a package and save one line of code (e.g. as when we do `drat::addRepo("courtiol")`.

Importantly, it is also a way to be explicit about where a function is coming from.

So `stats::lm()` means that `lm()` comes from the build-in package **{stats}** and `spaMM::fitme()` means that `fitme()` comes from the optional package **{spaMM}**.


## A note on the R package **{spaMM}**

**{spaMM}** is an R package developed by François Rousset (a population geneticist from Montpellier, France).

<br>

We will use it a lot because it is very flexible and powerful:

- it can fit from LM to GLMM (and more, e.g. DHGLM)
- it can handle temporal and spatial autocorrelation
- it allows for a straightforward computation of intervals, parametric bootstraps, and more


## Fitting the Alien data using ```lm()```

```{r fit alien lm}
fit_alien_lm <- lm(size ~ humans_eaten, data = Alien)
```

<br>

The object created stores a lot of information that you can retrieve directly or (better) indirectly using `lm()` companion functions:
```{r fit alien lm output}
names(fit_alien_lm)
```


## Fitting the Alien data using ```fitme()```

```{r fit alien lm2a, message=FALSE}
library(spaMM)
fit_alien_lm2 <- fitme(size ~ humans_eaten, data = Alien)
```

<br>

The object created also stores a lot of information that you can retrieve again directly or (better) indirectly using `fitme()` companion functions:
```{r fit alien lm2a output}
names(fit_alien_lm2)
```


## Extraction of the design matrix

### From fit with `lm`()
```{r dm lm, results='hold'}
model.matrix(fit_alien_lm)
```


## Extraction of the design matrix

### From fit with `fitme`()
```{r dm spaMM}
model.matrix(fit_alien_lm2)
```


## Extraction of model parameter estimates

### From fit with `lm`()
```{r fit coef, results='hold'}
coef(fit_alien_lm) ## you can also do fit_alien_lm$coefficients but using dedicated 
                   ## extractor (if available) is a better idea!
```

### From fit with `fitme`()
```{r fit coef spaMM, results='hold'}
fixef(fit_alien_lm2)
```


## Extraction of the error variance estimate

### From fit with `lm`()
```{r fit estimate error var}
summary(fit_alien_lm)$sigma^2 ## we expect something close to 25
```

### From fit with `fitme`()
```{r fit estimate error var spaMM}
fit_alien_lm2$phi * nrow(Alien) / (nrow(model.matrix(fit_alien_lm2)) - ncol(model.matrix(fit_alien_lm2)))
```

Note: in ML fit (default setting) `phi` contains the population residual variance (that is, it considers the sample as being the full population: denominator = $n$, not $n-p$), but the error variance estimate can be directly obtained from a REML fit (more on that later). The ability to switch between ML and REML fitting methods becomes particularly useful for more complex models that aim at estimating variances (mixed models). `lm()` does not have this flexibility.


## Extraction of the residuals

### From fit with `lm`()
```{r fit residuals}
residuals(fit_alien_lm)
```

### From fit with `fitme`()
```{r fit residuals spaMM}
residuals(fit_alien_lm2)
```


## Extraction of the population residual variance

### From fit with `lm`()
```{r fit estimate residual var}
summary(fit_alien_lm)$sigma^2 * (nrow(model.matrix(fit_alien_lm2)) - ncol(model.matrix(fit_alien_lm2))) / nrow(Alien)
```

Note: `sigma` contains the error standard deviation estimate, hence the transformation.

### From fit with `fitme`()
```{r fit estimate residual var spaMM}
fit_alien_lm2$phi
```


## Extraction of the log-likelihood

We will soon see what the likelihood is but for now consider that it is something quantifying how well the fit approximate the observations.

### From fit with `lm`()
```{r logLik}
logLik(fit_alien_lm)
```

### From fit with `fitme`()
```{r logLik spaMM}
logLik(fit_alien_lm2)
```


## Extraction of the residual sum of squares

This is another statistics quantifying how well the fit approximate the observations, which we will also study.

### From fit with `lm`()
```{r rss}
anova(fit_alien_lm)[2, "Sum Sq"] ## this is the same as sum(residuals(fit_alien_lm)^2)
```

### From fit with `fitme`()
```{r rss spaMM, results='hold'}
sum(residuals(fit_alien_lm2)^2) ## no extractor I can think of 
                                ## (even if in the specific case of LM you may use deviance(fit_alien_lm2))
```


# How fitting works?

## How to fit a LM without `lm()`?

In general, by:

### Maximum Likelihood (ML)
We want to find the $\widehat{\beta}$ (the column vector of model parameter estimates) maximizing the likelihood.

The likelihood of the model given the data is equal to the probability density assumed for those data given those parameter
values, that is: $\displaystyle {\mathcal {L}}(\theta \mid x) = P(x \mid \theta)$.

<br>

In the case of LM (and not GLM, LMM, GLMM), this is equivalent to performing the fit by:

### Ordinary Least Squares (OLS)

We want to find the $\widehat{\beta}$ minimizing the Residual Sum of Squares (RSS).

The RSS = $\displaystyle\sum_{i=1}^{n}{\varepsilon_i^2}$.


# Fitting LM by Ordinary Least Squares

## The Residual Sum of Squares (RSS)

<div class="columns-2">
```{r alien RSS plot, fig.align='center', fig.asp=1, fig.width=5, echo = FALSE, fig.cap="Residuals on best fit"}
par(las = 1)
pred <- fitted.values(fit_alien_lm)
resid <- residuals(fit_alien_lm)
plot(size ~ humans_eaten, data = Alien, ylab = "Alien size (cm)", xlab = "No. of humans eaten", asp = 1)
points(pred ~ humans_eaten, col = "blue", data = Alien, pch = 20)
with(Alien, segments(x0 = humans_eaten, x1 = humans_eaten, y0 = size, y1 = pred, col = "orange"))
legend("topleft", bty = "n", col = c("black", "blue"), pch = c(1, 20), legend = c("observation", "fitted value"))
```

```{r alien RSS plot2, fig.align='center', fig.asp=1, fig.width=5, echo = FALSE, fig.cap="Squared residuals on best fit"}
par(las = 1)
plot(
  size ~ humans_eaten,
  data = Alien,
  ylab = "Alien size (cm)",
  xlab = "No. of humans eaten",
  asp = 1
  )
points(pred ~ humans_eaten, col = "blue", data = Alien, pch = 20)
for (i in 1:nrow(Alien)) {
  with(Alien, polygon(
    x = c(
    humans_eaten[i],
    humans_eaten[i],
    humans_eaten[i] + abs(resid[i]),
    humans_eaten[i] + abs(resid[i])
    ),
    y = c(pred[i], size[i], size[i], pred[i]),
    col = rgb(red = 1, green = 0, blue = 0, alpha = 0.1),
    border = NA
    ))
}
with(Alien, segments(x0 = humans_eaten, x1 = humans_eaten, y0 = size, y1 = pred, col = "orange"))
legend("topleft", bty = "n", col = c("black", "blue"), pch = c(1, 20), legend = c("observation", "fitted value"))
text(-4, 65, paste("RSS =", round(sum(fit_alien_lm$resid^2), 1)), col = "red")
```
</div>


## The Residual Sum of Squares (RSS)

<div class="columns-2">
```{r alien RSS bad fit, fig.align='center', fig.asp=1, fig.width=5, echo = FALSE, fig.cap="Squared residuals on bad fit"}
par(las = 1)
plot(
  size ~ humans_eaten,
  data = Alien,
  ylab = "Alien size (cm)",
  xlab = "No. of humans eaten",
  asp = 1
  )
badslope <- 0.5
badintercept <- mean(Alien$size) - badslope * mean(Alien$humans_eaten)
badpred <- badintercept + badslope * Alien$humans_eaten
badresid <- Alien$size - badpred
points(badpred ~ humans_eaten, col = "blue", data = Alien, pch = 20)
for (i in 1:nrow(Alien)) {
  with(Alien, polygon(
    x = c(
    humans_eaten[i],
    humans_eaten[i],
    humans_eaten[i] + abs(badresid[i]),
    humans_eaten[i] + abs(badresid[i])
    ),
    y = c(badpred[i], size[i], size[i], badpred[i]),
    col = rgb(red = 1, green = 0, blue = 0, alpha = 0.1),
    border = NA
    ))
}
with(Alien, segments(x0 = humans_eaten, x1 = humans_eaten, y0 = size, y1 = badpred, col = "orange"))
legend("topleft", bty = "n", col = c("black", "blue"), pch = c(1, 20), legend = c("observation", "fitted value"))
text(-4, 65, paste("RSS =", round(sum(badresid^2), 1)), col = "red")
```
```{r alien RSS plot2 again, fig.align='center', fig.asp=1, fig.width=5, echo = FALSE, fig.cap="Squared residuals on best fit"}
par(las = 1)
plot(
  size ~ humans_eaten,
  data = Alien,
  ylab = "Alien size (cm)",
  xlab = "No. of humans eaten",
  asp = 1
  )
points(pred ~ humans_eaten, col = "blue", data = Alien, pch = 20)
for (i in 1:nrow(Alien)) {
  with(Alien, polygon(
    x = c(
    humans_eaten[i],
    humans_eaten[i],
    humans_eaten[i] + abs(resid[i]),
    humans_eaten[i] + abs(resid[i])
    ),
    y = c(pred[i], size[i], size[i], pred[i]),
    col = rgb(red = 1, green = 0, blue = 0, alpha = 0.1),
    border = NA
    ))
}
with(Alien, segments(x0 = humans_eaten, x1 = humans_eaten, y0 = size, y1 = pred, col = "orange"))
legend("topleft", bty = "n", col = c("black", "blue"), pch = c(1, 20), legend = c("observation", "fitted value"))
text(-4, 65, paste("RSS =", round(sum(fit_alien_lm$resid^2), 1)), col = "red")
```

</div>


## The Residual Sum of Squares (RSS)

<div class="columns-2">
```{r alien RSS true, fig.align='center', fig.asp=1, fig.width=5, echo = FALSE, fig.cap="Squared residuals on TRUE relationship"}
par(las = 1)
plot(
  size ~ humans_eaten,
  data = Alien,
  ylab = "Alien size (cm)",
  xlab = "No. of humans eaten",
  asp = 1
  )
badslope <- 1.5
badintercept <- 50
badpred <- badintercept + badslope * Alien$humans_eaten
badresid <- Alien$size - badpred
points(badpred ~ humans_eaten, col = "green", data = Alien, pch = 20)
for (i in 1:nrow(Alien)) {
  with(Alien, polygon(
    x = c(
    humans_eaten[i],
    humans_eaten[i],
    humans_eaten[i] + abs(badresid[i]),
    humans_eaten[i] + abs(badresid[i])
    ),
    y = c(badpred[i], size[i], size[i], badpred[i]),
    col = rgb(red = 1, green = 0, blue = 0, alpha = 0.1),
    border = NA
    ))
}
with(Alien, segments(x0 = humans_eaten, x1 = humans_eaten, y0 = size, y1 = badpred, col = "orange"))
legend("topleft", bty = "n", col = c("black", "green"), pch = c(1, 20), legend = c("observation", "mean response value"))
text(-4, 65, paste("RSS =", round(sum(badresid^2), 1)), col = "red")
```

```{r alien RSS plot2 again again, fig.align='center', fig.asp=1, fig.width=5, echo = FALSE, fig.cap="Squared residuals on best fit"}
par(las = 1)
plot(
  size ~ humans_eaten,
  data = Alien,
  ylab = "Alien size (cm)",
  xlab = "No. of humans eaten",
  asp = 1
  )
points(pred ~ humans_eaten, col = "blue", data = Alien, pch = 20)
for (i in 1:nrow(Alien)) {
  with(Alien, polygon(
    x = c(
    humans_eaten[i],
    humans_eaten[i],
    humans_eaten[i] + abs(resid[i]),
    humans_eaten[i] + abs(resid[i])
    ),
    y = c(pred[i], size[i], size[i], pred[i]),
    col = rgb(red = 1, green = 0, blue = 0, alpha = 0.1),
    border = NA
    ))
}
with(Alien, segments(x0 = humans_eaten, x1 = humans_eaten, y0 = size, y1 = pred, col = "orange"))
legend("topleft", bty = "n", col = c("black", "blue"), pch = c(1, 20), legend = c("observation", "fitted value"))
text(-4, 65, paste("RSS =", round(sum(fit_alien_lm$resid^2), 1)), col = "red")
```

</div>


## Home made LM fitting by Ordinary Least Squares

We compute a function that compute the RSS given the parameter estimates, the formula and the data:

```{r def compute_rss_Alien}
compute_rss <- function(vector.param, formula, data) {
  useful.data <- model.frame(formula = formula, data = data) ## ditch useless data and flag response
  X <- model.matrix(object = formula, data = useful.data) ## create the design matrix
  fitted_values <- X %*% matrix(vector.param) ## compute the fitted values given the coef
  response <- model.response(useful.data) ## extract the response
  sum((response - fitted_values)^2) ## compute and output the RSS
}
```

We can test that the function works by comparing the RSS at parameter values estimated with `lm()`:
```{r def compute_rss_Alien test}
compute_rss(vector.param = coef(fit_alien_lm), formula = size ~ humans_eaten, data = Alien)
anova(fit_alien_lm)[2, "Sum Sq"] ## value for reference
```


## Home made LM fitting by Ordinary Least Squares

We can now feed this function to an optimisation procedure that will find parameter values minimising the output by trying many combination of parameters:

```{r optim by rss}
res_optim_rss <- optim(par = c("intercept" = 0, "slope" = 1), ## the starting values for the parameters
                       fn = compute_rss, ## the function computing the output that must be minimised
                       formula = size ~ humans_eaten, ## any additional arguments for "fn"
                       data = Alien)                  ## any additional arguments for "fn"
```

Let's compare what we obtained from what `lm()` produced:
```{r optim by rss output}
res_optim_rss$par - coef(fit_alien_lm) ## very close to what is produced by lm() :-) !!!
res_optim_rss$value - sum(fit_alien_lm$residuals^2) ## very close to what is produced by lm() :-) !!!
```


# Fitting LM by Maximum (log-)Likelihood

## The maximum (log-)likelihood

```{r alien loglik}
Alien$fitted_values <- fitted.values(fit_alien_lm)
Alien$residuals <- residuals(fit_alien_lm)
Alien$population_residual_var <- sum(Alien$residuals^2) / nrow(Alien)
Alien$density <- dnorm(x = Alien$size, mean = Alien$fitted_values, sd = sqrt(Alien$population_residual_var))
Alien[1, ] ## we look in details at the first row, but you can look at all of them 
```

<div class="columns-2">

```{r alien lik plot, fig.align='center', fig.asp=1, fig.width=3.5, echo = FALSE}
par(mar = c(4, 4, 1, 1), las = 1)
curve(dnorm(x, mean = Alien[1, "fitted_values"],
            sd = sqrt(Alien$population_residual_var[1])),
      from = 30, to = 30 + Alien[1, "fitted_values"],
      ylab = "Probability density", xlab = "size")
abline(h = 0, lty = 2)
abline(v = Alien[1, "fitted_values"], lty = 2, col = "blue")
points(x = Alien[1, "fitted_values"], y = 0, pch = 20, col = "blue")
points(x = Alien[1, "size"], y = 0)
segments(x0 = Alien[1, "fitted_values"], x1 = Alien[1, "size"], y0 = 0, y1 = 0, col = "orange")
segments(x0 = Alien[1, "size"], x1 = Alien[1, "size"], y0 = 0, y1 = Alien[1, "density"], col = "purple")
arrows(x0 = Alien[1, "size"], x1 = 30, y0 = Alien[1, "density"], y1 = Alien[1, "density"], col = "purple", length = 0.1)
```

We can test that the function works by comparing the logLik at parameter values estimated with `lm()`:
```{r alien log density}
sum(log(Alien$density)) ## same as the log of the product!
logLik(fit_alien_lm)[1]
```

</div>


## Home made LM fitting by Maximum Likelihood

```{r def comput_logLik_Alien}
compute_logLik <- function(vector.param, formula, data) {
  useful.data <- model.frame(formula = formula, data = data) ## ditch useless data and flag response
  X <- model.matrix(object = formula, data = useful.data) ## create the design matrix
  fitted_values <- X %*% matrix(vector.param) ## compute the fitted values given the coef
  response <- model.response(useful.data) ## extract the response
  population_residual_var <- sum((response - fitted_values)^2) / nrow(useful.data)  ## var for the normal distrib
  sum(dnorm(response, mean = fitted_values, sd = sqrt(population_residual_var), log = TRUE))
}
```

We can test that the function works by comparing the logLik to that produced by `lm()`:
```{r def comput_logLik_Alien test}
compute_logLik(vector.param = coef(fit_alien_lm), formula = size ~ humans_eaten, data = Alien)
logLik(fit_alien_lm)[1] ## value for reference
```


## Home made LM fitting by Maximum Likelihood

We can now feed this function to an optimisation procedure that will find parameter values maximising the output:

```{r def logLik Alien computation}
res_optim_ML <- optim(par = c("intercept" = 0, "slope" = 1),
                      fn = compute_logLik,
                      formula = size ~ humans_eaten,
                      data = Alien,
                      control = list(fnscale = -1)) ## to indicate to maximise and not minimise
```

Let's compare what we obtained from what `lm()` produced:
```{r optim by ML output}
res_optim_ML$par - coef(fit_alien_lm) ## coef are very close!
res_optim_ML$value - logLik(fit_alien_lm)[1] ## logLiks are very close!
```


# Fitting methods comparison

## Comparing the three methods on a REAL dataset

```{r computation 3 methods}
fit_test_lm <- lm(height ~ weight*sex + cigarettes + drink, data = UK)

fit_test_homemade_OLS <- optim(par = rep(0, 8),
                               fn = compute_rss,
                               formula = height ~ weight*sex + cigarettes + drink,
                               data = UK,
                               method = "BFGS") ## better optim method (default is getting lost...)


fit_test_homemade_ML <- optim(par = rep(0, 8),
                              fn = compute_logLik,
                              formula = height ~ weight*sex + cigarettes + drink, data = UK,
                              control = list(fnscale = -1), method = "BFGS")
```
```{r computation 3 methods results}
round(rbind(lm = coef(fit_test_lm), OLS = fit_test_homemade_OLS$par, ML = fit_test_homemade_ML$par), 3)
```


## Comparing the three methods on a REAL dataset

Let's now perform a bench mark to compare the computational efficiency of the 3 fitting methods:
```{r computation 3 methods bench, warning=FALSE}
UKsmall <- UK[1:500, ] ## we reduce the sample size for saving time
microbenchmark::microbenchmark(
  lm = lm(height ~ weight*sex + cigarettes + drink, data = UKsmall),
  OLS = optim(par = rep(0, 8),
              fn = compute_rss,
              formula = height ~ weight*sex + cigarettes + drink,
              data = UKsmall,
              method = "BFGS"),
  ML = optim(par = rep(0, 8),
             fn = compute_logLik,
             formula = height ~ weight*sex + cigarettes + drink, data = UKsmall,
             control = list(fnscale = -1), method = "BFGS"),
  times = 10)
```
So our homemade functions work, but they are hugely inefficient!


# Fitting LM analytically

## What the statistical textbooks tell us

<center><font size = 8> $\widehat{\beta} = (X^\text{T}X)^{-1}X^\text{T}Y$ </font></center>

This gives best estimates **directly** (no need for optimisation / trials and errors):
```{r lin algebra}
Y <- matrix(Alien$size)
X <- model.matrix(fit_alien_lm)
solve(t(X) %*% X) %*% t(X) %*% Y  ## Tip: solve(x) returns the inverse of the matrix x
```

... but ```lm()``` does not even do that because it is still (somewhat) inefficient.

<br>

Geek note: $\widehat{Y} = X \widehat{\beta} = X (X^TX)^{-1}X^TY=HY$, and $H$ is called the hat matrix (it is a projection matrix).


## `lm()`-like fitting using QR decomposition

The goal is simply to decompose the design matrix into a product of two new matrices that present nice properties for doing maths super efficiently.

```{r QR}
qr_list <- qr(X)  ## same as fit_alien_lm$qr
Q <- qr.Q(qr_list, complete = TRUE)  ## orthogonal matrix n * n (transpose = inverse)
R <- qr.R(qr_list, complete = TRUE)  ## upper triangular matrix n * p
```
```{r QR2, eval = FALSE}
all.equal(Q %*% R, X, check.attributes = FALSE)  ## TRUE: Q %*% R is equal to X!!
```

```{r QR 3}
QTY <- t(Q) %*% Y ## same as fit_alien_lm$effects
backsolve(R, QTY) ## solve B from RB = QTY
```
<!-- using backsolve works on triangular matrix and is more efficient than solve(R[1:2, 1:2], QTY[1:2])) -->

* Note 1: `QTY` is a useful term that is also used by R to compute sum of squares efficiently (e.g. in `anova()`).
<!--Factor sum of squares is given by the first elements of QTY, it seems to depend on the design matrix. E.g. sum(QTY[1:2]^2) or just sum(QTY[2]^2). The residual sum of square would be given by sum(QTY[-c(1:2)]^2)```-->
* Note 2: other decompositions are sometimes used in other packages or other type of linear models (e.g. Cholesky decomposition).


## Actual `lm()` fitting procedure

First, `lm()` processes information for `lm.fit()`. It is quite a difficult function to read but are the important steps:
```{r lm guts, eval = FALSE}
lm(size ~ humans_eaten, data = Alien)
mf <- model.frame(size ~ humans_eaten, data = Alien)
Y  <- model.response(mf)
X  <- model.matrix(~ humans_eaten, data = Alien)
lm.fit(X, Y)
```

Then, `lm.fit()` calls a C function (Cdqrls) that calls a Fortran function (dqrls, which will compute estimates after QR) calling another Fortran function (dqrdc2, which performs the QR decomposition).

```
C     Dqrdc2 is a *modification* of Linpack's dqrdc ('DQRDC') for R
c
c     dqrdc2 uses householder transformations to compute the qr
c     factorization of an n by p matrix x.  a limited column
c     pivoting strategy based on the 2-norms of the reduced columns
c     moves columns with near-zero norm to the right-hand edge of
c     the x matrix.  this strategy means that sequential one
c     degree-of-freedom effects can be computed in a natural way.
c
c     i am very nervous about modifying linpack code in this way.
c     if you are a computational linear algebra guru and you really
c     understand how to solve this problem please feel free to
c     suggest improvements to this code.
```

<!--
It calls .Call(stats:::C_Cdqrls, X, Y, tol = 1e-7, TRUE)
The last logical only activates some checks about the input.
This function is defined in src/library/stats/src/lm.c
as SEXP Cdqrls(SEXP x, SEXP y, SEXP tol, SEXP chk).
It then calls
F77_CALL(dqrls)(REAL(qr), &n, &p, REAL(y), &ny, &rtol,
    REAL(coefficients), REAL(residuals), REAL(effects),
    &rank, INTEGER(pivot), REAL(qraux), work).
This fortan function is defined in src/appl/dqrls.f
as subroutine dqrls(x,n,p,y,ny,tol,b,rsd,qty,k,jpvt,qraux,work).
It is documented.
This then calls dqrdc2(x,n,n,p,tol,k,qraux,jpvt,work)
which is itself defined in the file src/appl/dqrdc2.f
as subroutine dqrdc2(x,ldx,n,p,tol,k,qraux,jpvt,work).
-->


## Dissecting the output from `lm()`

```{r lm output}
names(fit_alien_lm)
```
We have already seen all non-trivial components!

We have not yet seen:

* ```rank```: the number of columns that are linearly independent in the design matrix.
* ```assign```: it comes from ```attr(X, "assign")```; it indicates which parameters belong to each covariate.
* ```df.residuals```: nrow(Alien) - rank.
* ```xlevels```: here, empty.
* ```call```: the clean function call to ```lm()```.
* ```terms```: an object of class terms; i.e. the formula with many attributes (```?terms.object```).
* ```model```: it comes from ```model.frame()```; it gives the data used for the fit.


## What you need to remember

* that the variance of residuals is a biased estimator of the error variance
* how to extract key statistics for models fitted with `stats::lm()` & `spaMM::fitme()`
* how maximum likelihood estimations work


# Table of contents

## The Linear Model: LM

* 2.0 [Introduction](./LM_intro.html)
* 2.1 [Fitting procedure](./LM_fitting.html)
* 2.2 [Tests & Intervals](./LM_test_intervals.html)
* 2.3 [Assumptions & Outliers](./LM_assumptions.html)
* 2.4 [Let's practice](./LM_practice.html)

<br>

<div align="right">
[Back to main menu](./Title.html#2)
</div>
