---
title: "LM: Introduction"
subtitle: "exercises"
author: "Alexandre Courtiol"
date: "`r Sys.Date()`"
output:
  ioslides_presentation:
    widescreen: true
    smaller: true
vignette: >
  %\VignetteIndexEntry{2.0 z-------------------------------------------------------------exercises}
  %\VignetteEncoding{UTF-8}
  %\VignetteEngine{knitr::rmarkdown}
---

```{r setup, include=FALSE}
library(LM2GLMM)
knitr::opts_chunk$set(cache = FALSE,
                      fig.align = "center",
                      fig.width = 4,
                      fig.height = 4,
                      cache.path = "./cache_knitr/LM_intro_ex/",
                      fig.path = "./fig_knitr/LM_intro_ex/")
options(width = 110)
set.seed(1L)
```

# Design matrix

## Alternative representations of design matrices

If the average IQ for boys were 90 and 110 for girls, what would the parameter values be, for each of the following coding (using treatment contrasts):

* ```"boy"``` vs ```"girl"```
* ```"male"``` vs ```"female"```
* ```0``` vs ```1```
* ```1``` vs ```2```
* ```TRUE``` vs ```FALSE```

<br>

<div align="right">
[Back to main menu](./Title.html#2)
</div>


## Alternative representations of design matrices

Are these different representations of ages equivalent?

* ```"baby"``` vs ```"child"``` vs ```"adult"```
* ```0``` vs ```1``` vs ```2```
* ```1``` vs ```2``` vs ```3```


# Simulating data

## Turning script into function

Turn the following code into a function that takes the number of aliens to generate as an argument:

```{r alien data}
Alien <- data.frame(humans_eaten = rep(1:3, each = 2), intercept = 50, slope = 1.5)
Alien$error <- rnorm(n = 6, mean = 0, sd = sqrt(5))
Alien$size <- Alien$intercept + Alien$slope*Alien$humans_eaten + Alien$error
```


## Simulating interactions

Adapt the following code to include an interaction between `sex` and `bh_trips`.

```{r Alien}
set.seed(123)
Alien2 <- data.frame(planet = factor(c(rep("Chambr", 2), rep("Riplyx", 2), rep("Wickor", 2))),
                     sex = factor(rep(c("Z", "ZZ"), times = 3)),
                     bh_trips = 6:1)
Alien2$age <- rep(0, nrow(Alien2))
Alien2$age[Alien2$planet == "Chambr"] <- Alien2$age[Alien2$planet == "Chambr"] + 100 
Alien2$age[Alien2$planet == "Riplyx"] <- Alien2$age[Alien2$planet == "Riplyx"] + 1000
Alien2$age[Alien2$planet == "Wickor"] <- Alien2$age[Alien2$planet == "Wickor"] + 10000
Alien2$age[Alien2$sex == "Z"] <- Alien2$age[Alien2$sex == "Z"] + -20
Alien2$age[Alien2$sex == "ZZ"] <- Alien2$age[Alien2$sex == "ZZ"] + 20
Alien2$age <- Alien2$age + 0.5*Alien2$bh_trips
Alien2$age <- Alien2$age + rnorm(nrow(Alien2), mean = 0, sd = 4)
Alien2
```


# Model parameter estimates

## The UK dataset

For information on this dataset, check the details section in `?UK`.

```{r UK data}
head(UK)
fit_UK1 <- lm(height ~ drink + sex*weight, data = UK)
```


## Predictions with `fit_UK1`

Compute the following predictions by (1) creating a design matrix by hand & (2) using the function `predict()`:

* height of a 30 Kg boy whose mum drank most days?
* height of a 30 Kg girl whose mum drank most days?
* height of a 35 Kg boy whose mum drank 2 to 3 times a week?

```{r example predict, echo = FALSE, results='hide'}
newX <- matrix(
  c(1, 1, 0, 0, 0, 30, 0,
    1, 1, 0, 0, 1, 30, 30,
    1, 0, 0, 0, 0, 35, 0),
  nrow = 3, byrow = TRUE)
colnames(newX) <- names(coef(fit_UK1))
newX %*% coef(fit_UK1)
```

```{r example predict 3, echo = FALSE, results='hide'}
newdata1 <- data.frame(
  drink = c("Most days", "Most days", "2 to 3 times a week"),
  sex = c("Boy", "Girl", "Boy"),
  weight = c(30, 30, 35))
predict(fit_UK1, newdata = newdata1)
```


## Model parameter estimates of `fit_UK1`
```{r UK coef}
data.frame(coef(fit_UK1))
```
<br>

What do each of these estimates really mean?

(If you struggle interpreting parameter estimates, look at the fitted values)


## More predictions with `fit_UK1`

Predict and plot the influence of the weight of girls on their height! 


##

<div align="right">
[Back to main menu](./Title.html#2)
</div>
