---
title: "LM: Introduction"
subtitle: "Solutions"
author: "Alexandre Courtiol"
date: "`r Sys.Date()`"
output:
  html_vignette:
    toc: true
vignette: >
  %\VignetteIndexEntry{2.0 LM: Introduction -------------------------------------------solutions}
  %\VignetteEncoding{UTF-8}
  %\VignetteEngine{knitr::rmarkdown}
editor_options: 
  chunk_output_type: console
---

```{r setup, include=FALSE}
library(LM2GLMM)
knitr::opts_chunk$set(cache = FALSE,
                      fig.align = "center",
                      fig.width = 7,
                      fig.height = 5,
                      cache.path = "./cache_knitr/Exo_LM_intro_solution/",
                      fig.path = "./fig_knitr/Exo_LM_intro_solution/")
options(width = 90)
set.seed(1L)
```

# Design matrix

## Alternative representations of design matrices

### Instruction
If the average IQ for boys were 90 and 110 for girls, what would the parameter values be, for each of the following alternative coding for a boy and a girl (using treatment contrasts):

* ```"boy"``` vs ```"girl"```
* ```"male"``` vs ```"female"```
* ```0``` vs ```1```
* ```1``` vs ```2```
* ```TRUE``` vs ```FALSE```

### Solution

* ```"boy"``` vs ```"girl"```

```{r}
d <- data.frame(x = c("boy", "girl"))
model.matrix(~ x, data = d)
```
So the intercept represents here the IQ of males and thus should equate 90 and `xgirl` represents what needs to be added to the intercept to yield the value for females and thus should equate 20.

<br>

* ```"male"``` vs ```"female"```

```{r}
d <- data.frame(x = c("male", "female"))
model.matrix(~ x, data = d)
```
So the intercept represents here the IQ of females and thus should equate 110 and `xmale` represents what needs to be added to the intercept to yield the value for males and thus should equate -20.

<br>

* ```0``` vs ```1```

```{r}
d <- data.frame(x = c(0, 1))
model.matrix(~ x, data = d)
```
So the intercept represents here the IQ of males and thus should equate 90 and `x` represents a slope which multiplied by one and added to the intercept should yield the value for males.
That slope should thus equate 20.

<br>

* ```1``` vs ```2```

```{r}
d <- data.frame(x = c(1, 2))
model.matrix(~ x, data = d)
```
So the intercept represents nothing interesting, but when it is added to 1 times the slope `x` the total should equate the IQ of males.
Similarly when the intercept is added to 2 times the slope `x` the total should equate the IQ of females.

We thus have:

<center><font size="8"> $$\begin{cases} i + 1\times x = 90 \\ i + 2\times x = 110 \end{cases} $$ </font></center>

which you can solve as:

<center><font size="8"> $$90 -  1\times x + 2\times x = 110 \\ x = 20 $$ </font></center>

and thus

<center><font size="8"> $$i + 1\times 20 = 90 \\ i = 70$$ </font></center>

So here the intercept equates 70 and the slope `x` equates 20.

<br>


* ```TRUE``` vs ```FALSE```

Remember, in R `TRUE` and `FALSE` are represented by 1 and 0 respectively. If these are used in a model they will be treated as numeric values (1 and 0).

```{r}
as.numeric(c(TRUE, FALSE))
```

Therefore, this is essentially the opposite to the question with `0 vs 1`.

```{r}
d <- data.frame(x = c(TRUE, FALSE))
model.matrix(~ x, data = d)
```
So the intercept represents here the IQ of females and thus should equate 110 and `xTRUE` represents what needs to be added to the intercept to yield the value for males and thus should equate -20.


## Alternative representations of design matrices

### Instruction

Are these different representations of ages equivalent?

* ```"baby"``` vs ```"child"``` vs ```"adult"```
* ```0``` vs ```1``` vs ```2```
* ```1``` vs ```2``` vs ```3```

### Solution

Let's build all the design matrices:

```{r}
model.matrix(~ x, data = data.frame(x = c("baby", "child", "adult")))
model.matrix(~ x, data = data.frame(x = c(0, 1, 2)))
model.matrix(~ x, data = data.frame(x = c(1, 2, 3)))
```

No the representations are not equivalent and are actually all different.

# Simulating data

## Turning script into function

### Instruction

Turn the following code into a function that takes the number of aliens to generate as an argument:

*Tip:* Remember to check `?sample()` to understand what this new function does.

```{r alien data}
Alien <- data.frame(humans_eaten = sample(1:3, size = 6, replace = TRUE), intercept = 50, slope = 1.5)
Alien$error <- rnorm(n = 6, mean = 0, sd = sqrt(5))
Alien$size <- Alien$intercept + Alien$slope*Alien$humans_eaten + Alien$error
```

### Solution

```{r alien data solution}
simulate_myAliens <- function(N) {
  Alien <- data.frame(humans_eaten = sample(1:3, size = N, replace = TRUE),
                      intercept = 50, slope = 1.5)
  Alien$error <- rnorm(n = N, mean = 0, sd = sqrt(5))
  Alien$size <- Alien$intercept + Alien$slope*Alien$humans_eaten + Alien$error
  Alien
}
simulate_myAliens(N = 10)
```


## Simulating interactions

### Instruction

Adapt the following code to include an interaction between `sex` and `bh_trips`.

```{r Alien}
set.seed(123)
Alien2 <- data.frame(planet = factor(c(rep("Chambr", 2), rep("Riplyx", 2), rep("Wickor", 2))),
                     sex = factor(rep(c("Z", "ZZ"), times = 3)),
                     bh_trips = 6:1)
Alien2$age <- rep(0, nrow(Alien2))
Alien2$age[Alien2$planet == "Chambr"] <- Alien2$age[Alien2$planet == "Chambr"] + 100
Alien2$age[Alien2$planet == "Riplyx"] <- Alien2$age[Alien2$planet == "Riplyx"] + 1000
Alien2$age[Alien2$planet == "Wickor"] <- Alien2$age[Alien2$planet == "Wickor"] + 10000
Alien2$age[Alien2$sex == "Z"] <- Alien2$age[Alien2$sex == "Z"] + -20
Alien2$age[Alien2$sex == "ZZ"] <- Alien2$age[Alien2$sex == "ZZ"] + 20
Alien2$age <- Alien2$age + 0.5*Alien2$bh_trips
Alien2$age <- Alien2$age + rnorm(nrow(Alien2), mean = 0, sd = 4)
Alien2
```

### Solution

```{r Alien_other_int}
set.seed(123)
Alien3 <- data.frame(planet = factor(c(rep("Chambr", 2), rep("Riplyx", 2), rep("Wickor", 2))),
                     sex = factor(rep(c("Z", "ZZ"), times = 3)),
                     bh_trips = 6:1)
Alien3$age <- rep(0, nrow(Alien2))
Alien3$age[Alien2$planet == "Chambr"] <- Alien2$age[Alien2$planet == "Chambr"] + 100
Alien3$age[Alien2$planet == "Riplyx"] <- Alien2$age[Alien2$planet == "Riplyx"] + 1000
Alien3$age[Alien2$planet == "Wickor"] <- Alien2$age[Alien2$planet == "Wickor"] + 10000
Alien3$age[Alien2$sex == "Z"] <- Alien2$age[Alien2$sex == "Z"] + -20 + 5*Alien2$bh_trips[Alien2$sex == "Z"] 
Alien3$age[Alien2$sex == "ZZ"] <- Alien2$age[Alien2$sex == "ZZ"] + 20 + 10*Alien2$bh_trips[Alien2$sex == "ZZ"]
Alien3$age <- Alien2$age + 0.5*Alien2$bh_trips
Alien3$age <- Alien2$age + rnorm(nrow(Alien2), mean = 0, sd = 4)
Alien3
```


# Model parameter estimates

## The UK dataset

We start by fitting this model:

```{r UK data}
fit_UK1 <- lm(height ~ drink + sex*weight, data = UK)
```


## Predictions with `fit_UK1`

### Instruction

Compute the following predictions by (1) creating a design matrix by hand & (2) using the function `predict()`:

* height of a 30 Kg boy whose mum drank most days?
* height of a 30 Kg girl whose mum drank most days?
* height of a 35 Kg boy whose mum drank 2 to 3 times a week?

### Solution

```{r example predict}
names(coef(fit_UK1))

newX <- matrix(
  c(1, 1, 0, 0, 0, 30, 0,
    1, 1, 0, 0, 1, 30, 30,
    1, 0, 0, 0, 0, 35, 0), nrow = 3, byrow = TRUE)

colnames(newX) <- names(coef(fit_UK1))

newX

newX %*% coef(fit_UK1)
```

```{r example predict 3}
newdata1 <- data.frame(
  drink  = c("Most days", "Most days", "2 to 3 times a week"),
  sex    = c("Boy", "Girl", "Boy"),
  weight = c(30, 30, 35))
predict(fit_UK1, newdata = newdata1)
```


## Model parameter estimates of `fit_UK1`

### Instruction

```{r UK coef}
data.frame(coef(fit_UK1))
```

What do each of these estimates really mean?

(If you struggle interpreting parameter estimates, look at the fitted values)

### Solution

- `(Intercept)` gives the predicted height for an average boy weighing 0 Kg (silly but needed) and whose parents drink 2 to 3 times a week.
- `drinkMost_days` gives by how much the height of an average child increases if their parents drink most days, as compared to the average height of children whose parents drink 2 to 3 times a week.
- `drinkNat_at_all` gives by how much the height of an average child increases if their parents do not drink, as compared to the average height of children whose parents drink 2 to 3 times a week.
- `drinkOce_a_week_or_less` gives by how much the height of an average child increases if their parents drink once a week or less, as compared to the average height of children whose parents drink 2 to 3 times a week.
- `sexGirl` gives how much taller are, on average, girls of 0 Kg compared to boys of 0 Kg.
- `weight` gives by how much the height of an average boy increases when his weight increases by 1 Kg.
- `sexGirl:weight` gives by how much more the height of an average girl increases, compared to the increase in an average boy caused by an increase in weight of 1 Kg.


## More predictions with `fit_UK1`

### Instruction

Create a plot with marginal predictions showing the influence of the weight of girls on their height!

*Tip:* Look back at slides about marginal and conditional predictions to work out how to deal with the effect of 'drink'.

### Solution

For marginal predictions we use the *modal* value of non-focal qualitative variables within the set of all sampled observations.

```{r}

#The most frequent value of 'drink' is 'Not at all'
table(UK$drink)

```

Since weight influences predictions linearly, we only need predictions for the lightest and heaviest girl in our drink category.

```{r}
UK_girl <- UK[which(UK$sex == "Girl" & UK$drink == "Not at all"), ]

tab_weight <- data.frame(sex = "Girl",
                         drink = "Not at all",
                         weight = range(UK_girl$weight, na.rm = TRUE))

tab_weight
```

Now we can add the predicted values:

```{r}
tab_weight$height_pred <- predict(fit_UK1,
                                  newdata = tab_weight)
tab_weight
```

Create a plot to show the data.

```{r}

#Plot the points
plot(x = tab_weight$height_pred, y = tab_weight$weight)

#Or we can plot the points and lines
lines(x = tab_weight$height_pred, y = tab_weight$weight)

```

#### BONUS Solution `r .emo("nerd")`

How would we do this for each category of drink?

So let's start with identifying the relevant weights we need to use:

```{r range_weight}
UK_girl <- UK[which(UK$sex == "Girl"), ]

tapply(UK_girl$weight, UK_girl$drink, range, na.rm = TRUE)

tab_weight <- data.frame(sex = "Girl",
                         drink = rep(levels(UK_girl$drink), each = 2),
                         weight = unlist(tapply(UK_girl$weight, UK_girl$drink, range, na.rm = TRUE),
                                         use.names = FALSE))

tab_weight
```

Now we can add the predicted values:

```{r}
tab_weight$height_pred <- predict(fit_UK1, newdata = tab_weight)
tab_weight
```

And we can plot it (here using **{ggplot2}** for simplicity):

```{r plot_pred}
library(ggplot2)
ggplot(tab_weight) +
  aes(x = weight, y = height_pred, colour = drink) +
  geom_line() + 
  geom_point() +
  theme_bw()
```



