---
title: "LM: Introduction"
subtitle: "exercises"
author: "Alexandre Courtiol"
date: "`r Sys.Date()`"
output:
  ioslides_presentation:
    widescreen: true
    smaller: true
vignette: >
  %\VignetteIndexEntry{2.0 z-------------------------------------------------------------exercises}
  %\VignetteEncoding{UTF-8}
  %\VignetteEngine{knitr::rmarkdown}
---

```{r setup, include=FALSE}
library(LM2GLMM)
knitr::opts_chunk$set(cache = TRUE,
                      fig.align = "center",
                      fig.width = 4,
                      fig.height = 4,
                      cache.path = "./cache_knitr/LM_intro_ex/",
                      fig.path = "./fig_knitr/LM_intro_ex/")
options(width = 200)
set.seed(1L)
```


## Alternative representations of design matrices

### Are these different representations of gender equivalent?

* ```"boy"``` vs ```"girl"```
* ```"male"``` vs ```"female"```
* ```0``` vs ```1```
* ```1``` vs ```2```
* ```TRUE``` vs ```FALSE```

<br>

Create all corresponding design matrices to be sure!

<div align="right">
[Back to main menu](./Title.html#2)
</div>


## Alternative representations of design matrices

### If the average IQ for boys were 90 and 110 for girls, what would the parameter values be, for each of the following coding:

* ```"boy"``` vs ```"girl"```
* ```"male"``` vs ```"female"```
* ```0``` vs ```1```
* ```1``` vs ```2```
* ```TRUE``` vs ```FALSE```


## Alternative representations of design matrices

### Are these different representations of ages equivalent?

* ```"baby"``` vs ```"child"``` vs ```"adult"```
* ```0``` vs ```1``` vs ```2```
* ```1``` vs ```2``` vs ```3```

## Simulating data

Turn the following code into a function that takes the number of aliens to generate as an argument:

```{r alien data}
Alien <- data.frame(humans_eaten = rep(1:3, each = 2), intercept = 50, slope = 1.5)
Alien$error <- rnorm(n = 6, mean = 0, sd = sqrt(5))
Alien$size <- Alien$intercept + Alien$slope*Alien$humans_eaten + Alien$error
```


## Simulating data

Adapt the following code to include an interaction.

```{r Alien}
set.seed(123)
Alien2 <- data.frame(planet = factor(c(rep("Chambr", 2), rep("Riplyx", 2), rep("Wickor", 2))),
                     sex = factor(rep(c("Z", "ZZ"), times = 3)),
                     bh_trips = 6:1)
Alien2$age <- rep(0, nrow(Alien2))
Alien2$age[Alien2$planet == "Chambr"] <- Alien2$age[Alien2$planet == "Chambr"] + 100 
Alien2$age[Alien2$planet == "Riplyx"] <- Alien2$age[Alien2$planet == "Riplyx"] + 1000
Alien2$age[Alien2$planet == "Wickor"] <- Alien2$age[Alien2$planet == "Wickor"] + 10000
Alien2$age[Alien2$sex == "Z"] <- Alien2$age[Alien2$sex == "Z"] + -20
Alien2$age[Alien2$sex == "ZZ"] <- Alien2$age[Alien2$sex == "ZZ"] + 20
Alien2$age <- Alien2$age + 0.5*Alien2$bh_trips
Alien2$age <- Alien2$age + rnorm(nrow(Alien2), mean = 0, sd = 4)
Alien2
```

##

<div align="right">
[Back to main menu](./Title.html#2)
</div>

