---
title: "LM: Let's practice!"
subtitle: "solutions"
author: "Alexandre Courtiol"
date: "`r Sys.Date()`"
output: 
  html_vignette:
    toc: true
vignette: >
  %\VignetteEncoding{UTF-8}
  %\VignetteIndexEntry{2.4 z-------------------------------------------------------------solutions}
  %\VignetteEngine{knitr::rmarkdown}
editor_options: 
  chunk_output_type: console
---
```{r setup, include=FALSE}
library(LM2GLMM)
library(car)
knitr::opts_chunk$set(cache = FALSE, fig.align = "center", fig.width = 6, fig.height = 6,
                      cache.path = "./cache_knitr/LM_practice_ex/", fig.path = "./fig_knitr/LM_practice_ex/")
```

## Disclamer

In this vignette I illustrate the key steps required to solve the exercises. By no means I am trying to provide a detailed report of the analysis of each dataset. Any such report may need additional analyses and results would have to be written in good English, not in telegraphic style or lines of codes.

<br>

# Dataset: chickwts -- part 1 (difficulty = 0/5)

`r .emo("goal")` Compare the effectiveness of various feed supplements on the weights of six weeks old chickens.

## Exploring the data

We always start by looking at the data we have:

```{r, fig.width = 8}
str(chickwts)
plot(weight ~ feed, data = chickwts)
```

I don't know much about chicken but the values seem alright.

We look at the distribution of chicken among the sources of feed:
```{r}
table(chickwts$feed)
```

We check if there is any missing data:

```{r}
any(is.na(chickwts))
```

Great! There is no missing data.

## Fitting the model

We fit the model:

```{r}
(fit_chick <- lm(weight ~ feed, data = chickwts))
```

## Checking the assumptions

* linearity cannot be an issue here as we only have one qualitative predictor.
* multicollinearity is not an issue either as each chick has received a different feeding treatment.
* fixed values for the predictor: unless the diet has not been recorded correctly, there should not been any problem.

Let's have a look at the residuals of the model:

```{r}
par(mfrow = c(2, 2))
plot(fit_chick)
```

Everything looks very good. The errors should thus be independent, homoscedastic and normally distributed.


## Testing the effect of diet


Notes:

- because there is only one predictor, the following test also correspond to testing the fitted model against the fit of a null model.
- here anova() is sufficient because type I and II are the same with a single predictor.

```{r}
anova(fit_chick)
```

The diet seems to influence significantly the growth of chicken (F = `r round(anova(fit_chick)$F[1], 2)`, df1 = `r anova(fit_chick)$Df[1]`, df2 = `r anova(fit_chick)$Df[2]`, p-value < 0.001).

## Predicting the mean effect of each diet

```{r}
pred <- predict(fit_chick, newdata = data.frame(feed = levels(chickwts$feed)),
                interval = "confidence")
pred <- as.data.frame(pred)
pred$feed <- levels(chickwts$feed)
rownames(pred) <- pred$feed ## for automatic names in barplot
pred
```

Let's plot this:

```{r, results="hide"}
plot(pred[, "fit"] ~ I(1:nrow(pred) + 0.05), axes = FALSE, pch = 20,
     ylim = range(fit_chick$model$weight),
     ylab = "Chicken weight (gr)", xlab = "Diet")
axis(1, 1:nrow(pred), labels = rownames(pred))
axis(2)
box()
for (row in 1:nrow(pred)) {
  arrows(x0 = row + 0.05, y0 = pred[row, "lwr"], x1 = row + 0.05, y1 = pred[row, "upr"],
         code = 3, angle = 90, length = 0.1)
  diet <- levels(chickwts$feed)[row]
  weights.to.plot <- chickwts$weight[chickwts$feed == diet]
  points(rep(row, length(weights.to.plot)) - 0.05, weights.to.plot, col = "grey")
  }
```

The plot shows the predicted mean weight of chicken for each diet (black) with associated 95% confidence intervals. The observed weights of the chicken included in the analysis are represented by grey dots.

To make the plot even better we could have sorted the diet by increasing order of their effects.

We can also use **{ggplot2}** if you prefer:
```{r plot ggplot chicken}
library(ggplot2)
ggplot() +
  geom_point(aes(y = weight, x = feed),
             shape = 1,
             position = position_nudge(x = -0.1),
             data = chickwts) +
  geom_point(aes(y = fit, x = feed),
             position = position_nudge(x = 0.1),
             data = pred) +
  geom_errorbar(aes(ymin = lwr, ymax = upr, x = feed),
                width = 0.2,
                position = position_nudge(x = 0.1),
                data = pred) +
  labs(y = "Chicken weight (gr)", x = "Diet") +
  theme_minimal()
```


## Testing which diet differs from each others

```{r, message = FALSE, fig.width = 8}
summary(posthoc <- multcomp::glht(fit_chick, linfct = multcomp::mcp(feed = "Tukey")))
par(oma = c(0, 8, 0, 0))
plot(posthoc)
```

<br>

# Dataset: stackloss (difficulty = 1/5)

`r .emo("goal")` Find out if the acid concentration influences the loss of ammonia by the plants.

## Exploring the data

```{r}
str(stackloss)
summary(stackloss)
pairs(stackloss)
any(is.na(stackloss))
```


## Fitting the model

We will consider all predictors since water and air flow seem to be strong determinants of the loss of ammonia in plants.

```{r}
fit_stack <- lm(stack.loss ~ ., data = stackloss)
```

Note: the `.` is a shortcut to include all predictors at once.


## Checking the assumptions

### Linearity

Let’s examine the linearity assumption by plotting the residuals against each column of the design matrix (but the intercept):

```{r}
for (col in 2:ncol(model.matrix(fit_stack))) {
  plot(rstandard(fit_stack) ~ model.matrix(fit_stack)[, col])
  title(main = colnames(model.matrix(fit_stack))[col])
}
```

It looks messy but there is no obvious pattern suggesting a particular transformation of the predictors.

### Multicollinearity

```{r}
cor(model.matrix(fit_stack)[ , -1])
cov2cor(vcov(fit_stack))
vif(fit_stack)
```

There are some multicollinearity problems but it does not seem to impact much the effect of acidity which is our focus. Note that the very strong correlation between slopes and the intercept are to be expected, as well as the one between a variable and itself.

### Other assumptions

We will assume that no there is no measurement error and will focus on the analysis of the residuals of the model:

```{r}
par(mfrow = c(2, 2))
plot(fit_stack)
```

That is not fantastic but it is not specifically worse than what we may get under the assumption of LM with so few data. We can check this by simulating data under the assumption of LM and comparing the plots we get:

```{r}
par(mfrow = c(2, 2))
plot(update(fit_stack, as.matrix(simulate(fit_stack)) ~ .))  ## simulation 1
plot(update(fit_stack, as.matrix(simulate(fit_stack)) ~ .))  ## simulation 2
```

It does not look particularly better.

We can also use the package **{DHARMa}** which compares your residuals to many simulations under the data generating process hypothesised (more details in GLM course):

```{r}
DHARMa::simulateResiduals(fit_stack, plot = TRUE)
```

It is hard to be sure due to the size of the dataset but we will consider the assumptions fulfilled.


## Comparison to the null model

Because the model involves several predictors, before anything else we need to make sure that the overall model offers a higher goodness of fit than the null model:

```{r}
fit_stack_H0 <- update(fit_stack, . ~ 1)
anova(fit_stack_H0, fit_stack)
```

All good: the model predicts the data significantly better than the null model.


## Testing the effect of the acid concentration

```{r}
summary(fit_stack)$coef["Acid.Conc.", ]
```

The acid seems to have no significant influence on the loss of ammonia (t = ```r round(summary(fit_stack)$coef["Acid.Conc.", 3], 2)```, df = ```r fit_stack$df.residual```, p-value = ```r round( summary(fit_stack)$coef["Acid.Conc.", 4], 2)```).

Let's plot the effect very simply using the pacakge **{effects}**.

```{r}
plot(effects::effect("Acid.Conc.", mod = fit_stack))
```


## Considering interactions?

Because we are specifically interested in the effect of the acid and that this effect may depend on the other predictors, we could check if considering interactions between the acidity and other predictors would improve the goodness of fit of the model:

```{r}
fit_stack_int <- lm(stack.loss ~ Acid.Conc. * (Air.Flow + Water.Temp), data = stackloss)
anova(fit_stack, fit_stack_int)
vif(fit_stack_int)
```
The goodness of fit is not significantly better and there are huge multicollinearity issues, so let's not go there.


<br>


# Dataset: swiss (difficulty = 1/5)

`r .emo("goal")` Plot the influence of potential determinant of fertility.

## Exploring the data

```{r}
str(swiss)
summary(swiss)
pairs(swiss)
any(is.na(swiss))
```

## Fitting the model

```{r}
fit_swiss <- lm(Fertility ~ ., data = swiss)
```


## Checking the assumptions

### Linearity

Let's examine the linearity assumption by plotting the (standardised) residuals against each column of the design matrix (but the intercept):

```{r}
for (col in 2:ncol(model.matrix(fit_swiss))) {
  plot(rstandard(fit_swiss) ~ model.matrix(fit_swiss)[, col])
  title(main = colnames(model.matrix(fit_swiss))[col])
}
```

No obvious non-linear pattern emerges.

### Multicollinearity

We can look at the correlation between the column vectors of the model matrix:

```{r}
cor(model.matrix(fit_swiss)[, -1])  ## we discard the intercept for that
```

We can also look at the correlation between parameter estimates:

```{r}
cov2cor(vcov(fit_swiss))
```

There are some noticeable correlations between predictors and between parameter estimates. Let's double check the extent to which multicollinearity could be an issue by looking at the variance inflation factors:

```{r}
vif(fit_swiss)
```

The situation is not ideal but nothing looks catastrophic. Let's see if the situation gets better once the variable ```Examination``` is omitted:

```{r}
fit_swiss_bis <- update(fit_swiss, . ~ . - Examination)
vif(fit_swiss_bis)
```

The situation is much better so we will study predictions stemming from the fit with `Examination` and  from the fit without it, to examine the robustness of our results.

### Fixed values for predictor

Because we do not know on how many individuals the percentages have been based, we will assume that they did so on large sample and that this assumption is fulfilled.

### Errors

```{r}
par(mfrow = c(2, 2))
plot(fit_swiss)
plot(fit_swiss_bis)
```

Both model fits look alright (even if not super great).

## Comparison to the null model

Because the model involves several predictors, before anything else we need to make sure that the overall model offers a higher goodness of fit than the null model:

```{r}
fit_swiss_H0 <- update(fit_swiss, . ~ 1)
anova(fit_swiss_H0, fit_swiss)
anova(fit_swiss_H0, fit_swiss_bis)
```

Both of our models fit the data clearly better than the null model.

## Plotting the effect of the potential determinant of fertility

Because this is a simple LM, we can try to use the function ```allEffects``` from the package ```effects``` to do that automatically:

```{r plot effects, fig.height = 8, fig.width = 8, message=FALSE}
plot(effects::allEffects(fit_swiss))
plot(effects::allEffects(fit_swiss_bis))
```

Many other packages can probably do the same. One example is ```visreg```:

```{r plot visreg, fig.height = 8, fig.width = 8}
par(mfrow = c(2, 2))
plot(visreg::visreg(fit_swiss_bis))
```

Both models predict the same overall trends. To illustrate the difference in predicted values between the two models, one would have to do the plot manually instead of using some easy package. Let's do that to illustrate how the prediction for the mean effect of education differ between models:

```{r}
data.for.pred <- data.frame("Agriculture" = mean(swiss$Agriculture), 
                            "Examination" = mean(swiss$Examination),
                            "Education" = seq(min(swiss$Education), max(swiss$Education), length = 30),
                             "Catholic" = mean(swiss$Catholic),
                             "Infant.Mortality" = mean(swiss$Infant.Mortality)
                             )
pred1 <- predict(fit_swiss, newdata = data.for.pred, interval = "confidence")
pred2 <- predict(fit_swiss_bis, newdata = data.for.pred, interval = "confidence")
pred1_df_temp <- as.data.frame(pred1) ## to turn the matrix into a data frame
pred2_df_temp <- as.data.frame(pred2) ## to turn the matrix into a data frame
pred1_df <- cbind(data.for.pred, pred1_df_temp)  ## we add the new data
pred2_df <- cbind(data.for.pred, pred2_df_temp)  ## we add the new data
```

```{r}
plot(fit ~ Education, data = pred1_df, col = "blue", lwd = 2, type = "l",
     ylab = "Predicted Standardized Fertility (+/- 95% CI)", xlab = "Education (%)")
points(lwr ~ Education, data = pred1_df,
       col = "blue", lwd = 2, lty = 2, type = "l")
points(upr ~ Education, data = pred1_df,
       col = "blue", lwd = 2, lty = 2, type = "l")

points(fit ~ Education, data = pred2_df,
       col = "orange", lwd = 2, type = "l")
points(lwr ~ Education, data = pred2_df,
       col = "orange", lwd = 2, lty = 2, type = "l")
points(upr ~ Education, data = pred2_df,
       col = "orange", lwd = 2, lty = 2, type = "l")

legend("topright", fill = c("blue", "orange"),
       legend = c("fit_swiss", "fit_swiss_bis"), title = "Model:", bty = "n")
```

Note that while predicting the effect of ```Education``` we set all other covariates to their mean.


Same with **{ggplot2}**:

```{r}
full_pred <- rbind(cbind(pred1_df, mod = "fit_swiss"),
                   cbind(pred2_df, mod = "fit_swiss_bis")) ## we combine all the data
full_pred
ggplot(full_pred) +
  aes(fill = mod, x = Education) +
  geom_ribbon(aes(ymax = upr, ymin = lwr), alpha = 0.3) +
  geom_line(aes(y = fit, colour = mod)) +
  labs(y = "Predicted Standardized Fertility (+/- 95% CI)", x = "Education (%)") +
  theme_minimal()
```


We clean the session for the next exercise:

```{r}
rm(list = ls())
```

<br>


# Dataset: InsectSprays (difficulty = 2/5)

`r .emo("goal")` Compare the effectiveness of various insecticides on the number of insects in agricultural experimental units.

`r .emo("goal")` What is the efficiency of spray C?


## Exploring the data

```{r}
str(InsectSprays)
range(InsectSprays$count)
table(InsectSprays$spray)
plot(count ~ spray, data = InsectSprays)
any(is.na(InsectSprays))
```


## Fitting the model

```{r}
(fit_insect <- lm(count ~ spray, data = InsectSprays))
```

## Checking the assumptions

As for the dataset ```chickwts```, because we only study the effect of one qualitative predictor, we expect linearity and no multicollinearity, and that the values for the predictor are fixed (as assumed). We can thus directly turn to the analysis of the prediction errors:

```{r}
par(mfrow = c(2, 2))
plot(fit_insect)
```

That does not look so good. There seem to be some heteroscedasticity in the data. Let's test this possibility:

```{r}
lmtest::bptest(fit_insect)
```

The Breusch-Pagan test shows that we can reject the assumption of homoscedasticity. One option would be to try to use GLM but we have not yet seen this, so let's try to reduce the problems and apply a LM fit anyhow. The first thing top try to solve the problems is to use a power transformation of the response variable.

## Box Cox

For the Box Cox transformation to work, we need all the values to be strictly positive, but we saw that the minimum value is zero, so let's start by adding one to every count and let's refit the model.

```{r}
InsectSprays$count_bis <- InsectSprays$count + 1
fit_insect_bis <- update(fit_insect, count_bis ~ .)
```

We can now start the Box Cox analysis:

```{r}
car::boxCox(fit_insect_bis)
summary(bc <- car::powerTransform(fit_insect_bis))
```

The estimated value for lambda is close to 1/3 which corresponds to a cube root exponent to the observation but this is not particularly meaningful so we will stick to the best estimate for lambda which is ```r bc$lambda``` and stored in the object bc$lambda.

Let's fit the model on the transformed data and display the new diagnostic plots:

```{r}
InsectSprays$count_bc <- car::bcPower(InsectSprays$count_bis, bc$lambda)
fit_insect_bc <- update(fit_insect_bis, count_bc ~ .)
par(mfrow = c(2, 2))
plot(fit_insect_bc)
```

That looks fantastic! The heteroscedasticity is gone:

```{r}
lmtest::bptest(fit_insect_bc)
```

The lack of dependence should not be an issue (although the original paper where these data have been published showed that the experimental design was not as simple as the dataset here suggests), but let's check anyhow:

```{r}
lmtest::dwtest(fit_insect_bc)
car::durbinWatsonTest(fit_insect_bc)
```

Independence is not rejected, irrespective of whether doing a unilateral or bilateral test.

The normality looked good on the plots but we could also test it:

```{r}
nortest::lillie.test(residuals(fit_insect_bc))
shapiro.test(residuals(fit_insect_bc))
```
Again, no issue here.

## Testing the effect of insecticides

Note: once again, because there is only one predictor, the following test also correspond to testing the fitted model against the fit of a null model.

```{r}
anova(fit_insect_bc)  ## again, same here as car::Anova as only one predictor
```

The type of insecticide seems to influence significantly the number of insects counted (F = ```r round(anova(fit_insect_bc)$F[1], 2)```, df1 = ```r anova(fit_insect_bc)$Df[1]```, df2 = ```r anova(fit_insect_bc)$Df[2]```, p-value < 0.001).

## Predicting the effect of spray C

```{r}
(meanC_BC <- predict(fit_insect_bc, newdata = data.frame(spray = "C"), interval = "confidence"))
```
This corresponds to the number of insects + 1 on the Box Cox scale. Thus we need to transform this result back on the original scale:

```{r}
(meanC <- ((meanC_BC * bc$lambda) + 1)^(1/bc$lambda) - 1)
car::bcnPowerInverse(meanC_BC, lambda = bc$lambda, gamma = 0) - 1  ## same thing with less maths
```
We thus expect to find on average (95%CI) ```r  round(meanC[1, "fit"], 2)``` (```r round(meanC[1, "lwr"], 2)``` -- ```r round(meanC[1, "upr"], 2)```) insects on agricultural experimental units treated with the insecticide C.

We clean the session for the next exercise:

```{r}
rm(list = ls())
```

<br>


# Dataset: chickwts -- part 2 (difficulty = 0/5)

`r .emo("goal")`  Predict the proportion of chicken larger than 300gr for each feed supplements (difficulty 3/5).


## Predicting the proportion of chicken heavier than 300 grams

```{r}
fit_chick <- lm(weight ~ feed, data = chickwts)
```

We have already studied this model fit above, so we want do it again here.

```{r}
pred2 <- predict(fit_chick,
                newdata = data.frame(feed = levels(chickwts$feed)),
                interval = "prediction",
                se.fit = TRUE)

se.pred <- sqrt(pred2$se.fit^2 + pred2$residual.scale^2)

freq_more300 <- pnorm(q = 300,
                      mean = pred2$fit[, "fit"],
                      sd = se.pred, lower.tail = FALSE)

output <- as.data.frame(cbind(pred2$fit, "%>300gr" = round(freq_more300, 2)))

output$feed <- levels(chickwts$feed)

output
```

We can plot this: 

```{r, fig.width = 8}
barplot(sort(output$`%>300gr`), names.arg = output$feed[order(output$`%>300gr`)],
        ylim = c(0, 1), ylab = "Predicted proportion of chicken larger than 300 gr")
```

or using tidy packages:

```{r, fig.width = 8}
output$feed2 <- forcats::fct_reorder(output$feed, output$`%>300gr`)

ggplot(output) +
    geom_col(aes(x = feed2, y = `%>300gr`)) +
    labs(y = "Predicted proportion of chicken larger than 300 gr") +
    theme_minimal()
```

We clean the session for the next exercise:

```{r}
rm(list = ls())
```

<br>


# Dataset: mammals (difficulty = 3/5)

`r .emo("goal")` What is the allometric exponent for the growth of brain size with body mass?

`r .emo("goal")` Does this exponent agree with the usual expectation of 2/3?

`r .emo("goal")` How large do we expect a 1kg animal to be?

`r .emo("goal")` Rank organisms by relative brain size (i.e. controlled for body size)

`r .emo("goal")` Repeat the analysis with the dataset `Animals`


## Explore the data

```{r}
head(mammals)
str(mammals)
plot(brain ~ body, data = mammals)
plot(log(brain) ~ log(body), data = mammals, pch = 20, col = "blue")
text(log(mammals$body), log(mammals$brain) + 0.4,
     labels = rownames(mammals), cex = 0.8)
```

## Reshaping the data
```{r}
mammals$log_brain <- log(mammals$brain)
mammals$log_body <-  log(mammals$body)
```

## Fitting the model
```{r}
fit <- lm(log_brain ~ log_body, data = mammals)
plot(fit)
influence.measures(fit)
plot(fit, which = 4)
summary(fit)
fit_nohuman <- lm(log_brain ~ log_body, data = mammals[rownames(mammals) != "Human", ])
plot(fit_nohuman, which = 4)
```

## Comparing to 2/3

```{r}
t <- (fit$coefficients[2] - 2/3) / sqrt(diag(vcov(fit))[2])
2*pt(abs(t), df = fit$df.residual, lower.tail = FALSE)
confint(fit)
fit2 <- lm(log_brain ~ 1 + offset(2/3*log_body), data = mammals)
anova(fit2, fit)
```


```{r}
t <- (fit_nohuman$coefficients[2] - 2/3) / sqrt(diag(vcov(fit_nohuman))[2])
2*pt(abs(t), df = fit_nohuman$df.residual, lower.tail = FALSE)
confint(fit_nohuman)
fit2_nohuman <- lm(log_brain ~ 1 + offset(2/3*log_body), data = mammals[rownames(mammals) != "Human", ])
anova(fit2_nohuman, fit_nohuman)
```

We can plot the predictions from the models

```{r}
plot(log_brain ~ log_body, data = mammals, col = "blue")
text(mammals$log_body, mammals$log_brain + 0.5,
     labels = rownames(mammals), cex = 0.8)
abline(fit, col = "blue", lwd = 2)
abline(fit_nohuman, col = "blue", lwd = 2, lty = 2)
abline(coef(fit2), 2/3, col = "red", lwd = 2)
abline(coef(fit2_nohuman), 2/3, col = "red", lwd = 2, lty = 2)
legend("topleft", lty = c(1, 1, 2, 2), lwd = c(2, 2, 2, 2), col = c("blue", "red", "blue", "red"),
       legend = c("slope estimated", "slope = 2/3", "slope estimated (no human)", "slope = 2/3 (no human)"))
```


## Prediction of the brain size for an animal of 1Kg

```{r}
p <- predict(fit_nohuman, newdata = data.frame(log_body = log(1)))
exp(p)
```

An animal of 1 Kg is predicted to have a brain size of ```r round(exp(p), 2)``` grams.


## Let us rank the organism by brain size controling for body size

```{r}
matrix(names(sort(resid(fit), decreasing = TRUE)), ncol = 1)
```


# Dataset: Animals (difficulty = 3/5)

`r .emo("goal")` What is the allometric exponent for the growth of brain size with body mass?

`r .emo("goal")` Does this exponent agree with the usual expectation of 2/3?

`r .emo("goal")` How large do we expect a 1kg animal to be?

`r .emo("goal")` Rank organisms by relative brain size (i.e. controlled for body size)


## Explore the data

```{r}
head(Animals)
str(Animals)
plot(brain ~ body, data = Animals)
plot(log(brain) ~ log(body), data = Animals, col = "blue",
     ylim = c(min(log(brain)), max(log(brain)) + 0.5))
text(log(Animals$body), log(Animals$brain) + 0.5,
     labels = rownames(Animals), cex = 0.8)
```

## Reshaping the data
```{r}
Animals$log_brain <- log(Animals$brain)
Animals$log_body <-  log(Animals$body)
```

## Fitting the model
```{r}
fit <- lm(log_brain ~ log_body, data = Animals)
plot(fit)
influence.measures(fit)
plot(fit, which = 4)
summary(fit)
Animals_nodino <- Animals[!rownames(Animals) %in%
                          c("Triceratops", "Brachiosaurus", "Dipliodocus"), ]
fit_nodino <- lm(log_brain ~ log_body, data = Animals_nodino)
plot(fit_nodino)
influence.measures(fit_nodino)
plot(fit_nodino, which = 4)
summary(fit_nodino)
```

## Plot

```{r}
plot(log_brain ~ log_body, data = Animals, col = NULL)
text(x = Animals$log_body, y = Animals$log_brain,
     labels = rownames(Animals), cex = 0.8,
     col = ifelse( rownames(Animals) %in% c("Triceratops", "Brachiosaurus", "Dipliodocus"), "red", "black"))
p <- predict(fit, newdata = data.frame("log_body" = range(log(Animals$body))))
p_nodino <- predict(fit_nodino, newdata = data.frame("log_body" = range(log(Animals$body))))
points(x = range(Animals$log_body), y = p, col = "red", type = "l", lty = 2)
points(x = range(Animals$log_body), y = p_nodino, col = "blue", type = "l")
```

<br>


# Dataset: trees (difficulty = 3/5)

`r .emo("goal")` Compare the approximation of the volume of wood given by $\text{Volume} = c\times\text{Height}\times\text{Girth}^2$ (with $c$ to be estimated) to the approximation of a tree trunk as a cylinder.

## Exploring the data

```{r trees}
str(trees)
summary(trees)
pairs(trees)
any(is.na(trees))
```

## Fitting the model (Solution 1)

```{r solution 1}
trees$HG2 <- trees$Height*trees$Girth^2
fit_trees1 <- lm(Volume ~ HG2 - 1, data = trees)
```

### Checking the assumptions

There is no need to check our first 3 assumptions as the model structure is constrained, but we should check the assumptions concerning the errors.

```{r plot mod 1}
par(mfrow = c(2, 2))
plot(fit_trees1)
```

## Fitting the model (Solution 2)

$\text{Volume} = c \times \text{Height} \times \text{Girth}^2$ is not linear but we can turn it into a linear expression by using logs: 

$$\text{log(Volume)} = \text{log}(c) + \text{log(Height)} + 2 \times \text{log(Girth)}$$

We will thus substitute variables by their logs and try to fit the following model:

$$ \text{log_Volume} = k + 1 \times \text{log_Height} + 2 \times \text{log_Girth} + \epsilon$$

with $k = \text{log}(c)$. In this LM only $k$ must be estimated and the other slopes are considered as fixed. We thus fit the model as follow:

```{r}
trees$log_Volume <- log(trees$Volume)
trees$log_Height <- log(trees$Height)
trees$log_Girth <- log(trees$Girth)
fit_trees2 <- lm(log_Volume ~ 1 + offset(log_Height + 2 * log_Girth), data = trees)
```

### Checking the assumptions

```{r plot mod 2}
par(mfrow = c(2, 2))
plot(fit_trees2)
```

It looks alright, let's no forget that we only have ```r nrow(trees)``` rows in the datasets.

## Let us compare how each model fits the data

```{r r sq mod}
cor(predict(fit_trees1), trees$Volume)^2
cor(exp(predict(fit_trees2)), trees$Volume)^2
```


## Estimation of the parameter $c$

```{r}
coef(fit_trees1)
confint(fit_trees1)
```

The estimation with confidence interval of $k$ are:

```{r}
coef(fit_trees2)
confint(fit_trees2)
```

Because $k$ is correspond to the log of $c$, we must raise the obtained value to the exponential to obtain the estimation of $c$:

```{r}
exp(coef(fit_trees2))
exp(confint(fit_trees2))
```

We compare the two modelling approach in percentage of the estimation of $c$ by model 1:

```{r}
100*(coef(fit_trees1)[[1]] - exp(coef(fit_trees2))[[1]])/coef(fit_trees1)[[1]]
```


## Comparison to the approximation of a tree trunk by a cylinder

The volume of a cylinder is the height times the surface of a circle, so $h \pi r^2$ with $r$ the radius or  $h\pi(d/2)^2$ with $d$ the diameter. The girth is here the diameter as mentioned in the help file and not the perimeter.

In the formula, all units should be the same but here the diameter is in inches but the height is measured in feet and the volume in cubic feet. This thus lead to the volume of the tree as a cylinder being approximated by $\text{Height} \times \pi \times \left( \text{conv}_\text{in_to_ft} \text{Radius} \right)^2$. Since one foot is 12 inches, $\text{conv}_\text{in_to_ft} = 1/12$, leading to $\text{Volume} = \text{Height} \times \pi \times \left(\frac{\text{Girth}}{24}\right)^2$.

```{r mod 3}
trees$HPG2 <- trees$Height*pi*(trees$Girth/24)^2
fit_trees3 <- lm(Volume ~ -1 + offset(HPG2), data = trees)
```

We can compare both models explicitly:

```{r}
anova(fit_trees3, fit_trees1)
```

`fit_trees1` is a much better fit of the data than `fig_trees3`

Let us plot the differences in prediction between models:

```{r}
girth.for.pred <- seq(min(trees$Girth), max(trees$Girth), length = 4)
height.for.pred <- seq(min(trees$Height), max(trees$Height), length = 10)
data.for.pred <- expand.grid(Girth = girth.for.pred, Height = height.for.pred)

data.for.pred$HG2 <- data.for.pred$Height*data.for.pred$Girth^2
data.for.pred$HPG2 <- data.for.pred$Height*pi*(data.for.pred$Girth/24)^2

pred1 <- predict(fit_trees1, newdata = data.for.pred)
pred2 <- predict(fit_trees3, newdata = data.for.pred)

data.for.plot <- as.data.frame(cbind(data.for.pred, fit1 = pred1, fit2 = pred2))

plot(Volume ~ Height, data = trees, cex = 5*trees$Girth/max(trees$Girth), pch = 21,
     ylim = range(c(data.for.plot$fit, data.for.plot$fit2, trees$Volume)),
     ylab = "Volume (cubic feet)", xlab = "Height (meters)", col = "brown", bg = "yellow", lwd = 3)

for (girth_class in 1:length(unique(girth.for.pred))) {
  with(subset(data.for.plot,
              data.for.plot$Girth == unique(girth.for.pred)[girth_class]), {
    points(fit1 ~ Height, type = "l", lwd = girth_class)
    points(fit2 ~ Height, type = "l", lwd = girth_class, col = "red")
  })
}
```

The plot shows the influence of height on the mean volume for 4 specific girths (```r round(girth.for.pred, 2)``` inches). Thicker lines represent trees with larger girth. The black lines depict the predictions from the model ```fit_trees``` and the red ones are for the model ```fit_trees_cylinder```. The raw data are represented by the brown and yellow dots. The diameter of these dots is proportional to the actual measured girth.

That the approximation of the volume of trees as cylinders over-estimate the actual volume may be due to the fact that the diameter of trees goes down as you climb up the tree...


We clean the session for the next exercise:

```{r}
rm(list = ls())
```

<br>
