---
title: "LM: Let's practice!"
author: "Alexandre Courtiol"
date: "`r Sys.Date()`"
output:
  ioslides_presentation:
    widescreen: true
    smaller: true
vignette: >
  %\VignetteIndexEntry{2.4 Let's practice}
  %\VignetteEncoding{UTF-8}
  %\VignetteEngine{knitr::rmarkdown}
---

```{r setup, include=FALSE}
library(LM2GLMM)
knitr::opts_chunk$set(cache = TRUE, fig.align = "center", fig.width = 4, fig.height = 4,
                      cache.path = "./cache_knitr/Exo_LM/", fig.path = "./fig_knitr/Exo_LM/")
options(width = 200)
set.seed(1L)
```

## The Linear Model: LM

* 2.0 [Introduction](./LM_intro.html)
* 2.1 [Fitting procedure](./LM_fitting.html)
* 2.2 [Tests & Intervals](./LM_test_intervals.html)
* 2.3 [Assumptions & Outliers](./LM_assumptions.html)
* 2.4 [Let's practice](./LM_practice.html)

<br>

<div align="right">
[Back to main menu](./Title.html#2)
</div>


# Exercises

## Dataset: chickwts (difficulty = 0/5)

```{r chickwts}
head(chickwts)
levels(chickwts$feed)
```

### Goal:

* compare the effectiveness of various feed supplements on the weights of six weeks old chickens.

* optional: predict the proportion of chicken larger than 300gr for each feed supplements (difficulty 3/5)


## Dataset: InsectSprays (difficulty = 2/5)

```{r InsectSprays}
head(InsectSprays, n = 3)
levels(InsectSprays$spray)
```


### Goal

* Compare the effectiveness of various insecticides on the number of insects in agricultural experimental units.

* What is the efficiency of spray C?


## Dataset: swiss (difficulty = 1/5)

```{r swiss}
head(swiss)
```

<br>

### Goal

* Plot the influence of potential determinant of fertility variation between Swiss provinces.


## Dataset: trees (difficulty = 4/5)

```{r trees}
head(trees)
```

<br>

### Goal

* Compare the approximation of the volume of wood given by $\text{Volume} = c*\text{Height}*\text{Girth}^2$ (with c to be estimated) to the usual consideration that a tree trunk is a cylinder.



## Dataset: stackloss (difficulty = 1/5)

```{r stackloss}
head(stackloss)
```

<br>

### Goal

* Find out if the acid concentration influences the loss of ammonia by the plants.


## Dataset: mammals (difficulty = 3/5)

```{r}
head(mammals)
```

### Goal

* What is the allometric exponent for the growth of brain size with body mass?
* Does this exponent agree with the usual expectation of 2/3?
* How large do we expect a 1kg animal to be?
* Rank organisms by relative brain size (i.e. controlled for body size)
* Repeat the analysis with the dataset `Animals`


## Dataset: Bats (difficulty = 5/5)

```{r Bats}
head(Bats)
```

<br>

### Goal

* reproduce the results on body weight from the study: [Seltmann et al. (Conservation Physiology, 2017)](https://academic.oup.com/conphys/article/doi/10.1093/conphys/cox020/3103058/Habitat-disturbance-results-in-chronic-stress-and?rss=1).


# Solutions

## Answers to Exercises

<br>

You can find the solution to all problems but the bat one 
[here](./Exercise_Practice.html).

Give it a good try before checking how I did it!

# Table of contents

## The Linear Model: LM

* 2.0 [Introduction](./LM_intro.html)
* 2.1 [Fitting procedure](./LM_fitting.html)
* 2.2 [Tests & Intervals](./LM_test_intervals.html)
* 2.3 [Assumptions & Outliers](./LM_assumptions.html)
* 2.4 [Let's practice](./LM_practice.html)

<br>

<div align="right">
[Back to main menu](./Title.html#2)
</div>
