---
title: "LM: Let's practice more!"
author: "Alexandre Courtiol"
date: "`r Sys.Date()`"
output:
  ioslides_presentation:
    widescreen: true
    smaller: true
vignette: >
  %\VignetteIndexEntry{2.5 Practice}
  %\VignetteEncoding{UTF-8}
  %\VignetteEngine{knitr::rmarkdown}
---

```{r setup, include=FALSE}
library(LM2GLMM)
knitr::opts_chunk$set(cache = FALSE, fig.align = "center", fig.width = 4, fig.height = 4)
options(width = 200)
set.seed(1L)
```

## Dataset: chickwts (difficulty = 0/5)

```{r chickwts}
head(chickwts)
levels(chickwts$feed)
```

<br>

### Goal: compare the effectiveness of various feed supplements on the weights of six weeks old chickens.


## Dataset: InsectSprays (difficulty = 2/5)

```{r InsectSprays}
head(InsectSprays)
levels(InsectSprays$spray)
```

<br>

### Goal: compare the effectiveness of various insecticides on the number of insects in agricultural experimental units. What is the efficiency of spray C?

## Dataset: swiss (difficulty = 1/5)

```{r swiss}
head(swiss)
```

<br>

### Goal: plot the influence of potential determinant of fertility variation between Swiss provinces.


## Dataset: trees (difficulty = 3/5)

```{r trees}
head(trees)
```

<br>

### Goal: compare the approximation of the volume of wood given by $\text{Volume} = c*\text{Height}*\text{Girth}^2$ (with c to be estimated) to the usual consideration that a tree trunk is a cylinder.



## Dataset: stackloss (difficulty = 1/5)

```{r stackloss}
head(stackloss)
```

<br>

### Goal: find out if the acid concentration influences the loss of ammonia by the plants.


## Dataset: Bats (difficulty = 5/5)

```{r Bats}
head(Bats)
```

### Goal: reproduce the results on body weight from the study:

[Seltmann et al. (Conservation Physiology, 2017)](https://academic.oup.com/conphys/article/doi/10.1093/conphys/cox020/3103058/Habitat-disturbance-results-in-chronic-stress-and?rss=1).


# Table of content

## The Linear Model: LM

* 2.0 [Introduction](./LM_intro.html)
* 2.1 [Point estimates](./LM_point_estimates.html)
* 2.2 [Uncertainty in point estimates](./LM_uncertainty.html)
* 2.3 [Tests](./LM_tests.html)
* 2.4 [Assumptions and Outliers](./LM_assumptions.html)
* 2.5 [Let's practice more](./LM_practice.html)
