---
title: "LM: Let's practice!"
author: "Alexandre Courtiol"
date: "`r Sys.Date()`"
output:
  ioslides_presentation:
    widescreen: true
    smaller: true
vignette: >
  %\VignetteIndexEntry{2.4 Let's practice}
  %\VignetteEncoding{UTF-8}
  %\VignetteEngine{knitr::rmarkdown}
---

```{r setup, include=FALSE}
library(LM2GLMM)
knitr::opts_chunk$set(cache = TRUE, fig.align = "center", fig.width = 4, fig.height = 4,
                      cache.path = "./cache_knitr/Exo_LM/", fig.path = "./fig_knitr/Exo_LM/")
options(width = 200)
set.seed(1L)
```

## The Linear Model: LM

* 2.0 [Introduction](./LM_intro_course.html)
* 2.1 [Fitting procedure](./LM_fitting_course.html)
* 2.2 [Tests & Intervals](./LM_test_intervals_course.html)
* 2.3 [Assumptions & Outliers](./LM_assumptions_course.html)
* 2.4 [Let's practice](./LM_practice_exercises.html)

<br>

<div align="right">
[Back to main menu](./Title.html#2)
</div>


# Exercises

## Dataset: chickwts -- part 1 (difficulty = 0/5) `r .emo("practice")`

```{r chickwts}
head(chickwts)
levels(chickwts$feed)
```

<br>

`r .emo("goal")`  Test whether diet significantly affects the weights of six weeks old chickens.

`r .emo("goal")`  Compare the effect of different feed supplement types.

`r .emo("goal")`  Plot mean effects for each feed supplement type showing 95% confidence intervals.


## Dataset: stackloss (difficulty = 1/5) `r .emo("practice")`

```{r stackloss}
head(stackloss)
```

<br>

`r .emo("goal")` Find out if the acid concentration influences the loss of ammonia by the plants.


## Dataset: swiss (difficulty = 1/5) `r .emo("practice")`

```{r swiss}
head(swiss)
```

<br>

`r .emo("goal")` Plot the influence of potential determinants of fertility.

`r .emo("goal")` Test the effect of different predictors on fertility.


## Dataset: InsectSprays (difficulty = 2/5) `r .emo("practice")`

```{r InsectSprays}
head(InsectSprays, n = 3)
levels(InsectSprays$spray)
```

<br>

`r .emo("goal")`  Test whether insecticide type significantly affects the number of insects in agricultural experimental units.

`r .emo("goal")` What is the mean number of insects we would predict to find on a plot treated with insecticide C?


## Dataset: chickwts -- part 2 (difficulty = 3/5) `r .emo("practice")`

```{r chickwts2}
head(chickwts)
levels(chickwts$feed)
```

<br>

`r .emo("goal")`  Predict the proportion of chickens larger than 300gr for each feed supplements.


## Dataset: mammals (difficulty = 3/5) `r .emo("practice")`

```{r}
head(mammals)
```

<br>

`r .emo("goal")` What is the allometric exponent for the growth of brain size with body mass?

`r .emo("goal")` Does this exponent agree with the usual expectation of 2/3?

`r .emo("goal")` How large do we expect the brain of a 1kg animal to be?

`r .emo("goal")` Rank organisms by relative brain size (i.e. controlled for body size)

`r .emo("goal")` Repeat the analysis with the dataset `Animals`


## Dataset: trees (difficulty = 4/5) `r .emo("practice")`

```{r trees}
head(trees)
```

<br>

`r .emo("goal")` Compare the approximation of the volume of wood given by $\text{Volume} = c\times\text{Height}\times\text{Girth}^2$ (with $c$ to be estimated) to the approximation of a tree trunk as a cylinder.

<!-- The data here are not the good one and the exercise is very complex... -> let's drop it
## Dataset: Bats (difficulty = 5/5) `r .emo("practice")`

```{r Bats}
head(Bats)
```

<br>

`r .emo("goal")` Reproduce the results on body weight from the study: [Seltmann et al. (Conservation Physiology, 2017)](https://academic.oup.com/conphys/article/doi/10.1093/conphys/cox020/3103058/Habitat-disturbance-results-in-chronic-stress-and?rss=1).
-->

# Solutions

## Answers to Exercises `r .emo("practice")`

<br>

You can find the solution to all problems [here](./LM_practice_solutions.html).

Give it a good try before checking how I did it!

# Table of contents

## The Linear Model: LM

* 2.0 [Introduction](./LM_intro_course.html)
* 2.1 [Fitting procedure](./LM_fitting_course.html)
* 2.2 [Tests & Intervals](./LM_test_intervals_course.html)
* 2.3 [Assumptions & Outliers](./LM_assumptions_course.html)
* 2.4 [Let's practice](./LM_practice_exercises.html)

<br>

<div align="right">
[Back to main menu](./Title.html#2)
</div>
