---
title: "LM: Tests"
author: "Alexandre Courtiol"
date: "`r Sys.Date()`"
output:
  ioslides_presentation:
    widescreen: true
    smaller: true
vignette: >
  %\VignetteIndexEntry{2.3 Tests}
  %\VignetteEncoding{UTF-8}
  %\VignetteEngine{knitr::rmarkdown}
---

```{r setup, include=FALSE}
library(LM2GLMM)
library(spaMM)
library(lattice)
library(mvtnorm)
library(rgl)
library(car)
knitr::opts_chunk$set(cache = TRUE, cache.path = "./cache_knitr/LM_tests/", fig.path = "./fig_knitr/LM_tests/")
```

## You will learn in this session

* how to use the functions ```summary()```, ```anova()``` and ```Anova()```
* that parameters are tested using t-tests
* that variables are tested using F-tests or Likelihood Ratio Tests
* that type II ANOVA are the way to go for testing variables
* that doing many tests is dangerous


# The fitted model

## The UK model

```{r UK data}
(mod_UK1 <- lm(height ~ drink + sex*weight, data = UK))
```

# Testing parameter estimates

## Using the function ```summary()```

```{r summary UK}
summary(mod_UK1)
```

## t-test on parameter estimates

```{r summary random line}
summary(mod_UK1)$coefficient["drinkMost days", ]
```

### Nothing really new: we have seen all this before! (but the p-value)

```{r summary by hand}
estimate <- coef(mod_UK1)["drinkMost days"]
std.error <- sqrt(vcov(mod_UK1)["drinkMost days", "drinkMost days"])
t.value <- estimate/std.error
p.value <- 2*pt(t.value, df = mod_UK1$df.residual)  ## 2x as bilateral test
rbind(estimate, std.error, t.value, p.value)
```


## Comparing an estimate to a specific value

### Ex: Could the parameter ```drinkMost days``` be equal to 0.8799548?

```{r compare to 0.5}
new.t.value <- (estimate - 0.8799548)/std.error
2*pt(new.t.value, df = mod_UK1$df.residual)
```

Notice:

```{r confint Most days}
confint(mod_UK1, parm = "drinkMost days")
```

# Comparing the fitted model to the fit of a null model

## Using the function ```anova()```

```{r null fit}
mod_UK0 <- lm(height ~ 1, data = mod_UK1$model) ## we use model frame to keep same dataset
anova(mod_UK1, mod_UK0)  ## same line as the last one from summary(mod_UK1)
```

<br>

### This test should always be performed first before testing estimates individually!

## The F-test step by step

```{r anova null}
anova(mod_UK1, mod_UK0)
c(df = 11225 - 11231, SS = 256402 - 443561, F <- (-187158/-6) / (256402/11225))
c(256402/11225, summary(mod_UK1)$sigma^2)  ## estimate of the error variance
```

## The F-test step by step

```{r anova null 2}
anova(mod_UK1, mod_UK0)
c(df = 11225 - 11231, SS = 256402 - 443561, F <- (-187158/-6) / (256402/11225))
-187158/-6  ## change in estimated error variance
```

## The F-test step by step

```{r anova null 3}
anova(mod_UK1, mod_UK0)
c(df = 11225 - 11231, SS = 256402 - 443561, F <- (-187158/-6) / (256402/11225))
(p.value <- 1 - pf(F, df1 = 6, df2 = 11225))
```

## The likelihood ratio test

```{r}
mod_UK2 <- lm(height ~ milk, data = UK)
mod_UK0 <- lm(height ~ 1, data = mod_UK2$model)
anova(mod_UK2, mod_UK0, test = "Chisq")
1 - pchisq(2*(c(logLik(mod_UK2)) - c(logLik(mod_UK0))), df = 1)  ## df = K model 1 - K model 2
anova(mod_UK2, mod_UK0, test = "F")["Pr(>F)"]  ## for LM F-test and LRT are equivalent!
```


# Comparing two nested models

## t-test and F-test are equivalent for quantitative variables

```{r F test milk}
mod_UK3 <- lm(height ~ milk + sex*weight, data = UK)
summary(mod_UK3)$coefficient["milk", ]
mod_UK3_no_milk <- lm(height ~ sex*weight, data = mod_UK3$model)
anova(mod_UK3, mod_UK3_no_milk)
```

## Nested model comparison allows for testing the overall effect of several parameters

```{r F test weight}
mod_UK1 <- lm(height ~ drink + sex*weight, data = UK)
mod_UK1_no_weight <- lm(height ~ drink + sex, data = mod_UK1$model)
anova(mod_UK1, mod_UK1_no_weight)
```

## Nested model comparison thus allows for testing the overall effect of a qualitative variable

```{r F test alcool}
mod_UK1 <- lm(height ~ drink + sex*weight, data = UK)
mod_UK1_no_alcool <- lm(height ~ sex*weight, data = mod_UK1$model)
anova(mod_UK1, mod_UK1_no_alcool)
```

## Testing all variables in one go with ```anova()```

### Sequential test: Type 1 ANOVA (dangerous)

```{r anova 1}
anova(mod_UK1)
```

## Testing all variables in one go with ```Anova()```

### Marginal test: Type 2 ANOVA (recommended)

```{r anova 2}
library(car)
Anova(mod_UK1)
```

<br>

Note : test of simple terms consider other simple terms to be there, but not interactions.

Tip: never use type 3 ANOVA!



# Comparing two non-nested models

## We should not use F-test or LRT


### You may use Akaike Information Criterion (AIC = -2 * (logLik - K))

```{r non nested}
UK_for_both <- na.omit(UK[, c("height", "drink", "sex", "weight", "milk")])
mod_UK1 <- lm(height ~ drink + sex*weight, data = UK_for_both)
mod_UK5 <- lm(height ~ milk + sex*weight, data = UK_for_both)
AIC(mod_UK5)
AIC(mod_UK1)
exp((AIC(mod_UK5) - AIC(mod_UK1)) / 2) ## evidence ratio
```

The model ```mod_UK5``` is ```r round(exp((AIC(mod_UK5) - AIC(mod_UK1)) / 2), 1)``` times more
likely to minimizes the (estimated) information loss than ```mod_UK1```.


# Comparing too many things

## Simulate nothingness

```{r simu_null}
simu_null <- function(N, K) {
  d <- as.data.frame(replicate(K + 1, rnorm(N)))
  colnames(d)[1] <- "y"
  colnames(d)[2:(K + 1)] <- paste0("x", 1:K)
  return(d)
}

set.seed(1L)
simu_null(N = 5, K = 3)
```

## Testing nothingness

```{r robustness}
K <- 3
set.seed(1L)
lm(y ~ ., data = simu_null(N = 5, K = K))
set.seed(1L)
(p.values <- Anova(lm(y ~ ., data = simu_null(N = 5, K = K)))["Pr(>F)"][1:K, 1])
min(p.values) <= 0.05
```

## Comparing many estimates/models decreases the robustness!!!

```{r robustness 2}
set.seed(1L)
mean(replicate(1000, min(Anova(lm(y ~ ., data = simu_null(N = 100, K = 1)))["Pr(>F)"][1, 1]) <= 0.05))

mean(replicate(1000, min(Anova(lm(y ~ ., data = simu_null(N = 100, K = 3)))["Pr(>F)"][1:3, 1]) <= 0.05))

mean(replicate(1000, min(Anova(lm(y ~ ., data = simu_null(N = 100, K = 10)))["Pr(>F)"][1:10, 1]) <= 0.05))

mean(replicate(1000, min(Anova(lm(y ~ ., data = simu_null(N = 1000, K = 50)))["Pr(>F)"][1:50, 1]) <= 0.05))
```

## What shall I do?

* Don't go fishing (forget backward selection!!!)
* Always compare the fit to null model fit and do not do more tests if not signif

```{r null fit test}
mean(replicate(1000, {
  d <- simu_null(N = 1000, K = 50)
  mod <- lm(y ~ ., data = d)
  mod0 <- lm(y ~ 1, data = d)
  anova(mod, mod0)[2, "Pr(>F)"] <= 0.05
}))

```

* Always test the qualitative variable as a whole before comparing levels


## What shall I do?

* Correct for multiple comparisons

```{r ghlt, eval = FALSE}
mod_drink <- lm(height ~ drink, data = UK)
library(multcomp)
summary(glht(mod_drink, linfct = mcp(drink = "Tukey")))
```

```{r ghlt echo, echo = FALSE}
options(width = 200)
mod_drink <- lm(height ~ drink, data = UK)
library(multcomp)
summary(glht(mod_drink, linfct = mcp(drink = "Tukey")))
```

## What shall I do?

* Correct for multiple comparisons

```{r ghlt plot, fig.align = "center", fig.width = 8, fig.height = 4}
par(oma = c(0, 15, 0, 0))  ## increase left outer margin
plot(glht(mod_drink, linfct = mcp(drink = "Tukey")))
```

## What shall I do?

* Use at least well choosen contrasts

```{r drink}
UK$drink2 <- relevel(UK$drink, ref = "Not at all")
summary(mod_drink2 <- lm(height ~ drink2, data = UK))$coef
## compare to summary(glht(mod_drink2, linfct = mcp(drink2 = "Dunnett")))
```


## What you need to remember

* how to use the functions ```summary()```, ```anova()``` and ```Anova()```
* that parameters are tested using t-tests
* that variables are tested using F-tests or Likelihood Ratio Tests
* that type II ANOVA are the way to go for testing variables
* that doing many tests is dangerous


# Table of content

## The Linear Model: LM

* 2.0 [Introduction](./LM_intro.html)
* 2.1 [Point estimates](./LM_point_estimates.html)
* 2.2 [Uncertainty in point estimates](./LM_uncertainty.html)
* 2.3 [Tests](./LM_tests.html)
* 2.4 [Assumptions](./LM_assumptions.html)

