---
title: "LM: Uncertainty in point estimates"
subtitle: "exercises"
author: "Alexandre Courtiol"
date: "`r Sys.Date()`"
output:
  ioslides_presentation:
    widescreen: true
    smaller: true
vignette: >
  %\VignetteIndexEntry{2.2 z-------------------------------------------------------------exercises}
  %\VignetteEncoding{UTF-8}
  %\VignetteEngine{knitr::rmarkdown}
---

```{r setup, include=FALSE}
library(LM2GLMM)
knitr::opts_chunk$set(cache = FALSE,
                      fig.align = "center",
                      fig.width = 4,
                      fig.height = 4,
                      cache.path = "./cache_knitr/LM_uncertainty_ex/",
                      fig.path = "./fig_knitr/LM_uncertainty_ex/")
options(width = 110)
set.seed(1L)
```

# Distributions of point estimates

## Theoretical vs actual distribution and sample size

Adapt this example from the course using to compare the match between the theoretical and the actual distribution of the slope for different values of $n$. What do you notice?

```{r redraw samples, eval = FALSE}
set.seed(123)
Alien <- simulate_Aliens()
fit_stats <- lm(size ~ humans_eaten, data = Alien)
new_beta_estimates <- t(replicate(1000, coef(update(fit_stats, data = simulate_Aliens()))))
```

```{r plot ecdf intercept, fig.align = "center", fig.height = 3.5, fig.width = 3.5, eval = FALSE}
curve(pnorm(x, mean = coef(fit_stats)[1],
    sd = sqrt(vcov(fit_stats)[[1]])),
    from = 30, to = 70, ylab = "cdf", lwd = 3)
plot(ecdf(new_beta_estimates[, 1]), col = "red",
     add = TRUE)
```


# Confidence intervals 

## CI on point estimates

Compare the CI on the estimate of the slope for `humans_eaten` using:

- the asymptotic method
- non-parametric bootstrap
- likelihood profiling



