---
title: "Simulating data"
author: "Alexandre Courtiol"
date: "`r Sys.Date()`"
output:
  ioslides_presentation:
    widescreen: true
    smaller: true
vignette: >
  %\VignetteIndexEntry{Simulating data}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}

---
```{r setup, include=FALSE}
library(LM2GLMM)
```

## Why simulating data?
* it helps to understand what exactly linear models are
* it is a good way for you to know that you understand the meaning of your estimates
* it allows for the study of the power and robustness of LM in general and particular cases

## Pseudo-random numbers in R

```{r rnorm}
rnorm(5, mean = 2, sd = 0.1)  ## Tip: check ?Distributions for other distribution
rnorm(5, mean = 2, sd = 0.1)
set.seed(14353L)
rnorm(5, mean = 2, sd = 0.1)
set.seed(14353L)
rnorm(5, mean = 2, sd = 0.1)
```

## Step one: simulating the independent variables

### Goal
Complete the function that follows, so that is returns a dataframe containing
1 continuous independent variable and 1 categorical independent variable.

```{r goal function}
simulate_my_predictors_1C1G <- function(n, k){
  ## n is the number of datapoints per group
  ## k is the number of groups
  ...
  }
```

## Step one: simulating the independent variables

### Goal
Complete the function that follows, so that is returns a dataframe containing
1 continuous independent variable and 1 categorical independent variable.

### Possible solution
```{r target function}
simulate_predictors_1C1G
```

## Step two: simulating the response variables
