---
title: "Answers to Exercises: Let's practice more GLM"
author: "Alexandre Courtiol"
date: "`r Sys.Date()`"
output:
  html_vignette:
    toc: true
vignette: >
  %\VignetteIndexEntry{3.X Answers to Exercises: Let's practice more GLM}
  %\VignetteEncoding{UTF-8}
  %\VignetteEngine{knitr::rmarkdown}
---
```{r setup, include=FALSE}
library(LM2GLMM)
library(car)
knitr::opts_chunk$set(cache = FALSE, fig.align = "center", fig.width = 6, fig.height = 6)
```

## Disclamer

In this vignette I illustrate the key steps required to solve the exercises. By no means I am trying to provide a detailed report of the analysis of each dataset. Any such report would need additional analyses and results would have to be written in good English, not in telegraphic style or lines of codes. I have also written this vignette very quickly, so double check I did not mess anything up...


# Dataset: InsectSprays (difficulty = 1/5)

## Goal

* compare the effectiveness of various insecticides on the number of insects in agricultural experimental units
* compare the predictions from each insecticides from a GLM fit to those done while practicing the LM

Note: at this stage I will not check the assumptions and detail the dataset since we did it before and I will not look at the GLM assumptions since we have not yet seen this.


## LM model
```{r}
mod_insect <- lm(count ~ spray, data = InsectSprays)
```

## LM BoxCox
```{r}
InsectSprays$count_bis <- InsectSprays$count + 1
mod_insect_bis <- update(mod_insect, count_bis ~ .)
bc <- car::powerTransform(mod_insect_bis)
InsectSprays$count_bc <- car::bcPower(InsectSprays$count_bis, bc$lambda)
mod_insect_bc <- update(mod_insect_bis, count_bc ~ .)
```

## GLM
```{r}
mod_insect_glm <- glm(count ~ spray, data = InsectSprays, family = poisson())
```

## Tests comparing the fitted models to their respective null models
```{r}
anova(mod_insect, update(mod_insect, . ~ 1))
anova(mod_insect_bc, update(mod_insect_bc, . ~ 1))
anova(mod_insect_glm, update(mod_insect_glm, . ~ 1), test = "LR")
```

## Predictions
```{r}
data.for.pred <- data.frame(spray = levels(InsectSprays$spray))
pred_lm <- predict(mod_insect, newdata = data.for.pred, interval = "confidence")
pred_bc <- predict(mod_insect_bc, newdata = data.for.pred, interval = "confidence")
unboxcox <- function(x, lambda) (x*lambda + 1)^(1/lambda)
pred_bc_unboxcox <- unboxcox(x = pred_bc, lambda = bc$lambda) - 1  ## we remove one as we added one so to be able to do the BoxCox
pred_glm_eta <- predict(mod_insect_glm, newdata = data.for.pred, se.fit = TRUE)
pred_glm_table <- data.frame(fit = exp(pred_glm_eta$fit))
pred_glm_table$lwr <- exp(pred_glm_eta$fit + qnorm(0.025)*pred_glm_eta$se.fit)
pred_glm_table$upr <- exp(pred_glm_eta$fit + qnorm(0.975)*pred_glm_eta$se.fit)
```

## Plot
```{r}
plot(InsectSprays$count ~  as.numeric(InsectSprays$spray), xlim = c(0.5, 6.5), ylim = c(0, 30),
     ylab = "Predicted number of instects", xlab = "Sprays", axes = FALSE)
axis(2, las = 2)
axis(1, at = 1:6, labels = levels(InsectSprays$spray))
box()

points(pred_lm[, "fit"] ~  I(-0.2 + 1:6), col = "blue", pch = 20)

arrows(x0 = -0.2 + 1:6, x1 = -0.2 + 1:6, y0 = pred_lm[, "lwr"],
       y1 = pred_lm[, "upr"], code = 3, col = "blue", angle = 90, length = 0.1)

points(pred_bc_unboxcox[, "fit"] ~  I(-0.1 + 1:6), col = "green", pch = 20)

arrows(x0 = -0.1 + 1:6, x1 = -0.1 + 1:6, y0 = pred_bc_unboxcox[, "lwr"],
       y1 = pred_bc_unboxcox[, "upr"], code = 3, col = "green", angle = 90, length = 0.1)

points(pred_glm_table[, "fit"] ~  I(0.1 + 1:6), col = "red", pch = 20)

arrows(x0 = 0.1 + 1:6, x1 = 0.1 + 1:6, y0 = pred_glm_table[, "lwr"],
       y1 = pred_glm_table[, "upr"], code = 3, col = "red", angle = 90, length = 0.1)

legend("top", horiz = TRUE, fill = c("blue", "green", "red"), legend = c("LM", "LM + BoxCox", "GLM Poisson"), bty = "n")
```





