---
title: "GLM: Uncertainty & Tests"
author: "Alexandre Courtiol"
date: "`r Sys.Date()`"
output:
  ioslides_presentation:
    widescreen: true
    smaller: true
vignette: >
  %\VignetteIndexEntry{3.1 Uncertainty & Tests}
  %\VignetteEncoding{UTF-8}
  %\VignetteEngine{knitr::rmarkdown}
---

```{r setup, include=FALSE}
library(LM2GLMM)
options(width = 120)
knitr::opts_chunk$set(cache = FALSE, fig.width = 5, fig.height = 5, fig.align = "center")
```

## You will learn in this session


## Our toy models

```{r}
set.seed(1L)
Aliens <- data.frame(humans_eaten = round(runif(n = 100, min = 0, max = 15)))
Aliens$size  <- rnorm( n = 100, mean = 50 + 1.5 * Aliens$humans_eaten, sd = 5)
Aliens$eggs  <- rpois( n = 100, lambda = exp(-1 + 0.1 * Aliens$humans_eaten))
Aliens$happy <- rbinom(n = 100, size = 1, prob = plogis(-3 + 0.3 * Aliens$humans_eaten))
Aliens$all_eyes  <- round(runif(nrow(Aliens), min = 1, max = 12))
Aliens$blue_eyes <- rbinom(n = nrow(Aliens), size = Aliens$all_eyes, prob = plogis(-2 + 0.5 * Aliens$humans_eaten))
Aliens$pink_eyes <- Aliens$all_eyes - Aliens$blue_eyes

mod_gauss <- glm(size  ~ humans_eaten, family = gaussian(), data = Aliens)
mod_poiss <- glm(eggs  ~ humans_eaten, family = poisson(),  data = Aliens)
mod_binar <- glm(happy ~ humans_eaten, family = binomial(), data = Aliens)
mod_binom <- glm(cbind(blue_eyes, pink_eyes) ~ humans_eaten, family = binomial(), data = Aliens)
```


# Uncertainty in parameter estimates

## Uncertainty in parameter estimates

Poisson family:

<br>

```{r}
X <- model.matrix(mod_poiss)
W <- matrix(0, ncol = nrow(mod_poiss$model), nrow = nrow(mod_poiss$model))
diag(W) <- mod_poiss$weights
t(X) %*% W %*% X
(XTWX <- crossprod(mod_poiss$R))
```

## Uncertainty in parameter estimates

Poisson family:

<br>

```{r}
vcov(mod_poiss)
phi <- 1
phi*solve(XTWX)
```

## Uncertainty in parameter estimates

Gaussian family:

<br>

```{r}
vcov(mod_gauss)
XTWX <- crossprod(mod_gauss$R)
phi <- mod_gauss$deviance / mod_gauss$df.residual
phi*solve(XTWX)
```

## Uncertainty in parameter estimates

Binomial family (binary case):

<br>

```{r}
vcov(mod_binar)
XTWX <- crossprod(mod_binar$R)
phi <- 1
phi*solve(XTWX)
```

## Uncertainty in parameter estimates

Binomial family (general case):

<br>

```{r}
vcov(mod_binom)
XTWX <- crossprod(mod_binom$R)
phi <- 1
phi*solve(XTWX)
```


## What you need to remember



# Table of content

## The Generalized Linear Model: GLM

* 3.0 [Introduction](./GLM_intro.html)
* 3.1 [Uncertainty & Tests](./GLM_uncertainty.html)
* 3.2 [Residuals & Assumptions](./GLM_assumptions.html)
* 3.3 [Overdispersion & Zero-inflation](./GLM_overdispersion.html)
* 3.4 [Let's practice more](./GLM_practice.html)

