---
title: "GLM: Let's practice"
author: "Alexandre Courtiol"
date: "`r Sys.Date()`"
output:
  ioslides_presentation:
    widescreen: true
    smaller: true
vignette: >
  %\VignetteIndexEntry{3.3 Practice}
  %\VignetteEncoding{UTF-8}
  %\VignetteEngine{knitr::rmarkdown}
---

```{r setup, include=FALSE}
library(LM2GLMM)
options(width = 100)
knitr::opts_chunk$set(cache = TRUE, fig.width = 5, fig.height = 5, fig.align = "center",
                      cache.path = "./cache_knitr/Exo_GLM/", fig.path = "./fig_knitr/Exo_GLM/")
```

## Dataset: esoph

```{r}
head(esoph)
```

### Goal

* Find out whether you should rather limit alcohol consumption or tabacco consumption in order to avoid developing an oesophageal cancer? Does it depends on your age?


## Dataset: effects::TitanicSurvival

```{r}
data(TitanicSurvival, package = "carData")
head(TitanicSurvival)
```

### Goal

* Find out who mostly survived and who mostly died during the Titanic disaster of 1912.
* Compute the odds-ratio for survival between the two most contrasting groups of individuals. 


## Dataset: LM2GLMM::Challenger

```{r}
head(Challenger)
```

### Goal

* What was the probability of O-ring failure when the space shuttle Challenger took off on the 28th of January 1986 by 31 degrees F?


## Dataset: LM2GLMM::UK

```{r}
UK[1:2, ]
```

### Goal

* Try to identify the determinant of the smoking behaviour of children.
* Try to identify the determinant of bronchitis in children.
* Try to identify the determinant of the variable ```backward```.


## Dataset: LM2GLMM::HSE98women

```{r}
head(HSE98women)
```

### Goal

* How does the probability of not having periods changes with age, smoking status and BMI?
* How does the probability of having been drunk recently changes with age?
* Does the number of drinking problems experienced during life depends on education?


## Dataset: MASS::mammals

```{r}
data(mammals, package = "MASS")
head(mammals)
```

### Goal

* What is the allometric exponent for the growth of brain size with body mass?
* Does this exponent agree with the usual expectation of 2/3?
* How large do we expect a 1kg animal to be?
* Rank organisms by relative brain size (i.e. controlled for body size)
* Repeat the analysis with the dataset ```MASS::Animals```

## Dataset: LM2GLMM::Surprise

```{r}
head(Surprise)
```

### Goal

* Do children value more the type of the present, the cost of the present, or both?


# Table of contents

## The Generalized Linear Model: GLM

* 3.0 [Introduction](./GLM_intro.html)
* 3.1 [Intervals & Tests](./GLM_intervals.html)
* 3.2 [Residuals & Assumptions](./GLM_assumptions.html)
* 3.3 [Let's practice more](./GLM_practice.html)
