---
title: "LMM: Solving LM problems"
author: "Alexandre Courtiol"
date: "`r Sys.Date()`"
output:
  ioslides_presentation:
    widescreen: true
    smaller: true
vignette: >
  %\VignetteIndexEntry{4.2 Solving LM problems using LMM}
  %\VignetteEncoding{UTF-8}
  %\VignetteEngine{knitr::rmarkdown}
editor_options: 
  chunk_output_type: console
---

```{r setup, include=FALSE}
library(LM2GLMM)
library(spaMM)
library(lme4)
spaMM.options(nb_cores = 4L)
options(width = 120)
knitr::opts_chunk$set(cache = TRUE, cache.path = "./cache_knitr/LMM_solving_pb/", fig.path = "./fig_knitr/LMM_solving_pb/", fig.width = 5, fig.height = 5, fig.align = "center", error = FALSE)
```

## Mixed-effects models

* 4.0 [Introduction to LMM](./LMM_intro.html)
* 4.1 [Random effects](./LMM_random.html)
* 4.2 [Solving LM problems using LMM](./LMM_solving_pb.html)
* 4.3 [A showcase of some useful applications](./LMM_showcase.html)
* 4.4 [Introduction to GLMM](./GLMM_intro.html)
* 4.5 [More complex mixed models](./MM_more.html)


## You will learn in this session

* how to handle temporal and spatial autocorrelation
* how to model heteroskedasticity


# Temporal autocorrelation

# Temporal autocorrelation in discrete (equaly spaced) time steps

## The ```AirPassengers``` data

```{r airpassenger data}
AirPassengers
```


## The ```AirPassengers``` data

```{r plot airpassenger data}
plot(AirPassengers)
```


## Reformating the dataset for the fit

```{r reshape airpassenger}
air <- data.frame(passengers = as.numeric(AirPassengers),
                  year = rep(1949:1960, each = 12),
                  month = factor(rep(1:12, 12)))
air$time <- 1:nrow(air)
air
```


## Looking at the average trend per year

```{r airpassenger time trend per year}
plot(with(air, tapply(passengers, year, mean)) ~ I(1949:1960),
     ylab = "Mean number of passengers", xlab = "Year", type = "b")
```


## Looking at the average trend per month

```{r airpassenger time trend per month}
plot(with(air, tapply(passengers, month, mean)) ~ I(1:12),
     ylab = "Mean number of passengers", xlab = "Month", type = "h")
```


## Simple fit

```{r airpassenger simple fit}
(mod_air <- lm(passengers ~ year * month, data = air))
```


## The problem

```{r airpassenger residuals}
plot(residuals(mod_air), type = "b")
abline(h = 0, lty = 2, col = "red")
```


## The problem

```{r airpassenger dw test}
lmtest::dwtest(mod_air)
```


## The problem

```{r airpassenger acf}
acf(residuals(mod_air))
```


## Solution

```{r airpassenger AR1, message = FALSE}
library(nlme)
MAR1 <- corAR1(value = 0.5, form = ~ 1|year, fixed = FALSE)
MAR1 <- Initialize(MAR1, data = air)
round(corMatrix(MAR1)[["1950"]], 2)
```


## Solution
```{r airpassenger lme}
(mod_air2 <- lme(passengers ~ month * year, random = ~ 1 | year, data = air,
                 correlation = MAR1, method = "REML"))  ## confusing: correlation is modelled between months!
```


## Alternative code

```{r airpassenger lme2}
(mod_air2b <- lme(passengers ~ month * year, random = ~ 1 | year, data = air,
                 correlation = corAR1(form = ~ 1|year), method = "REML"))
```


## Testing the temporal autocorrelation

```{r airpassenger test temp autocorr}
mod_air3 <- lme(passengers ~ month * year, random = ~ 1 | year, data = air, method = "REML")
anova(mod_air2, mod_air3)
```


## Alternative autocorrelation structures

```{r airpassenger corARMA}
mod_airARMA1 <- update(mod_air2, correlation = corARMA(form = ~ 1 | year, p = 1, q = 0))
mod_airARMA2 <- update(mod_air2, correlation = corARMA(form = ~ 1 | year, p = 4, q = 0))
mod_airARMA3 <- update(mod_air2, correlation = corARMA(form = ~ 1 | year, p = 2, q = 2))

rbind(mod_air2 = AIC(mod_air2),
      mod_airARMA1 = AIC(mod_airARMA1),
      mod_airARMA2 = AIC(mod_airARMA2),
      mod_airARMA3 = AIC(mod_airARMA3))
```

<br>

Note: do not compare AICs or likelihoods from ```nlme``` to those from other packages!

(it seems they have failed to consider a constant term...)


## Fitted values

```{r airpassenger fitted}
mod_air4 <- update(mod_air2, correlation = corARMA(form = ~ 1 | year, p = 4, q = 0), method = "ML")
data.for.plot <- expand.grid(month = factor(1:12), year = 1949:1960)
data.for.plot$obs <- air$passengers
data.for.plot$time <- seq(1949, 1960, length = (1960 - 1949 + 1) * 12)
data.for.plot$fit_lm <- predict(mod_air)
data.for.plot$fit_lme <- predict(mod_air4)
```


## Fitted values

```{r airpassenger plot fitted}
plot(obs ~ time, data = data.for.plot, type = "l", ylim = c(0, 700), ylab = "Passengers")
points(fit_lm ~ time, data = data.for.plot, type = "l", col = "red")
points(fit_lme ~ time, data = data.for.plot, type = "l", col = "blue")
```


## Better, but good enough?

```{r airpassenger final}
plot(residuals(mod_air4), type = "l")
abline(h = 0, lty = 2, col = "red")
```



## Dealing with temporal auto-correlation using ```spaMM```

```{r spaMM air}
library(spaMM)
air$year_z <- scale(air$year) ## otherwise hard to fit!
mod_air_spaMM1 <- fitme(passengers ~ month * year_z + AR1(1|time %in% year), data = air, method = "REML")
mod_air_spaMM2 <- fitme(passengers ~ month * year_z, data = air, method = "REML")

print(AIC(mod_air_spaMM1))
print(AIC(mod_air_spaMM2))
```


## Examining the best model

```{r air best}
mod_air_spaMM1
```


## Examining the best model

```{r air best2}
mod_air_spaMM1$corrPars[[1]]
res <- matrix(residuals(mod_air_spaMM2)[,1], ncol = 12)
(res_autocorr <- sapply(1:11, function(month)
  cor(res[month, ], res[(month + 1), ])
))
summary(res_autocorr)

```


## Fitted values

```{r air fitted, fig.height = 4, fig.width = 4}
data.for.plot$pred_spaMM <- predict(mod_air_spaMM1)
plot(obs ~ time, data = data.for.plot, type = "l", lwd = 3, ylim = c(0, 700), ylab = "Passengers")
points(pred_spaMM ~ time, data = data.for.plot, type = "l", col = "green")
```

Note: never extrapolate using such model! The perfect fit is not unusual.


## Testing the effect of years

### ```spaMM```
```{r spaMM air 2}
mod_air_spaMM2ML <- fitme(passengers ~ month*year_z + AR1(1|time %in% year), data = air, method = "ML")

mod_air_no_year <- fitme(passengers ~ month + AR1(1|time %in% year), data = air, method = "ML")
```

```{r air spaMM anova}
anova(mod_air_spaMM2ML, mod_air_no_year)

c(logLik(mod_air_spaMM2ML), logLik(mod_air_no_year))
```


## Testing the effect of years

### ```nlme```

```{r air years}
mod_air3ML <- lme(passengers ~ month * year, random = ~ 1 | time, data = air,
                 correlation = corAR1(), method = "ML")

mod_air_no_year2 <- lme(passengers ~ month, random = ~ 1 | time, data = air,
                 correlation = corAR1(), method = "ML")
```

```{r air years anova}
anova(mod_air3ML, mod_air_no_year2)
```



# Temporal autocorrelation in continuous time

## The ```nlme::BodyWeight``` dataset

```{r bodyweight data, fig.width = 10}
data("BodyWeight", package = "nlme")
plot(BodyWeight)
```


## The ```nlme::BodyWeight``` dataset

```{r bodyweight reshape}
body <- as.data.frame(BodyWeight)
body$Rat <- factor(body$Rat, levels = 1:16, order = FALSE)
str(body)
unique(body$Time)
```


## Fitting the model

```{r bodyweight lme}
(mod_rat1 <- lme(weight ~ Diet * Time, random = ~ Time|Rat, data = body))
```


## Checking residuals

```{r bodyweight residuals}
plot(mod_rat1) ## there is some homoscedasticity but we will ignore it for now
```


## Checking residuals

```{r bodyweight plot residuals}
plot(residuals(mod_rat1), type = "b")
```


## Fitting continuous temporal autocorrelation

```{r bodyweight corExp}
(mod_rat2 <- lme(weight ~ Diet * Time, random = ~ Time|Rat, correlation = corExp(form = ~ Time), data = body))
```


## Model comparison using nlme

```{r bodyweight anova}
anova(mod_rat1, mod_rat2)
```

<br>

Note: the comparison makes sense as models are nested and fitted with REML.


## Same fit with ```spaMM```

```{r bodyweight fitme spaMM}
(mod_rat_spaMM <- fitme(weight ~ Diet * Time + (Time|Rat) + Matern(1|Time),
                        fixed = list(nu = 0.5),
                        data = body, method = "REML"))
```


## AR1 vs Matern

```{r bodyweight fitme AR1 spaMM}
mod_rat_spaMM_AR1 <- fitme(weight ~ Diet * Time + (Time|Rat) + AR1(1|Time),
                           fixed = list(nu = 0.5),
                           data = body, method = "REML")
logLik(mod_rat_spaMM)
logLik(mod_rat_spaMM_AR1)
```

<br>

Note 1: AR1 and Matern (with nu = 0.5) are equivalent if time is discrete!

Note 2: AR1 cannot work if time is not discrete!


## Fitted values: ```nlme``` vs ```spaMM```

```{r bodyweight spaMM plot}
plot(predict(mod_rat_spaMM), predict(mod_rat2))
abline(0, 1, col = "red")
```


## Model comparison

```{r corrHLfit spaMM2}
mod_rat_spaMM2 <- fitme(weight ~ Diet * Time + (Time|Rat), data = body,
                        method = "REML")

print(AIC(mod_rat_spaMM))
print(AIC(mod_rat_spaMM2))
```


## Testing the overall effect of diet

### ```spaMM```

```{r spaMM rat}
mod_rat_spaMM3ML <- fitme(weight ~ Diet * Time + (Time|Rat) + Matern(1|Time),
                          fixed = list(nu = 0.5), data = body, method = "ML")
mod_rat_no_diet <- fitme(weight ~ 1 + Time + (Time|Rat) + Matern(1|Time),
                         fixed = list(nu = 0.5), data = body, method = "ML")
```

```{r bodyweight spaMM anova}
anova(mod_rat_spaMM3ML, mod_rat_no_diet)

c(logLik(mod_rat_spaMM3ML), logLik(mod_rat_no_diet))
```


## Testing the overall effect of diet

### ```nlme```

```{r nlme rat}
mod_rat3ML <- lme(weight ~ Diet * Time, random = ~ Time|Rat,
                  correlation = corExp(form = ~ Time, nugget = TRUE), data = body, method = "ML")

mod_rat_no_diet2 <- lme(weight ~ 1 + Time, random = ~ Time|Rat,
                        correlation = corExp(form = ~ Time, nugget = TRUE), data = body, method = "ML")
```

```{r nlme rat anova}
anova(mod_rat3ML, mod_rat_no_diet2)
```


## More complex autocorrelation functions

There are more complex autocorrelation functions out there.

```spaMM```:
```{r spaMM full Matern}
mod_rat_Matern <- fitme(weight ~ Diet * Time + (Time|Rat) + Matern(1|Time), data = body)
c(logLik(mod_rat_spaMM)[[1]], logLik(mod_rat_Matern)[[1]])
```

```nlme```:
```{r other correlations}
mod_rat_corExp <- update(mod_rat1, corr = corExp(form = ~ Time, nugget = TRUE))
mod_rat_corRatio <- update(mod_rat1, corr = corRatio(form = ~ Time))
mod_rat_corSpher <- update(mod_rat1, corr = corSpher(form = ~ Time))
mod_rat_corLin <- update(mod_rat1, corr = corLin(form = ~ Time))
mod_rat_corGaus <- update(mod_rat1, corr = corGaus(form = ~ Time))
c(logLik(mod_rat2)[[1]], logLik(mod_rat_corExp)[[1]], logLik(mod_rat_corRatio)[[1]], 
  logLik(mod_rat_corSpher)[[1]], logLik(mod_rat_corLin)[[1]], logLik(mod_rat_corGaus)[[1]])
```

Note: do not compare likelihood across packages!

# Spatial autocorrelation

## Maximum normalised-difference vegetation index in north Cameroon

```{r loaloa data}
data("Loaloa")
ndvi <- Loaloa[, c("maxNDVI", "latitude", "longitude")]
head(ndvi)
```


## Visualising the data

```{r loaloa plot, fig.width = 9}
library(maps)
spaMMplot2D(x = ndvi$longitude, y = ndvi$latitude, z = ndvi$maxNDVI, add.map = TRUE,
            xlab = "Longitude", ylab = "Latitude", plot.title = title(main = "max NDVI"))
```


## Visualising the data

```{r loaloa pairs}
pairs(ndvi)
```


## Fitting the model

```{r mod ndvi}
(mod_ndvi1 <- fitme(maxNDVI ~ 1 + Matern(1|longitude + latitude), data = ndvi, method = "REML"))
```


## Predictions

```{r mapMM 1, fig.width = 9}
mapMM(mod_ndvi1, add.map = TRUE, plot.title = title(xlab = "Longitude", ylab = "Latitude"))
```


## Predictions

```{r mapMM 2, fig.width = 9}
filled.mapMM(mod_ndvi1, add.map = TRUE, plot.title = title(xlab = "Longitude", ylab = "Latitude"))
```


## Prediction uncertainty

```{r loaloa pred uncertainty}
x.for.pred <- seq(min(ndvi$longitude), max(ndvi$longitude), length.out = 100)
y.for.pred <- seq(min(ndvi$latitude), max(ndvi$latitude), length.out = 50)
data.for.pred <- expand.grid(longitude = x.for.pred, latitude = y.for.pred)
gridpred <- predict(mod_ndvi1, newdata = data.for.pred, variances = list(predVar = TRUE))
data.for.pred$predVar <- attr(gridpred, "predVar")
m <- matrix(data.for.pred$predVar, ncol = length(y.for.pred))
```


## Prediction uncertainty

```{r loaloa plot uncertainty, fig.width = 9}
spaMM.filled.contour(x = x.for.pred, y = y.for.pred, z = m, plot.axes = {
  points(ndvi[, c("longitude", "latitude")])}, col = spaMM.colors(redshift = 3))
```


# Heteroscedasticity

## Heteroscedasticity

We used random effects to model variance components from random variables.

We can thus use the same kind of tools to also model the variance components of the errors.

Do not mix-up random variance components from residual ones: remember that in the error, all covariance terms are null (by definition).


## Let's revisit the rats

```{r rat hetero}
mod_rat_spaMM <- fitme(weight ~ Diet * Time + (Time|Rat) + AR1(1|Time), data = body,
                            method = "REML")
```

```{r rat hetero plot}
coplot(residuals(mod_rat_spaMM) ~ I(1:nrow(body)) | body$Diet, show.given = FALSE)
```


## Let's revisit the rats

```{r rat hetero 2}
mod_rat_hetero <- fitme(weight ~ Diet * Time + (Time|Rat) + AR1(1|Time), data = body,
                        method = "REML", resid.model = ~ Diet)
```

```{r rat summary, results="hide"}
summary.tables <- summary(mod_rat_hetero)
```

```{r rat phi}
summary.tables$phi_table
```

```{r rat AIC spaMM}
print(rbind(AIC(mod_rat_spaMM),
            AIC(mod_rat_hetero)))
```


## Let's re-test the overal effect of the diet

```{r hetero 3}
mod_rat_hetero <- fitme(weight ~ Diet * Time + (Time|Rat) + AR1(1|Time), data = body,
                           method = "ML", resid.model = ~ Diet)

mod_rat_hetero0 <- fitme(weight ~ Time + (Time|Rat) + AR1(1|Time), data = body,
                           method = "ML", resid.model = ~ Diet)

anova(mod_rat_hetero, mod_rat_hetero0)
```


## Heteroscedasticity in LM

### Just do the same!

```{r hetero simple}
set.seed(1L)
d <- data.frame(y = c(rnorm(100, mean = 10, sd = sqrt(10)),
                      rnorm(100, mean = 20, sd = sqrt(20))),
                group = factor(c(rep("A", 100), rep("B", 100))))

m <- fitme(y ~ group, resid.model = ~ group, data = d, method = "REML")
unique(m$phi)
```


## What you need to remember

* how to handle temporal and spatial autocorrelation
* how to model heteroskedasticity


# Table of contents

## Mixed-effects models

* 4.0 [Introduction to LMM](./LMM_intro.html)
* 4.1 [Random effects](./LMM_random.html)
* 4.2 [Solving LM problems using LMM](./LMM_solving_pb.html)
* 4.3 [A showcase of some useful applications](./LMM_showcase.html)
* 4.4 [Introduction to GLMM](./GLMM_intro.html)
* 4.5 [More complex mixed models](./MM_more.html)
